---
title: "Wilms Tumour Methylation Preprocessing Validation Cohorts"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_smooth: true
    toc_depth: 2
    number_sections: true
    number-depth: 2
    css: "/hpc/pmc_kuiper/Wilms/METHYLATION/doc.css"
date: "`r Sys.Date()`"
author: "Franziska Oberhammer"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Environment Setup and Library Loading
We load the required packages for DNA methylation analysis, quality control (sesame), batch effect evaluation, visualization, and general data manipulation.

```{r load_packages, warning=FALSE, message=FALSE}
library(circlize)
library(cowplot)
library(details)
library(eulerr)
library(ggplot2)
library(ggpubr)
library(ggthemes)
library(grid)
library(gridExtra)
library(patchwork)
library(plotly)
library(purrr)
library(reactable)
library(readxl)
library(RColorBrewer)
library(sesame)
library(sesameData)
library(summarytools)
library(sva)
library(tidyverse)
library(qs)

```


Important Prerequisites & Technical Notes:

- R Version: EPICv2 preprocessing requires R version 4.4.x or higher.
- QC Note: Quality control filtering relies on standard deviation thresholds computed from the sample population. Because samples failing QC are excluded prior to public upload, recalculating metrics on a subsetted dataset will result in different standard deviations and potentially different filtering results.

# Global Parameters and Settings
We establish directory paths and execution flags to govern script behaviour (re-running computationally expensive steps vs. loading pre-cached data).

```{r configuration, results='hide', cache=FALSE}
parameters <- list()

# Project Directory Paths ------------------------------------------------------
parameters$PROJECT_FOLDER          <- "/hpc/pmc_kuiper/Wilms/METHYLATION" 
parameters$CODE_FOLDER             <- file.path(parameters$PROJECT_FOLDER, "CODE") 
parameters$dataFolderInternal      <- file.path(parameters$PROJECT_FOLDER, "DATA/PMCPR001ACC/") 
parameters$dataFolderExternal      <- file.path("/hpc/pmc_kuiper/Wilms/", "EXTERNAL_DATA/2025_Treger_CancerDiscovery_SomaticFootprintPredisposition/DATA/METHYLATION_EPIC/GSE292896_idat") 
parameters$ANALYSIS_FOLDER         <- file.path(parameters$PROJECT_FOLDER, "ANALYSIS") 
parameters$PROCESSED_DATA_FOLDER   <- file.path(parameters$ANALYSIS_FOLDER, "PROCESSED_DATA") 
parameters$DATA_FOLDER             <- file.path(parameters$PROJECT_FOLDER, "DATA")

parameters$sampleSheetPathInternal <- file.path(parameters$PROJECT_FOLDER, "ANALYSIS/PROCESSED_DATA/sample_sheet_updated_20251015.rds")
parameters$sampleSheetPathExternal <- file.path("/hpc/pmc_kuiper/Wilms/EXTERNAL_DATA/2025_Treger_CancerDiscovery_SomaticFootprintPredisposition/DATA/sample_sheet_externalValidationCohort_preped.rds")



# Computational Workflow Flags -------------------------------------------------
# Processing raw IDATs takes quite long.
# Set these switches to TRUE only when recalculation or re-saving is needed.
CHANGE_SAMPLESHEET <- FALSE
RERUN_PROCESSING   <- FALSE
SAVE_PLOTS         <- FALSE
JUST_LOAD_BETAS    <- TRUE

# External Scripts and Palettes ------------------------------------------------
source(file.path(parameters$CODE_FOLDER, "00_theme_Publication.R"))
source(file.path(parameters$CODE_FOLDER, "00_WT_colour_palettes.R"))
source(file.path(parameters$CODE_FOLDER, "00_WT_helper_functions.R"))
```

# Overview of available samples

```{r}
sample_sheets <- list()

# internal sample sheet ------------
sample_sheets$sample_sheet_internal <- readRDS(parameters$sampleSheetPathInternal)$SAMPLESHEET
filter(Cohort == "Test")

# external sample sheet ------------
sample_sheets$sample_sheet_external  <- readRDS(parameters$sampleSheetPathExternal) %>% 
  rename(Sex = Gender)

# combine sample sheets -------------------------------------
sample_sheets$sample_sheet_combined <- full_join(sample_sheets$sample_sheet_internal, sample_sheets$sample_sheet_external)

```

# Quality Control (SeSAMe)
We import raw IDAT intensity files into SeSAMe SigSet objects (sdfs).


```{r SeSAMe load}
# Load or import raw IDAT files into SeSAMe objects
## Internal validation cohort ------------
if (CHANGE_SAMPLESHEET == TRUE) {
  all_prefixes_internal <- searchIDATprefixes(parameters$dataFolderInternal)
  
  selected_samples_internal <- sample_sheets$sample_sheet_internal$Methylation_barcode
  selected_prefixes_internal <- all_prefixes_internal[
    basename(all_prefixes_internal) %in% selected_samples_internal
  ]
  
  # this has to be done on R 4.4.2
  sdfs_internal <- lapply(selected_prefixes_internal, readIDATpair)
  
  qsave(
    file = file.path(parameters$PROCESSED_DATA_FOLDER, "sesame_raw_ValidationInternal.qs"),
    x = sdfs_internal
  )
  
} else {
  sdfs_internal <- qread(file.path(parameters$PROCESSED_DATA_FOLDER, "sesame_raw_ValidationInternal.qs"))
}


## External validation cohort ------------
if (CHANGE_SAMPLESHEET == TRUE) {
  all_prefixes_external <- searchIDATprefixes(parameters$dataFolderExternal)
  
  selected_samples_external <- sample_sheets$sample_sheet_external$Basename
  selected_prefixes_external <- all_prefixes_external[
    basename(all_prefixes_external) %in% selected_samples_external
  ]
  
  sdfs_external <- lapply(selected_prefixes_external, readIDATpair)
  
  qsave(
    file = file.path(parameters$PROCESSED_DATA_FOLDER, "sesame_raw_ValidationExternal.qs"),
    x = sdfs_external
  )
  
} else {
  sdfs_external <- qread(file.path(parameters$PROCESSED_DATA_FOLDER, "sesame_raw_ValidationExternal.qs"))
}

# Map array barcodes to unique Sample IDs
## Internal validation cohort ------------
names(sdfs_internal) <- sample_sheets$sample_sheet_internal$Internal_Patient_ID[match(names(sdfs_internal), sample_sheets$sample_sheet_internal$Methylation_barcode)]

### External  validation cohort ------------
names(sdfs_external) <- sample_sheets$sample_sheet_external$Internal_Patient_ID[match(names(sdfs_external), sample_sheets$sample_sheet_external$Methylation_barcode)]


# Combine validation cohorts
sdfs <- c(sdfs_internal, sdfs_external)



```



### SeSAMe Quality Metrics Computation
We compute key array quality parameters including probe count statistics, signal detection rates, intensity distributions, channel switching, and dye bias.
We plot them for each EPIC array version individually.

```{r QC_calculation}
# 1. Compute Individual Metric Sets --------------------------------------------

# Probe counts
if(CHANGE_SAMPLESHEET==TRUE){
  qcs_numProbes = openSesame(sdfs, prep="", func=sesameQC_calcStats, funs="numProbes")
  qsave(qcs_numProbes, file.path(parameters$ANALYSIS_FOLDER, "QC", "validationCohort", "qcs_numProbes_validation.qs"))
} else {
  qcs_numProbes <- qread(file.path(parameters$ANALYSIS_FOLDER, "QC", "validationCohort", "qcs_numProbes_validation.qs"))
}

qcs_numProbes_all <-do.call(rbind, lapply(qcs_numProbes, as.data.frame))

# Signal detection p-values
if(CHANGE_SAMPLESHEET==TRUE){
  qcs_detection = openSesame(sdfs, prep="", func=sesameQC_calcStats, funs="detection")
  qsave(qcs_detection, file.path(parameters$ANALYSIS_FOLDER, "QC", "validationCohort", "qcs_detection_validation.qs"))
} else {
  qcs_detection <- qread(file.path(parameters$ANALYSIS_FOLDER, "QC", "validationCohort", "qcs_detection_validation.qs"))
}

qcs_detection_all <-do.call(rbind, lapply(qcs_detection, as.data.frame))

# Signal intensities
if(CHANGE_SAMPLESHEET==TRUE){
  qcs_intensity = openSesame(sdfs, prep="", func=sesameQC_calcStats, funs="intensity")
  qsave(qcs_intensity, file.path(parameters$ANALYSIS_FOLDER, "QC", "validationCohort", "qcs_intensity_validation.qs"))
} else {
  qcs_intensity <- qread(file.path(parameters$ANALYSIS_FOLDER, "QC", "validationCohort", "qcs_intensity_validation.qs"))
}
qcs_intensity_all <-do.call(rbind, lapply(qcs_intensity, as.data.frame))

# Channel switching metrics
if(CHANGE_SAMPLESHEET==TRUE){
  qcs_channel = openSesame(sdfs, prep="", func=sesameQC_calcStats, funs="channel")
  qsave(qcs_channel, file.path(parameters$ANALYSIS_FOLDER, "QC", "validationCohort", "qcs_channel_validation.qs"))
} else {
  qcs_channel <- qread(file.path(parameters$ANALYSIS_FOLDER, "QC", "validationCohort", "qcs_channel_validation.qs"))
}
qcs_channel_all <-do.call(rbind, lapply(qcs_channel, as.data.frame))

# Dye bias metrics
if(CHANGE_SAMPLESHEET==TRUE){
  qcs_dyeBias = openSesame(sdfs, prep="", func=sesameQC_calcStats, funs="dyeBias")
  qsave(qcs_dyeBias, file.path(parameters$ANALYSIS_FOLDER, "QC", "validationCohort", "qcs_dyeBias_validation.qs"))
} else {
  qcs_dyeBias <- qread(file.path(parameters$ANALYSIS_FOLDER, "QC", "validationCohort", "qcs_dyeBias_validation.qs"))
}
qcs_dyeBias_all <- do.call(rbind, lapply(qcs_dyeBias, as.data.frame))


# 2. Merge QC Metrics with Sample Metadata --------------------------------------
stopifnot(rownames(qcs_dyeBias_all) == rownames(qcs_channel_all))
stopifnot(rownames(qcs_numProbes_all) == rownames(qcs_intensity_all)) 
stopifnot(rownames(qcs_channel_all) == rownames(qcs_intensity_all) )
stopifnot(rownames(qcs_detection_all) == rownames(qcs_intensity_all) )

qcs_all <- cbind(qcs_dyeBias_all,
                 qcs_channel_all, 
                 qcs_numProbes_all, 
                 qcs_intensity_all, 
                 qcs_detection_all)

qcs_all_with_mastersamplesheet <- left_join(qcs_all %>% rownames_to_column(var = "Internal_Patient_ID"),
                                            sample_sheets$sample_sheet_combined, by = "Internal_Patient_ID")

```



```{r QC_plots}
# Format QC metrics into long structure for interactive plotting
qcs_all_with_mastersamplesheet_long <- qcs_all_with_mastersamplesheet[,c(1:45, 98)] %>% 
  dplyr::select(-c("na_intensity_ig", "na_intensity_ir", "num_probes", "num_probes_II", "num_probes_IR", "num_probes_IG", "num_probes_cg", "num_probes_ch", "num_probes_rs", "na_intensity_M", "na_intensity_U", "na_intensity_ii", "num_dtna", "frac_dtna", "Internal_Patient_ID_Group" )) %>% 
  reshape2::melt( id.vars = c("PMCID", "Internal_Patient_ID", "Sample_Group", "EPIC_Version")
  )


```

### Reference: Evaluated QC Metrics Overview

Dye Bias
| QC | Metric | Description |
| ---| --- | --- |
| Dye Bias | `medR` | Median Inf.I Intens. Red |
| Dye Bias | `medG` | Median Inf.I Intens. Grn |
| Dye Bias | `topR` | Median of Top 20 Inf.I Intens. Red |
| Dye Bias | `topG` | Median of Top 20 Inf.I Intens. Grn |
| Dye Bias | `RGratio` | Ratio of Red-to-Grn median Intens. |
| Dye Bias | `RGdistort` | Ratio of Top vs. Global R/G Ratios |


Colour Channel
| QC | Metric | Description |
| ---| --- | --- |
| Colour Channel | `InfI_switch_R2R` | N. Inf.I Probes Red -> Red | 
| Colour Channel | `InfI_switch_G2G` | N. Inf.I Probes Grn -> Grn | 
| Colour Channel | `InfI_switch_R2G` | N. Inf.I Probes Red -> Grn | 
| Colour Channel | `InfI_switch_G2R` | N. Inf.I Probes Grn -> Red | 


Signal Intensity
| QC | Metric | Description |
| ---| --- | --- |
| Signal Intensity | `mean_intensity` |  Mean sig. intensity | 
| Signal Intensity | `mean_intensity_MU` |  Mean sig. intensity (M+U) |
| Signal Intensity | `mean_ii` |  Mean sig. intensity (Inf.II) |
| Signal Intensity | `mean_inb_grn` |  Mean sig. intens.(I.Grn IB) |
| Signal Intensity | `mean_inb_red` |  Mean sig. intens.(I.Red IB) |
| Signal Intensity | `mean_oob_grn` |  Mean sig. intens.(I.Grn OOB) |
| Signal Intensity | `mean_oob_red` |  Mean sig. intens.(I.Red OOB) |
| Signal Intensity | `na_intensity_M` |  N. NA in M (all probes) |
| Signal Intensity | `na_intensity_U` |  N. NA in U (all probes) |
| Signal Intensity | `na_intensity_ig` |  N. NA in raw intensity (IG) |
| Signal Intensity | `na_intensity_ir` |  N. NA in raw intensity (IR) |
| Signal Intensity | `na_intensity_ii` |  N. NA in raw intensity (II) |

Detection
| QC | Metric | Description |
| ---| --- | --- |
| Detection | `num_dtna` | N. Probes w/ Missing Raw Intensity |
| Detection | `frac_dtna` | % Probes w/ Missing Raw Intensity |
| Detection | `num_dt` | N. Probes w/ Detection Success |
| Detection | `frac_dt` | % Detection Success |
| Detection | `num_dt_mk` | N. Detection Succ. (after masking) |
| Detection | `frac_dt_mk` | % Detection Succ. (after masking) |
| Detection | `num_dt_cg` | N. Probes w/ Detection Success (cg) |
| Detection | `frac_dt_cg` | % Detection Success (cg) |
| Detection | `num_dt_ch` | N. Probes w/ Detection Success (ch) |
| Detection | `frac_dt_ch` | % Detection Success (ch) |
| Detection | `num_dt_rs` | N. Probes w/ Detection Success (rs) |
| Detection | `frac_dt_rs` | % Detection Success (rs) |

### QC Plots: EPIC v1 Arrays

```{r QC_plots_EPICV1, warning=FALSE}
# plot EPICv1
ggplotly(
  ggplot(
    qcs_all_with_mastersamplesheet_long %>% filter(EPIC_Version == "EPIC_V1"),
    aes(x = variable, y = value, fill = variable))+
    geom_violin()+
    geom_jitter(aes(text = paste("Intenral Patient ID:", Internal_Patient_ID)), height = 0)+
    facet_wrap(~variable, scales = "free")+
    xlab("")+
    ggtitle("EPIC v1")
  , tooltip=c("text", "x", "y")) %>% 
  style(showlegend = FALSE)
```

### QC Plots: EPIC v2 Arrays

```{r QC_plots_EPICV2, warning=FALSE}
# plot EPICv2
ggplotly(
  ggplot(
    qcs_all_with_mastersamplesheet_long %>% filter(EPIC_Version == "EPIC_V2"),
    aes(x = variable, y = value, fill = variable))+
    geom_violin()+
    geom_jitter(aes(text = paste("Intenral Patient ID:", Internal_Patient_ID)), height = 0)+
    facet_wrap(~variable, scales = "free")+
    xlab("")+
    ggtitle("EPIC v2")
  , tooltip=c("text", "x", "y")) %>% 
  style(showlegend = FALSE)
```

## Outlier Calculation and Sample Filtering
We evaluate samples against a 3 Standard Deviation threshold across all computed metrics. Samples failing >10% of total QC metrics are flagged for exclusion.
Because the 3–standard deviation threshold is calculated across all computed metrics, re-running QC on only the remaining samples would recalculate the baseline and inappropriately flag previously approved samples. Since all uploaded samples have already passed initial QC (failing <10% of metrics), this step should be skipped to preserve the dataset.

```{r QC_final_calculations}
# Calculate platform-specific metric means and standard deviations
qc_summary_df <- qcs_all_with_mastersamplesheet_long %>%
  group_by(EPIC_Version, variable) %>%
  summarise(mean=mean(value), sd=sd(value))

# Standard Deviation boundary helper
is_within_3sd <- function(value, mean, sd) {
  abs(value - mean) <= 3 * sd
}

# Flag metric passes/failures per sample
qcs_all_with_mastersamplesheet_long$QC_PASS_variable <- NA

for (i in 1: dim(qcs_all_with_mastersamplesheet_long)[1]){
  EPIC_V_temp <- qcs_all_with_mastersamplesheet_long[i, "EPIC_Version"]
  var_temp <-  qcs_all_with_mastersamplesheet_long[i, "variable"] %>% as.character()
  value_temp <- qcs_all_with_mastersamplesheet_long[i, "value"]
  qc_temp <- qc_summary_df %>% filter(EPIC_Version==EPIC_V_temp, variable == var_temp)
  qc_pass <- is_within_3sd(value = value_temp,
                           mean = qc_temp$mean,
                           sd = qc_temp$sd)
  
  qcs_all_with_mastersamplesheet_long[i,"QC_PASS_variable"] <- qc_pass
}

# Pivot back to wide format and quantify total failures
QCtests <- reshape2::dcast(qcs_all_with_mastersamplesheet_long, formula = Internal_Patient_ID ~ variable, value.var = "QC_PASS_variable")
colnames(QCtests) <- paste0(colnames(QCtests), "_QC")
QCtests <- QCtests %>%
  column_to_rownames("Internal_Patient_ID_QC") %>%
  mutate(QC_metrixs_failed = rowSums(!.)) %>%  #get number of qc fails per sample
  rownames_to_column("Internal_Patient_ID")

# Merge QC metrics back into metadata
qcs_all_with_mastersamplesheet <- left_join(qcs_all_with_mastersamplesheet, QCtests)

# Mark overall failure if more than 10% of individual tests fail
number_of_QC_columns <- length(colnames(qcs_all_with_mastersamplesheet)[grepl(colnames(qcs_all_with_mastersamplesheet), pattern  = "_QC")])

qcs_all_with_mastersamplesheet$QC_FAIL <- NA

qcs_all_with_mastersamplesheet <- qcs_all_with_mastersamplesheet %>%
  mutate(QC_FAIL = ifelse(QC_metrixs_failed > number_of_QC_columns/10, "FAIL", QC_FAIL))

sample_sheets$sample_sheet_combined_QC <- left_join(sample_sheets$sample_sheet_combined, qcs_all_with_mastersamplesheet %>%
                                                      dplyr::select(Internal_Patient_ID, QC_FAIL))

sample_sheets$sample_sheet_combined_QC %>% filter(!is.na(QC_FAIL))
```


```{r cleanGlobalEnviron3}
# Clean temporary QC calculation objects
rm(qcs_numProbes, qcs_detection, qcs_intensity, qcs_channel, qcs_dyeBias, qcs_dyeBias_all, qcs_channel_all, qcs_numProbes_all, qcs_intensity_all, qcs_detection_all, qcs_all, qcs_all_with_mastersamplesheet, qcs_all_with_mastersamplesheet_long, QCtests, qc_temp , qc_summary_df, i, EPIC_V_temp, number_of_QC_columns, value_temp, var_temp, qc_pass)
```

## Bisulfite Conversion Quality Control

We quantify residual incomplete bisulfite conversion using GCT scores calculated from C/T-extension probes (Zhou et al. 2017). Values close to 1.0 indicate optimal conversion.

```{r}
#Split up in EPICv1 and EPICv2
sdfs_V1 <- sdfs[sample_sheets$sample_sheet_combined %>% filter(EPIC_Version == "EPIC_V1") %>% pull(Internal_Patient_ID)]
sdfs_V2 <- sdfs[sample_sheets$sample_sheet_combined  %>% filter(EPIC_Version == "EPIC_V2") %>% pull(Internal_Patient_ID)]

# 1. Compute Bisulfite Conversion for EPIC v1 ----------------------------------
bisConversionControl_result_V1 <- lapply(setNames(names(sdfs_V1), names(sdfs_V1)), function(i) {
  return(bisConversionControl(sdfs_V1[[i]]))
})

# 2. Compute Bisulfite Conversion for EPIC v2 ----------------------------------
EPICv2_manifest <- read.csv(file.path(parameters$DATA_FOLDER, "BIN", "EPICv2.hg38.manifest.tsv"), sep = "\t")

extR = EPICv2_manifest[!is.na(EPICv2_manifest$nextBase) & EPICv2_manifest$nextBase=="R", "Probe_ID"]
extA = EPICv2_manifest[!is.na(EPICv2_manifest$nextBase) & EPICv2_manifest$nextBase=="A", "Probe_ID"]

# if this doesnt work, check that sesame >1.21.13
bisConversionControl_result_V2 <- lapply(setNames(names(sdfs_V2), names(sdfs_V2)), function(i) {
  return(bisConversionControl(sdfs_V2[[i]], extR = extR, extA = extA))
})

# 3. Combine and Evaluate Outliers ---------------------------------------------
bisConversionControl_result_both <- c(bisConversionControl_result_V1, bisConversionControl_result_V2)

bisConversionControl_result_both_df <- do.call(rbind, bisConversionControl_result_both) %>% as.data.frame() %>% rownames_to_column()
colnames(bisConversionControl_result_both_df) <- c("Internal_Patient_ID", "bisConversionControl")

bisConversionControl_result_both_df <- left_join(bisConversionControl_result_both_df, sample_sheets$sample_sheet_combined  %>% select(Internal_Patient_ID, EPIC_Version))

# Statistical summary
bisConversionControl_result_both_df_summary <- bisConversionControl_result_both_df %>% 
  summarise(mean=mean(bisConversionControl), sd=sd(bisConversionControl))

bisConversionControl_result_both_df <- bisConversionControl_result_both_df %>% 
  mutate(mean = bisConversionControl_result_both_df_summary$mean,
         sd = bisConversionControl_result_both_df_summary$sd)

# Interactive conversion score visualization
ggply <- ggplotly(ggplot(bisConversionControl_result_both_df, aes(x = Internal_Patient_ID, y = bisConversionControl))+
                    geom_line(aes(x = Internal_Patient_ID, y = mean, group = 1, colour = "mean"))+
                    geom_line(aes(x = Internal_Patient_ID, y = mean + 3*sd, group = 1, colour = "mean + 3sd"), linetype = "dashed")+
                    geom_bar(position = "dodge", stat = "identity")+
                    theme_minimal()+
                    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), legend.title = element_blank())+
                    labs(colour = ""))%>%
  style(textposition = "none")

ggply


# 4. Outlier Flagging and Metadata Updating ------------------------------------
bisConversionControl_result_both_df$bisConversionControl_testResult <- NA

for (r in 1:dim(bisConversionControl_result_both_df)[1]){
  bisConversionControl_temp <- bisConversionControl_result_both_df[r,"bisConversionControl"]
  bisConversionControl_mean <- bisConversionControl_result_both_df_summary$mean
  bisConversionControl_sd <- bisConversionControl_result_both_df_summary$sd
  
  is_within_3sd_temp <- is_within_3sd(bisConversionControl_temp, mean = bisConversionControl_mean, sd = bisConversionControl_sd)
  
  bisConversionControl_result_both_df[r,"bisConversionControl_testResult"] <- is_within_3sd_temp
}

sample_sheets$sample_sheet_combined_QC_bisulfite <- left_join(
  sample_sheets$sample_sheet_combined_QC %>% dplyr::select(-bisConversionControl_testResult), 
  bisConversionControl_result_both_df %>% dplyr::select(Internal_Patient_ID, bisConversionControl_testResult),
  by = "Internal_Patient_ID") %>% 
  mutate(QC_FAIL = ifelse(bisConversionControl_testResult == FALSE, "FAIL", QC_FAIL))

# Clean temporary bisulfite workflow objects
rm(bisConversionControl_result_both_df, ggply, bisConversionControl_result_V1, bisConversionControl_result_V2, EPICv2_manifest, sdfs_V1, sdfs_V2, bisConversionControl_mean, bisConversionControl_sd, bisConversionControl_temp, extA, extR, r)
```

# Preprocessing Pipeline Executions

We execute the target SeSAMe pre-processing string (QCDPB):

| One letter code| Function | Description | 
| --- | --- | --- |
| `Q` | `qualityMask` | Mask probes of poor design |
| `C` | `inferInfiniumIChannel` | Infer channel for Infinium-I probes |
| `D` | `dyeBiasNL` | Dye bias correction (non-linear) |
| `P` | `pOOBAH` | Detection p-value masking using oob |
| `B` | `noob` | Background subtraction using oob |


*oob = out of band*


```{r runPreprocess}
# Note: If preprocessCore fails during multithreading, reinstall via:
# BiocManager::install("preprocessCore", configure.args="--disable-threading")
if(RERUN_PROCESSING == TRUE){
  # this has to be done with R 4.3.1. With the newer version it did not work
  sdfs_preped <- lapply(sdfs, function(s) (openSesame(s, prep="QCDPB", func=NULL)))
  qsave(sdfs_preped, file.path(parameters$PROCESSED_DATA_FOLDER, "sesame_sdfs_preped_ValidationCombined.qs"))
} else {
  sdfs_preped <- qread(file.path(parameters$PROCESSED_DATA_FOLDER, "sesame_sdfs_preped_ValidationCombined.qs"))
}

```

## Preprocessing QC Evaluation

### Quality Check: Background Subtraction

We verify background subtraction by comparing pre- and post-processing Beta-value distributions across probe design types. Results are exported directly to PDF.


```{r QC_backgroundsubtraction}

if (CHANGE_SAMPLESHEET == TRUE){
  pdf(file = file.path(parameters$ANALYSIS_FOLDER, "QC", "validationCohort", "QC_plotBetaByDesign.pdf"), width = 10, height = 4)
  
  for (i in names(sdfs)){
    
    sdf_raw <- sdfs[[i]]
    sdf_processed <- sdfs_preped[[i]]
    
    # Calculate uniform density upper limits for plot alignment
    ylimit_p <- max(c(density(na.omit(sesame:::getBetas(sdf_raw)))$y,
                      density(na.omit(sesame:::getBetas(sdf_processed)))$y,
                      density(na.omit(sesame:::getBetas(sesame:::InfIR(sdf_raw))))$y,
                      density(na.omit(sesame:::getBetas(sesame:::InfIR(sdf_processed))))$y,
                      density(na.omit(sesame:::getBetas(sesame:::InfIG(sdf_raw))))$y,
                      density(na.omit(sesame:::getBetas(sesame:::InfIG(sdf_processed))))$y,
                      density(na.omit(sesame:::getBetas(sesame:::InfII(sdf_raw))))$y,
                      density(na.omit(sesame:::getBetas(sesame:::InfII(sdf_processed))))$y
    ))
    
    # adjust a bit so nothing is cut of
    ylimit_plot <- ylimit_p+ylimit_p*0.1
    
    # Generate side-by-side comparative density plots # function is adapted from sesame. can be found in helper functions
    raw_plot <- QC_plotBetaByDesign(sdf = sdf_raw, ggtitle("raw"))+
      theme(legend.position="none") +
      ylim(0, ylimit_plot)
    
    processed_plot <- QC_plotBetaByDesign(sdf = sdf_processed, ggtitle("processed"))+
      ylim(0, ylimit_plot)
    
    final_plot <- raw_plot + processed_plot + plot_annotation(i)
    
    print(final_plot)
  }
  dev.off()
  # clean global environment
  rm(i, ylimit_p, ylimit_plot, final_plot, processed_plot, sdf_processed, sdf_raw, raw_plot)
  
}


```

### Quality Check: Dye Bias Correction

We verify linear channel equalization by inspecting Red-Green QQ plots before and after correction.
We use linear scaling:
"Dye bias refers to the difference in signal intensity between the two color channel. SeSAMe offers two flavors of dye bias correction: linear scaling (dyeBiasCorr) and nonlinear scaling (dyeBiasCorrTypeINorm). Linear scaling equalize the mean of all probes from the two color channel." -sesame

```{r QC_qqplot}

# Generate QQ plots comparing channel distributions pre- and post-correction
QQ_plots <- lapply(setNames(names(sdfs), names(sdfs)), function(i) {
  raw_plot <- QC_plotRedGrnQQ(sdfs[[i]], main = "raw")
  processed_plot <- QC_plotRedGrnQQ(sdfs_preped[[i]], main = "processed")
  
  final_plot <- raw_plot + processed_plot + plot_annotation(paste("R-G QQ Plot:", i))
  return(final_plot)
})

# Save output plots to disk
if(SAVE_PLOTS){
  pdf(file = file.path(parameters$ANALYSIS_FOLDER, "QC","validationCohort", "QC_plotRedGrnQQ.pdf"), width = 10, height = 6)
  for (i in names(QQ_plots)){
    print(QQ_plots[[i]])
  }
  dev.off()
}

# clean global environment
rm(QQ_plots)

```

```{r cleanGlobalEnvironment}
# clean raw SigSet object
rm(sdfs)
```


# Beta Value Extraction and Cross-Platform Merging
We extract methylation Beta values, collapsing multi-probe CPI suffixes on EPIC v2 to prefix-level target CpG identifiers, and harmonise shared CpG content between EPIC v1 and v2 platforms.

```{r betas}


if (JUST_LOAD_BETAS == FALSE){
  if(RECALCULATE_BETAS){
    # Extract Beta values collapsing suffixes to cpg prefixes using minimum p-value selection
    betas_dfs <- lapply(setNames(names(sdfs_preped), names(sdfs_preped)), function(i) {
      print(i)
      tmp_obj <- sdfs_preped[[i]]
      res <- getBetas(tmp_obj, collapseToPfx = TRUE, collapseMethod = "minPval")
      return(res)
    })
    qsave(betas_dfs, file.path(parameters$PROCESSED_DATA_FOLDER, "betas_dfs_validationCohort.qs"))
  } else{
    betas_dfs <- qread(file.path(parameters$PROCESSED_DATA_FOLDER, "betas_dfs_validationCohort.qs"))
  }
  
  # Split datasets by platform version
  betas_dfs_V1 <- betas_dfs[sample_sheets$sample_sheet_combined_QC_bisulfite %>%
                              filter(EPIC_Version == "EPIC_V1") %>%
                              pull(Internal_Patient_ID)]
  
  betas_dfs_V2 <- betas_dfs[sample_sheets$sample_sheet_combined_QC_bisulfite %>%
                              filter(EPIC_Version == "EPIC_V2") %>%
                              pull(Internal_Patient_ID)]
  
  betas_V1 <- data.frame(betas_dfs_V1)
  betas_V2 <- data.frame(betas_dfs_V2)
  
  # Intersect probe sets common to both platforms
  CpGs_in_both <- intersect(rownames(betas_V1), rownames(betas_V2))
  
  # Merge matrix and align column order with primary sample sheet
  betas_combined <- cbind(betas_V1[CpGs_in_both,], betas_V2[CpGs_in_both,])
  
  betas_combined <- betas_combined[,match(sample_sheets$sample_sheet_combined_QC_bisulfite %>%
                                            pull(Internal_Patient_ID), colnames(betas_combined)
  )]
  
  
  qsave(betas_combined, file = file.path(parameters$PROCESSED_DATA_FOLDER, "betas_combined_validationCohort.qs"))
  
  # clean environment
  rm(betas_V1, betas_V2, betas_dfs_V1, betas_dfs_V2, betas_all, sdfs, betas_dfs)
  
} else {
  betas_combined <- qread(file = file.path(parameters$PROCESSED_DATA_FOLDER, "betas_combined_validationCohort.qs"))
}
```

# Probe masking

EPIC v1 probes were masked during SeSAMe preprocessing. 
At the time of this analysis, the automated probe masking function was only implemented for the Illumina HumanMethylationEPIC v1 array. Consequently, EPIC v2 probe masking was performed manually by retrieving the SeSAMe-recommended masking manifests and removing these CpG probes from the combined beta-value matrix.


```{r EPIC_V2_recomendedMask}
# Retrieve recommended exclusion probe list for EPIC v2
CpGs_to_mask_EPIC_V2 <- getMask(platform = "EPICv2", mask_names = "recommended")
names(CpGs_to_mask_EPIC_V2) <- CpGs_to_mask_EPIC_V2

# Lift over EPIC v2 probe IDs to target EPIC v1 probe standard
# in case of error, check the version of SeSAMe; it should be >1.21.13
CpGs_to_mask_EPIC_V2_liftover <- mLiftOver(CpGs_to_mask_EPIC_V2, target_platform = "EPIC", source_platform = "EPICv2") 

# Exclude identified probe positions from the Beta matrix
betas_combined_filtered <- betas_combined %>%
  filter(!rownames(betas_combined) %in% CpGs_to_mask_EPIC_V2_liftover)

rm(betas_combined, CpGs_to_mask_EPIC_V2,CpGs_to_mask_EPIC_V2_liftover)
```

# QC Sex Inference Verification
Sex inference was performed to verify reported sample identities. Samples demonstrating discordance between genetically inferred sex and phenotypic metadata were identified and flagged for downstream quality filtering.

```{r}
# QC: Custom Sex Inference for Merged EPIC v1/v2 Arrays
#
# Standard sesame::inferSex expects array-specific manifest formats. 
# To evaluate the combined dataset (intersected EPIC v1 and v2 probes, mapped 
# to v1 CpG identifiers), the underlying sex inference function was manually 
# adapted to target the overlapping sex-linked CpG set.
# Source implementation: https://github.com/zwdzwd/sesame/blob/devel/R/sex.R
inferSex2 <- function(betas, platform=NULL) {
  
  hypoMALE <- c(
    "cg21983484","cg23696472","cg11673471","cg01742836","cg13574945",
    "cg08059778","cg24186901","cg26023405","cg15977272","cg13023833",
    "cg20766178","cg20455959","cg26584339","cg13130271","cg13244998",
    "cg05872808","cg21290550","cg05806018","cg07861180","cg20015269",
    "cg12576145","cg10991108","cg02333283","cg16357225","cg25206026",
    "cg20749341","cg03773146","cg04872051","cg16590821","cg09520212",
    "cg22221554","cg11152253","cg23429746","cg00813156","cg25132467",
    "cg16221895","cg09307104","cg15165114","cg18998000","cg00723973",
    "cg06041068","cg10860619","cg09514431","cg07912337","cg03334316",
    "cg17399684","cg05534333","cg23493872","cg12413138","cg05374090",
    "cg27501007","cg08855111","cg21159768","cg16488754","cg12075609",
    "cg07446674","cg01342901","cg02869694","cg12277627","cg19992190",
    "cg10717149","cg14191108","cg01869765","cg26505478","cg23685102",
    "cg02195366","cg06334238","cg02615131","cg15565409","cg15693668",
    "cg03505772","cg00845806","cg26439324","cg12935118","cg18932686",
    "cg24264679","cg08782677","cg13649400","cg06779802","cg23554546",
    "cg23951868","cg00337921","cg08479532","cg00114625","cg03391801",
    "cg22776211","cg07674503","cg22452543","cg18140045","cg15450782",
    "cg07674075","cg06510592","cg21137943","cg24479484","cg27501723",
    "cg20439892","cg18107314","cg08405463","cg09146364","cg16894263")
  
  hyperMALE <- c(
    "cg26359388","cg02540440","cg11049634","cg22874828","cg09182733",
    "cg01123965","cg15822015","cg05130312","cg17072671","cg22655232",
    "cg05695959","cg21010298","cg06143713","cg22759686","cg11143827",
    "cg04303560","cg11717280","cg14372935","cg05533223","cg16405492",
    "cg15765801","cg08156775","cg24183173","cg21797452","cg03161453",
    "cg10474871","cg11516614","cg18813691","cg08614574","cg08456555",
    "cg16440909","cg13326840","cg16822540","cg03801901","cg09039264",
    "cg01383599","cg14931238","cg04071644","cg22208280","cg05559023",
    "cg23317607","cg26327984","cg07801607","cg06870560","cg24156613",
    "cg04101819","cg07422795","cg14261068","cg12622895","cg09192294",
    "cg26695278","cg12653510","cg03554089","cg11166197","cg04032096",
    "cg25047306","cg07818713","cg21258987","cg07981033","cg14492530",
    "cg18157587","cg12030638","cg17498624","cg01816615","cg08723064",
    "cg05193067","cg27167763","cg15521097","cg25456959","cg16576300",
    "cg07318999","cg22417678","cg22671388","cg23644934","cg00267352",
    "cg22223709","cg23698976","cg06780606","cg13920260","cg15861835",
    "cg10039267","cg12454245","cg22067189","cg00150874","cg08401365",
    "cg13781721","cg02931660","cg01316390","cg14746118","cg21294096",
    "cg11871337","cg00408231","cg09641151","cg05226646","cg11291200",
    "cg01109660","cg23607813","cg04624564","cg07452499","cg18123612")
  
  vals <- mean(betas[hyperMALE], na.rm = TRUE) - betas[hypoMALE]
  dd <- density(na.omit(vals))
  
  
  if (dd$x[which.max(dd$y)] > 0.4) {
    "Male"
  } else {
    "Female"
  }
}

# Run sex inference across all samples
pred.sex = apply(betas_combined_filtered, MARGIN = 2, function(x) {
  inferSex2(x)
})

pred.sex_df <- pred.sex %>% 
  as.data.frame() %>% 
  rownames_to_column()

colnames(pred.sex_df) <- c("Internal_Patient_ID", "Sex_predicted")

# Integrate sex predictions and flag discrepancies
sample_sheets$sample_sheet_combined_QC_bisulfite_predictSex <- left_join(sample_sheets$sample_sheet_combined_QC_bisulfite %>% dplyr::select(-Sex_predicted), pred.sex_df)

sample_sheets$sample_sheet_combined_QC_bisulfite_predictSex %>% 
  filter( tolower(Sex)!= tolower(Sex_predicted)) %>%  dplyr::select(Internal_Patient_ID, Sex, Sex_predicted) %>% View()

# Flag samples with discordant recorded vs. predicted sex
sample_sheets$sample_sheet_combined_QC_bisulfite_predictSex <- sample_sheets$sample_sheet_combined_QC_bisulfite_predictSex %>% 
  mutate(QC_FAIL = ifelse(tolower(Sex)!= tolower(Sex_predicted), "FAIL", QC_FAIL)) 

# clean global environment
rm(pred.sex_df, pred.sex)
```



# Sample Exclusion and Data Export
Finally, we remove samples that failed any quality control threshold (array metrics, bisulfite conversion, or sex concordance) and export the cleaned Beta value matrix and sample sheet

```{r}
# Filter sample sheet to include only QC-passed samples
sample_sheets$mastersamplesheet_after_QC <-  sample_sheets$sample_sheet_combined_QC_bisulfite_predictSex %>% 
  filter(is.na(QC_FAIL) )

# Subset Beta matrix to passing samples
betas_combined_filtered_QCSample <- betas_combined_filtered[,sample_sheets$mastersamplesheet_after_QC$Internal_Patient_ID]

# Export final processed objects
# qsave(betas_combined_filtered_QCSample, file = file.path(parameters$PROCESSED_DATA_FOLDER, "betas_V1V2_combined_probeFiltered_SampleQCFiltered_validationCohort.qs"))
# saveRDS(sample_sheets$mastersamplesheet_after_QC, file = file.path(parameters$PROCESSED_DATA_FOLDER, "mastersamplesheet_afterQC_validationCohort.rds"))
```


