---
title: "Manuscript_WT_Figure1"
author: "Franziska Oberhammer"
date: "2026-09-10"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

```{r libraries}

# DEPENDENCY MANAGEMENT & ENVIRONMENT PREPARATION
# Load all required R packages for differential methylation, differential expression,
# multi-omics data integration, unsupervised clustering, statistical testing, and visualization.

suppressPackageStartupMessages({
  library(DESeq2)
  library(cowplot)
  library(ggExtra)
  library(Cairo)
  library(ggpubr)
  library(limma)
  library(genefilter)
  library(corrplot)
  library(dendextend)
  library(viridis)
  library(EnhancedVolcano)
  library(plotly)
  library(tidyverse)
  library(circlize)
  library(ggalluvial)
  library(qs)
  library(ConsensusClusterPlus)
  library(gridExtra)
  library(RColorBrewer)
  library(ggsignif)
  library(dplyr)
  library(rstatix)
  library(officer)
  library(ggplot2)
  library(cluster)
  library(rcompanion)
  library(mclust)
  library(aricode)
})
```


```{r Configuration}
# PATH & PARAMETER CONFIGURATION
# Defines input/output project paths and explicit file references for 
# processed DNA methylation beta matrices, RNA-seq counts, and clinical sample metadata.

parameters <- list()

# Configuration ----------------------------------------------------------------
# NOTE: Update directory paths below to point to local or server environment locations
parameters$WILMS_FOLDER          <- "/hpc/pmc_kuiper/Wilms"
parameters$PROJECT_FOLDER        <- file.path(parameters$WILMS_FOLDER, "METHYLATION")
parameters$CODE_FOLDER           <- file.path(parameters$PROJECT_FOLDER, "CODE")
parameters$DATA_FOLDER           <- file.path(parameters$PROJECT_FOLDER, "DATA")
parameters$ANALYSIS_FOLDER       <- file.path(parameters$PROJECT_FOLDER, "ANALYSIS")
parameters$PROCESSED_DATA_FOLDER <- file.path(parameters$ANALYSIS_FOLDER, "PROCESSED_DATA")
parameters$RESULTS_FOLDER        <- file.path(parameters$PROJECT_FOLDER, "RESULTS")

# Project Input File Paths
parameters$CURRENT_SAMPLESHEET          <- file.path(parameters$PROCESSED_DATA_FOLDER, 
                                                     "samplesheet_afterQC_onlyMethSamples.rds")
parameters$CURRENT_BETAS                <- file.path(parameters$PROCESSED_DATA_FOLDER,
                                                     "betas_V1V2_combined_probeFiltered_SampleQCFiltered_SNPsQC_postMortemAndVersionFiltered_SexMissmatchFiltered.qs")
parameters$CURRENT_BETAS_batchCorrected <- file.path(parameters$PROCESSED_DATA_FOLDER,
                                                     "betas_V1V2_combined_probeFiltered_SampleQCFiltered_SNPsQC_postMortemAndVersionFiltered_batchCorrected.qs")
parameters$CURRENT_COUNTS               <- file.path("/hpc/pmc_kuiper/Wilms/DATA/RNAseq", "matrices/Wilms.all.samples.feature.counts.20250626.Counts.tab.txt")



# Source auxiliary scripts and themes -----------------------------------------------------
source(file.path(parameters$CODE_FOLDER, "00_theme_Publication.R"))
source(file.path(parameters$CODE_FOLDER, "00_WT_colour_palettes.R"))
source(file.path(parameters$CODE_FOLDER, "00_WT_helper_functions.R"))
```




```{r LoadData}
# Load sample metadata, methylation beta matrices, clustering objects, and RNA-seq counts
sample_sheet         <- readRDS(parameters$CURRENT_SAMPLESHEET)$SAMPLESHEET %>% as.data.frame()
betas                <- qread(parameters$CURRENT_BETAS)
betas_batchCorrected <- as.data.frame(qread(parameters$CURRENT_BETAS_batchCorrected))
counts_df            <- read.csv2(parameters$CURRENT_COUNTS, sep = "\t", row.names = "FEATURE")
```


```{r PrepData}
# Preprocess methylation beta matrices (keep only train cohort samples)
betas <- betas %>% 
  dplyr::select(-c("CI_V1", "BC", "EB"))

betas_batchCorrected <- betas_batchCorrected %>% 
  dplyr::select(-c("EB")) #CI_V1 and BC already removed in batch corrected data

# Standardize sample column names by removing version suffixes
colnames(betas) <- sub("_.*", "", colnames(betas))
colnames(betas_batchCorrected) <- sub("_.*", "", colnames(betas_batchCorrected))

# Verify sample ordering matches the sample sheet
stopifnot(colnames(betas) == sample_sheet$Internal_Patient_ID)
stopifnot(colnames(betas_batchCorrected) == sample_sheet$Internal_Patient_ID)


# Preprocess RNA-seq count matrix and align ordering -----------------------------------------------------------------------------------------
# filter samplesheet for only RNAseq data
sample_sheet_RNAseq_prep <- sample_sheet %>% 
  filter(!is.na(Used_RNA_ID)) %>% 
  filter(!Internal_Patient_ID %in% c("DB",  "DG",  "DI",  "DX",  "EB")) # failed QC (number of reads etc)

rownames(sample_sheet_RNAseq_prep) <- sample_sheet_RNAseq_prep$Internal_Patient_ID

# Reorder RNA-seq columns to match sample sheet order
RNA_colnames_ordered <- colnames(counts_df)[match(sample_sheet_RNAseq_prep$Used_RNA_ID, colnames(counts_df))]
counts_df <- counts_df[,RNA_colnames_ordered]

stopifnot(colnames(counts_df) == sample_sheet_RNAseq_prep$Used_RNA_ID)

# Map internal RNA IDs to M-Labels
colnames(counts_df) <- sample_sheet_RNAseq_prep$M_Label[match(colnames(counts_df), sample_sheet_RNAseq_prep$Used_RNA_ID)]

# clean workspace -----------
rm(RNA_colnames_ordered)
```


# Figure 1A Heatmap

## Clustering

```{r find_most_variable_cpgs_batchCorrected}
# Feature selection: isolate top 5,000 most variable CpGs across samples by row variance

NUMBER_OF_MOST_VARIABLE_CpGs <- 5000

betas_batchCorrected_mvCpGs <- head(betas_batchCorrected[
  order(matrixStats:::rowVars(
    betas_batchCorrected %>% as.matrix()),
    decreasing = TRUE),],
  NUMBER_OF_MOST_VARIABLE_CpGs)

#Change labels to M-labels
colnames(betas_batchCorrected_mvCpGs) <- sample_sheet[match(colnames(betas_batchCorrected_mvCpGs), sample_sheet$Internal_Patient_ID),]$M_Label
```

```{r}
# Scale and transpose transformed mvCpG matrix for distance and clustering calculations
betas_batchCorrected_mvCpGs_scaledTransformed <- scale(t(betas_batchCorrected_mvCpGs))

stopifnot(rownames(betas_batchCorrected_mvCpGs_scaledTransformed)== sample_sheet$M_Label)

```


```{r}
# Unsupervised consensus clustering via ConsensusClusterPlus across top variable CpGs
innerLinkage <- "ward.D2"
finalLinkage <- "ward.D2"
clusterAlg <- "hc"
distanceMeth <- "pearson"

results = ConsensusClusterPlus(
  betas_batchCorrected_mvCpGs_scaledTransformed%>% t(),
  maxK=20,
  reps=1000,
  pItem=0.8,
  pFeature=1,
  innerLinkage = innerLinkage,
  finalLinkage = finalLinkage,
  clusterAlg = clusterAlg,
  distance = distanceMeth,
  title = file.path(parameters$ANALYSIS_FOLDER,
                    "EXPLORATORY_ANALYSIS/clustering/ConsensusClusterPlus_5000mvCpGs_manuscript",
                    paste0("innerLinkage", innerLinkage,
                           "_finalLinkage", finalLinkage,
                           "_clusterAlg", clusterAlg,
                           "_distanceMeth", distanceMeth)),
  seed=1262118388.71279,
  # plot="pdf" # Uncomment to export consensus cluster decision plots
)

# Calculate item consensus (ICL) metrics
icl = calcICL(
  results,
  title = file.path(parameters$ANALYSIS_FOLDER,
                    "EXPLORATORY_ANALYSIS/clustering/ConsensusClusterPlus_5000mvCpGs_manuscript",
                    paste0("innerLinkage", innerLinkage,
                           "_finalLinkage", finalLinkage,
                           "_clusterAlg", clusterAlg,
                           "_distanceMeth", distanceMeth)),
  # plot = "pdf" # Uncomment to export plots
)
```


Evaluating the clustering metrics identified a five-cluster model as optimal.   


```{r}

# Extract consensus clustering hierarchy at k = 5 clusters
consensus_clustering_5 <- results[[5]]
consensus_clustering_5_tree <- consensus_clustering_5$consensusTree
consensus_clustering_5_tree$labels <- rownames(betas_batchCorrected_mvCpGs_scaledTransformed)

qsave(consensus_clustering_5_tree, 
      file = file.path(
        parameters$ANALYSIS_FOLDER,
        "EXPLORATORY_ANALYSIS/clustering/ConsensusClusterPlus_5000mvCpGs_manuscript",
        paste0("innerLinkage", innerLinkage,
               "_finalLinkage", finalLinkage,
               "_clusterAlg", clusterAlg,
               "_distanceMeth", distanceMeth), "clustering_tree_5clusters.qs"))

# clean global environment
rm(innerLinkage, finalLinkage, clusterAlg, distanceMeth)

```


```{r}
# Map numeric consensus clusters (k=5) to manuscript subgroup nomenclature (1a-2c)
clustering_result <- consensus_clustering_5$consensusClass %>%
  as.data.frame(nm = "DNAm_5_cluster") %>%
  rownames_to_column("M_Label") %>% 
  mutate(DNAm_5_cluster_letters = 
           case_when(
             DNAm_5_cluster == 1 ~ "2a",
             DNAm_5_cluster == 2 ~ "2c",
             DNAm_5_cluster == 3 ~ "1b",
             DNAm_5_cluster == 4 ~ "2b",
             DNAm_5_cluster == 5 ~ "1a",
           ))

sample_sheet_clustering <- left_join(sample_sheet %>% mutate(DNAm_5_cluster_letters = NULL), clustering_result)

#clean global environment
rm(clustering_result, results, icl)
```

## Heatmap

```{r}

stopifnot(colnames(betas_batchCorrected_mvCpGs) == names(consensus_clustering_5$consensusClass))

methHeatmap <- ComplexHeatmap::Heatmap(
  matrix = betas_batchCorrected_mvCpGs,
  use_raster = F,
  cluster_columns = as.dendrogram(consensus_clustering_5_tree),
  column_split = 5,
  
  show_row_names = F,
  show_column_names = T,

  row_title = "CpG",
  column_title = "Sample",
  column_title_side = "bottom",
  col = circlize::colorRamp2(seq(0, 1, 0.01), colours_WT$meth),
  name = "ß-value",
  heatmap_legend_param = list(
    direction = "horizontal",
    width = unit(4, "cm")
  ),
  column_names_gp = grid::gpar(fontsize = 5),
  row_names_gp = grid::gpar(fontsize = 5)
)

# Set figure dimensions for export
pdf_width <- 180/25.4   # Convert mm to inches
pdf_height <- 80/25.4   
dpi <- 300        # raster resolution

# Render Heatmap output
# CairoPDF(file.path(parameters$RESULTS_FOLDER, "FIGURE1", "1A_DNAm_clustering.pdf"), width = pdf_width, height = pdf_height) # convert to inches
ComplexHeatmap::draw(methHeatmap, heatmap_legend_side = "bottom")
# dev.off()


# High-res Raster Export
png_width <- pdf_width * dpi
png_height <- pdf_height * dpi

# CairoPNG(file.path(parameters$RESULTS_FOLDER, "FIGURE1", "1A_DNAm_clustering.png"), width = png_width, height = png_height, dpi = dpi)
ComplexHeatmap::draw(methHeatmap, heatmap_legend_side = "bottom"
                     # , show_heatmap_legend = FALSE, show_annotation_legend = FALSE
                     )  
# dev.off()


# write.table(betas_batchCorrected_mvCpGs, quote = FALSE, row.names = TRUE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE1", "RAW_DATA", "1A_DNAm_clustering.tsv"), sep = "\t")



```


# Figure 1B PCA RNAseq

```{r}
set.seed(42)

sample_sheet_RNAseq <- left_join(sample_sheet_RNAseq_prep %>% mutate(DNAm_5_cluster_letters = NULL), sample_sheet_clustering %>% 
                                   dplyr::select(Internal_Patient_ID, DNAm_5_cluster_letters)) 


# =====================================
# 1. NORMALIZATION PIPELINE
# =====================================

cat("Step 1: DESeq2 Normalization Pipeline\n")
cat("=====================================\n")

# Construct DESeq2 object with main experimental covariate design
dds <- DESeqDataSetFromMatrix(
  countData = counts_df,
  colData = sample_sheet_RNAseq,
  design = ~ Batch_RNAseq + Sex + DNAm_5_cluster_letters
)

cat("Original dataset dimensions:", dim(dds), "\n")

# Quality filtering: Remove low-expressed genes
# Keep genes with ≥10 counts in ≥3 samples (adjust based on sample size)
keep <- rowSums(counts(dds) >= 10) >= max(3, ncol(dds) * 0.1)
dds <- dds[keep, ]
cat("After filtering low-expressed genes:", nrow(dds), "genes retained\n")

# Estimate size factors (normalization factors)
dds <- estimateSizeFactors(dds)
cat("Size factors calculated:\n")
print(round(sizeFactors(dds), 3))

# Get normalized counts
norm_counts <- counts(dds, normalized = TRUE)
plot(colSums(norm_counts))

# =====================================
# 1b. REMOVE SAMPLES WITH EXTREME LIBRARY SIZES
# =====================================
# This was already performed on our raw data. RNAseq data from patient M222AAH was an extreme outlier in terms of library size. Therefore we had to remove it, and also did not upload it to EGA. This next step only needs to be performed if that sample is still in your data. 

if("M222AAH" %in% colnames(counts_df)){
  # Total counts per sample
  totals <- colSums(norm_counts)
  
  # Cutoffs (mean ± 3*SD)
  low <- mean(totals) - 3*sd(totals)
  high <- mean(totals) + 3*sd(totals)
  
  # Identify outlier sample IDs
  outliers <- names(totals[totals < low | totals > high])
  cat("Removing outlier samples (library size extreme):", outliers, "\n")
  
  # Remove outliers from DESeqDataSet
  dds <- dds[, !colnames(dds) %in% outliers]
  
  # Update sample sheet and normalized counts accordingly
  sample_sheet_RNAseq <- as.data.frame(colData(dds))
  norm_counts <- counts(dds, normalized = TRUE)
  
  # Quick check
  cat("Dataset dimensions after outlier removal (dds):", dim(dds), "\n")
  cat("Dataset dimensions after outlier removal (sample sheet):", dim(sample_sheet_RNAseq), "\n")
  cat("Dataset dimensions after outlier removal (normalised counts):", dim(norm_counts), "\n")
  
  # Visualize library sizes and highlight removed samples
  df_lib <- data.frame(
    Sample = names(totals),
    TotalCounts = totals,
    Status = ifelse(names(totals) %in% outliers, "Removed", "Kept")
  )
  
  # pdf(file.path(parameters$RESULTS_FOLDER, "FIGURES1", "100_RNAseqLibrarySize.pdf"), width = 8, height = 7)
  ggplot(df_lib, aes(x = Sample, y = TotalCounts, fill = Status)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = c("Kept" = "skyblue", "Removed" = "red")) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = "Library Sizes per Sample", y = "Normalized Counts")
  # dev.off()
  
}


# =====================================
# 2. VARIANCE STABILIZING TRANSFORMATION
# =====================================

cat("\nStep 2: Variance Stabilizing Transformation\n")
cat("==========================================\n")

# Apply VST transformation to eliminate mean-variance dependency across count range
vst_data <- vst(dds, blind = FALSE)
vst_mat <- assay(vst_data)

cat("Generating mean-variance plots...\n")

top_gene_vst <- which.max(rowMeans(vst_mat))
gene_name <- rownames(vst_mat)[top_gene_vst]

# Get normalized count data for comparison
count_data <- plotCounts(dds, gene = gene_name, intgroup = "DNAm_5_cluster_letters", returnData = TRUE, normalized = TRUE)
count_data$vst_counts <- vst_mat[gene_name, rownames(count_data)]

# Side-by-side comparison plot
p1 <- ggplot(count_data, aes(x = DNAm_5_cluster_letters, y = count, color = DNAm_5_cluster_letters)) +
  geom_jitter(width = 0.2, size = 3, alpha = 0.7) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +
  scale_y_log10() +
  labs(title = "DESeq2 Normalized Counts", y = "Normalized counts (log10)") +
  theme_minimal() + theme(legend.position = "none")+
  scale_color_manual(values = colours_WT$DNAm_clustering2)

p2 <- ggplot(count_data, aes(x = DNAm_5_cluster_letters, y = vst_counts, color = DNAm_5_cluster_letters)) +
  geom_jitter(width = 0.2, size = 3, alpha = 0.7) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +
  labs(title = "VST Transformed Counts", y = "VST counts") +
  theme_minimal() + theme(legend.position = "none")+
  scale_color_manual(values = colours_WT$DNAm_clustering2)


grid.arrange(p1, p2, ncol = 2, top = paste("Gene:", gene_name))



# =====================================
# 3. PRINCIPAL COMPONENT ANALYSIS
# =====================================

cat("\nStep 3: Principal Component Analysis\n")
cat("===================================\n")


vst_mat <- assay(vst_data)

# Remove technical batch and biological sex covariates using limma
covars <- model.matrix(~ Sex + Batch_RNAseq, colData(vst_data))[ , -1]
vst_corrected <- limma::removeBatchEffect(vst_mat, covariates = covars)

# Select top 500 most variable genes post-batch correction
topVar <- head(order(rowVars(vst_corrected), decreasing = TRUE), 500)
vst_top <- vst_corrected[topVar, ]

# Run PCA on adjusted expression values
pca_result <- prcomp(t(vst_top), center = TRUE, scale. = FALSE)


# Calculate variance explained by each PC
var_explained <- (pca_result$sdev^2) / sum(pca_result$sdev^2) * 100
cum_var_explained <- cumsum(var_explained)

cat("Variance explained by first 5 PCs:\n")
for(i in 1:min(5, length(var_explained))) {
  cat(sprintf("PC%d: %.1f%% (Cumulative: %.1f%%)\n", 
              i, var_explained[i], cum_var_explained[i]))
}

# Create PCA data frame
pca_df <- data.frame(
  Sample = rownames(pca_result$x),
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  PC3 = pca_result$x[, 3],
  PC4 = pca_result$x[, 4],
  PC5 = pca_result$x[, 5]
) %>%
  left_join(sample_sheet_RNAseq, by = c("Sample" = "M_Label"))

# PCA plots colored by different metadata variables ---------- 
plot_list <- list()

# Main PCA plot by cluster
# Define the PC pairs you want to plot
pc_pairs <- list(
  c(1, 2),
  c(1, 3), 
  c(2, 3),
  c(2,4)
)

# Loop over categorical variables and PC pairs to make PCA plots
for (pair in pc_pairs) {
  pcx <- pair[1]
  pcy <- pair[2]
  
    p <- ggplot(pca_df,  
                aes(x = .data[[paste0("PC", pcx)]], 
                    y = .data[[paste0("PC", pcy)]], 
                    color = .data[["DNAm_5_cluster_letters"]])
    ) +
      geom_point(size = 3, alpha = 0.8) +
  labs(
    #title = "PCA: Samples colored by Cluster",
    x = paste0("PC", pcx, " (", round(var_explained[pcx], 1), "%)"),
    y = paste0("PC", pcy, " (", round(var_explained[pcy], 1), "%)")
  ) +
  theme(legend.position = NULL) +
  theme_Publication()+
  scale_color_manual(values = colours_WT$DNAm_clustering2, name = "DNAm cluster")
      
    p_all <- ggMarginal(
  p,
  data = pca_df,
  groupColour = TRUE,
  groupFill   = TRUE,
  margins     = "both",
  type        = "density"
)
    # Save to plot_list with informative name
    plotname <- paste0("DNAm_clustering", "_PC", pcx, "_PC", pcy)
    plot_list[[plotname]] <- p_all
  }

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURE1", "1B_RNAseqPCA.pdf"), width = 8, height = 6)
plot_list[1] 
# dev.off()

# save raw data used to generate figures
# write.table(pca_df %>% dplyr::select(Sample, PC1, PC2, DNAm_5_cluster_letters), quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE1", "RAW_DATA", "1B_RNAseqPCA_data.tsv"), sep = "\t")


# Additional PCA plots for other categorical variables
categorical_vars <- c("Sex", "Age_diagnosis_log", "Batch_RNAseq", "Preoperative_chemotherapy", "Sample_Group", "Corrected_histology")
stopifnot(categorical_vars %in% colnames(sample_sheet_RNAseq))

# Loop over categorical variables and PC pairs to make PCA plots
for (vari in categorical_vars) {
  for (pair in pc_pairs) {
    pcx <- pair[1]
    pcy <- pair[2]
    
    p <- ggplot(
      pca_df, 
      aes(x = .data[[paste0("PC", pcx)]], 
          y = .data[[paste0("PC", pcy)]], 
          color = .data[[vari]])
    ) +
      geom_point(size = 3, alpha = 0.8) +

      labs(
        title = paste("PCA:", vari, "colored (PC", pcx, "vs PC", pcy, ")"),
        x = paste0("PC", pcx, " (", round(var_explained[pcx], 1), "%)"),
        y = paste0("PC", pcy, " (", round(var_explained[pcy], 1), "%)")
      ) +
      theme_minimal() +
      theme(legend.position = "bottom")
    
    # Save to plot_list with informative name
    plotname <- paste0(vari, "_PC", pcx, "_PC", pcy)
    plot_list[[plotname]] <- p
  }
}




# Display PCA plots
do.call(grid.arrange, c(plot_list[1:8],   ncol = 4))
do.call(grid.arrange, c(plot_list[9:16],   ncol = 4))
do.call(grid.arrange, c(plot_list[17:24],   ncol = 4))
do.call(grid.arrange, c(plot_list[25:32],   ncol = 4))
do.call(grid.arrange, c(plot_list[33:40],   ncol = 4))
do.call(grid.arrange, c(plot_list[41:48],   ncol = 4))
do.call(grid.arrange, c(plot_list[49:56],   ncol = 4))
do.call(grid.arrange, c(plot_list[57:64],   ncol = 4))

# =====================================
# 4. SCREE PLOT
# =====================================

cat("\nStep 4: Scree Plot Analysis\n")
cat("===========================\n")

scree_data <- data.frame(
  PC = 1:length(var_explained),
  Variance_Explained = var_explained,
  Cumulative_Variance = cum_var_explained
)

# Scree plot with elbow point identification
scree_plot <- ggplot(scree_data[1:min(20, nrow(scree_data)), ], 
                     aes(x = PC, y = Variance_Explained)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_line(color = "red", size = 1) +
  geom_point(color = "red", size = 2) +
  labs(
    title = "Scree Plot: Variance Explained by Principal Components",
    x = "Principal Component",
    y = "Variance Explained (%)"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = 1:min(20, nrow(scree_data)))

# Cumulative variance plot
cum_var_plot <- ggplot(scree_data[1:min(20, nrow(scree_data)), ], 
                       aes(x = PC, y = Cumulative_Variance)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "blue", size = 2) +
  geom_hline(yintercept = c(80, 90, 95), linetype = "dashed", alpha = 0.5) +
  labs(
    title = "Cumulative Variance Explained",
    x = "Principal Component",
    y = "Cumulative Variance (%)"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = 1:min(20, nrow(scree_data))) +
  ylim(0, 100)

# Display scree plots
# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURES1", "100_RNAseqScreePlot.pdf"), width = 8, height = 7)
grid.arrange(scree_plot, cum_var_plot, ncol = 1)
# dev.off()

# Identify optimal number of PCs (Kaiser criterion: eigenvalue > 1)
eigenvalues <- pca_result$sdev^2
optimal_pcs_kaiser <- sum(eigenvalues > 1)
cat(sprintf("Optimal PCs (Kaiser criterion): %d\n", optimal_pcs_kaiser))



# =====================================
# 5. PC CORRELATION ANALYSIS
# =====================================

cat("\nStep 5: PC Correlation with Continuous Variables\n")
cat("===============================================\n")

# Identify continuous variables in metadata
continuous_vars <- names(sample_sheet_RNAseq)[sapply(sample_sheet_RNAseq, is.numeric)]
cat("Continuous variables found:", paste(continuous_vars, collapse = ", "), "\n")

if(length(continuous_vars) > 0) {
  # Extract continuous variables for correlation
  cont_data <- sample_sheet_RNAseq[continuous_vars]
  
  # Calculate correlations between PCs and continuous variables
  pc_scores <- pca_result$x[, 1:min(10, ncol(pca_result$x))]
  
  # Use Spearman correlation (non-parametric, robust to outliers)
  correlation_method <- "spearman"
  
  # Calculate correlation matrix (rectangular: PCs x metadata variables)
  cor_matrix <- cor(pc_scores, cont_data, method = correlation_method, use = "complete.obs")
  
  # Calculate p-values for correlations with exact=FALSE to handle ties
  cor_pvalues <- matrix(NA, nrow = nrow(cor_matrix), ncol = ncol(cor_matrix))
  rownames(cor_pvalues) <- rownames(cor_matrix)
  colnames(cor_pvalues) <- colnames(cor_matrix)
  
  cat("Calculating correlation p-values...\n")
  for(i in 1:nrow(cor_matrix)) {
    for(j in 1:ncol(cor_matrix)) {
      # Use exact=FALSE to avoid ties warning and use normal approximation
      test_result <- cor.test(pc_scores[, i], cont_data[, j], 
                             method = correlation_method, exact = FALSE)
      cor_pvalues[i, j] <- test_result$p.value
    }
  }
  
  # Apply multiple testing correction (Benjamini-Hochberg FDR)
  cor_pvalues_adj <- matrix(p.adjust(as.vector(cor_pvalues), method = "BH"), 
                           nrow = nrow(cor_pvalues), ncol = ncol(cor_pvalues))
  rownames(cor_pvalues_adj) <- rownames(cor_pvalues)
  colnames(cor_pvalues_adj) <- colnames(cor_pvalues)
  
  # Create significance matrix for visualization (using adjusted p-values)
  sig_matrix <- ifelse(cor_pvalues_adj < 0.001, "***",
                       ifelse(cor_pvalues_adj < 0.01, "**",
                              ifelse(cor_pvalues_adj < 0.05, "*", "")))
  
  # CORRPLOT FOR NON-SQUARE MATRIX: Use order = "original" instead of "hclust"
  library(corrplot)
  corrplot(cor_matrix,
           method = "color",
           type = "full",
           order = "original",
           tl.col = "black",
           tl.srt = 45,
           title = paste("PC-Metadata Correlations (", stringr::str_to_title(correlation_method), ")", sep = ""),
           mar = c(0, 0, 2, 0),
           addCoef.col = "black",
           number.cex = 0.7,
           # Color scale
           col = colorRampPalette(c("#BB4444", "#EE9988", "#FFFFFF", "#77AADD", "#4477AA"))(200),
           # Add significance indicators
           p.mat = cor_pvalues_adj,
           sig.level = 0.05,
           insig = "blank",  # Blank out non-significant correlations
           is.corr = FALSE)  # tells corrplot this is not a square correlation matrix
  
  # Alternative: ggplot2 visualization with better control for rectangular matrices
  library(reshape2)
  cor_melt <- melt(cor_matrix)
  pval_melt <- melt(cor_pvalues_adj)
  cor_melt$pvalue <- pval_melt$value
  cor_melt$significance <- ifelse(cor_melt$pvalue < 0.001, "***",
                                 ifelse(cor_melt$pvalue < 0.01, "**",
                                        ifelse(cor_melt$pvalue < 0.05, "*", "")))
  
  # Enhanced ggplot with better aesthetics for PC-metadata correlations
  ggplot_corr <- ggplot(cor_melt, aes(x = Var2, y = Var1, fill = value)) +
    geom_tile(color = "white", linewidth = 0.5) +
    scale_fill_gradient2(low = "#2166AC", high = "#B2182B", mid = "white", 
                        midpoint = 0, limit = c(-1, 1), space = "Lab",
                        name = "Spearman\nCorrelation") +
    geom_text(aes(label = paste0(round(value, 2), "\n", significance)), 
              color = "black", size = 3.5, fontface = "bold") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.text.y = element_text(size = 10),
      axis.title = element_blank(),
      panel.grid = element_blank(),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 10),
      legend.position = "right"
    ) +
    labs(title = "Principal Component - Metadata Correlations",
         subtitle = "*, **, *** indicate FDR-corrected p < 0.05, 0.01, 0.001") +
    coord_fixed(ratio = 1) 
  
  print(ggplot_corr)
  
  # Optional: Create a clustered heatmap using pheatmap for better clustering visualization
  library(pheatmap)
  
  # For clustering, we need to handle the rectangular matrix differently
  # We can cluster rows (PCs) separately from columns (metadata variables)
  if(ncol(cor_matrix) > 2 && nrow(cor_matrix) > 2) {
    
    # Create annotation for significance
    annotation_colors <- list()
    
    pheatmap(cor_matrix,
             color = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
             breaks = seq(-1, 1, length.out = 101),
             cluster_rows = TRUE,    # Cluster PCs
             cluster_cols = TRUE,    # Cluster metadata variables  
             clustering_distance_rows = "correlation",
             clustering_distance_cols = "correlation",
             clustering_method = "complete",
             display_numbers = TRUE,
             number_format = "%.2f",
             fontsize_number = 8,
             main = "PC-Metadata Correlations (Clustered)",
             angle_col = 45,
             cellwidth = 30,
             cellheight = 20)
  }
  
  # Print significant correlations (using FDR-corrected p-values)
  cat("\nSignificant PC-metadata correlations (FDR-corrected p < 0.05):\n")
  significant_cors <- which(cor_pvalues_adj < 0.05, arr.ind = TRUE)
  if(nrow(significant_cors) > 0) {
    sig_results <- data.frame(
      PC = rownames(cor_matrix)[significant_cors[, 1]],
      Variable = colnames(cor_matrix)[significant_cors[, 2]],
      Correlation = cor_matrix[significant_cors],
      P_value = cor_pvalues[significant_cors],
      P_adjusted = cor_pvalues_adj[significant_cors]
    )
    sig_results <- sig_results[order(sig_results$P_adjusted), ]
    print(sig_results)
    
    # Export significant correlations
   #write.csv(sig_results, "Significant_PC_correlations.csv", row.names = FALSE)
  } else {
    cat("No significant correlations found after FDR correction.\n")
  }
  
  # Summary statistics
  cat(sprintf("\nCorrelation Analysis Summary:\n"))
  cat(sprintf("- Method: %s correlation\n", stringr::str_to_title(correlation_method)))
  cat(sprintf("- PCs analyzed: %d\n", nrow(cor_matrix)))
  cat(sprintf("- Continuous variables: %d\n", ncol(cor_matrix)))
  cat(sprintf("- Total correlations tested: %d\n", length(cor_matrix)))
  cat(sprintf("- Significant (raw p < 0.05): %d\n", sum(cor_pvalues < 0.05, na.rm = TRUE)))
  cat(sprintf("- Significant (FDR-corrected p < 0.05): %d\n", sum(cor_pvalues_adj < 0.05, na.rm = TRUE)))
  
} else {
  cat("No continuous variables found in metadata for correlation analysis.\n")
}


# =====================================
# 6. PC ASSOCIATION WITH CATEGORICAL VARIABLES
# =====================================

cat("\nStep 6: PC Associations with Categorical Variables\n")
cat("===================================================\n")

# Identify categorical variables in metadata
categorical_vars <- unique(c("DNAm_5_cluster_letters", categorical_vars))
categorical_vars <- categorical_vars[!categorical_vars %in% c("sizeFactor")]
cat("Categorical variables found:", paste(categorical_vars, collapse = ", "), "\n")

if(length(categorical_vars) > 0) {
  # Extract PC scores
  pc_scores <- pca_result$x[, 1:min(10, ncol(pca_result$x))]  # first 10 PCs
  sig_cat_results <- data.frame()
  
  for(vari in categorical_vars) {
    for(pc in colnames(pc_scores)) {
      # ANOVA: PC ~ categorical variable
      group <- factor(sample_sheet_RNAseq[[vari]])
      test <- aov(pc_scores[, pc] ~ group)
      pval <- summary(test)[[1]][["Pr(>F)"]][1]
      
      sig_cat_results <- rbind(sig_cat_results, 
                               data.frame(PC = pc,
                                          Variable = vari,
                                          P_value = pval))
    }
  }
  
  # Multiple testing correction
  sig_cat_results$P_adjusted <- p.adjust(sig_cat_results$P_value, method = "BH")
  
  # Keep only significant associations
  sig_cat_results <- sig_cat_results[sig_cat_results$P_adjusted < 0.05, ]
  sig_cat_results <- sig_cat_results[order(sig_cat_results$P_adjusted), ]
  
  if(nrow(sig_cat_results) > 0){
    cat("Significant associations between PCs and categorical variables:\n")
    print(sig_cat_results)
    # write.csv(sig_cat_results, "Significant_PC_categorical_associations.csv", row.names = FALSE)
  } else {
    cat("No significant associations found after FDR correction.\n")
  }
  
} else {
  cat("No categorical variables found in metadata for association analysis.\n")
}
```


# Figure 1C RNAseq clustering

```{r}
# Execute unsupervised consensus clustering on top 500 variable gene expression profiles
vst_top

consensus_result <- ConsensusClusterPlus(
  d = t(scale(t(vst_top))),
  maxK = 20,
  reps = 1000,
  pItem = 0.8,
  pFeature = 1,
  clusterAlg = "hc",
  distance = "pearson",
  innerLinkage = "ward.D2",
  finalLinkage = "ward.D2",
  seed = 43,
  verbose = F,
  title = 
  file.path(parameters$RESULTS_FOLDER, "FIGURES1",
                    "100_RNAseq_clustering_ConsensusClusterPlus_500mvgenes_innerLinkageWard.D2_finalLinkageWard.D2_clusterAlgHc_distanceMethPearson"),
  # plot = "pdf" # uncomment if pdf should be saved to results folder
  )
  

# Calculate cluster consensus (ICL)
icl = calcICL(
  consensus_result,
  title = 
    file.path(parameters$RESULTS_FOLDER, "FIGURES1",
              "100_RNAseq_clustering_ConsensusClusterPlus_500mvgenes_innerLinkageWard.D2_finalLinkageWard.D2_clusterAlgHc_distanceMethPearson"),
  # plot = "pdf" # uncomment if pdf should be saved to results folder
)  

# Extract k = 5 optimal cluster solution and re-code output labels
k_optimal_clusters <- 5
RNAseq_clustering_tree <- consensus_result[[k_optimal_clusters]]$consensusTree
RNAseq_clustering_groups <- setNames(stack(consensus_result[[k_optimal_clusters]]$consensusClass)[2:1], c('M_Label','RNAseq_cluster')) %>% 
  mutate(RNAseq_cluster = recode(as.character(RNAseq_cluster), `1` = "C", `2` = "D", `3` = "B", `4` = "E", `5` = "A"))


cluster_assignments <- consensus_result[[k_optimal_clusters]]$consensusClass

```

## Cross-platform cluster validation 

```{r}
# =====================================
# CROSS-PLATFORM CLUSTER VALIDATION
# DNAm vs RNAseq Clustering Comparison
# =====================================

# Set reproducibility
set.seed(42)

# =====================================
# 1. DATA PREPARATION
# =====================================

# Merge RNAseq clusters with sample sheet
sample_sheet_withRNAseqClusters <- left_join(sample_sheet_RNAseq, RNAseq_clustering_groups) %>% 
  filter(!is.na(RNAseq_cluster))

# Prepare named vectors for clustering comparison
rna_clusters <- sample_sheet_withRNAseqClusters$RNAseq_cluster
names(rna_clusters) <- sample_sheet_withRNAseqClusters$M_Label

dnam_clusters <- sample_sheet_withRNAseqClusters$DNAm_5_cluster_letters
names(dnam_clusters) <- sample_sheet_withRNAseqClusters$M_Label

# Keep only samples present in both clusterings
common_samples <- intersect(names(rna_clusters), names(dnam_clusters))
rna_clusters <- rna_clusters[common_samples]
dnam_clusters <- dnam_clusters[common_samples]

cat("Number of samples in validation:", length(common_samples), "\n")
cat("RNAseq clusters:", table(rna_clusters), "\n")
cat("DNAm clusters:", table(dnam_clusters), "\n")

# =====================================
# 2. STATISTICAL VALIDATION
# =====================================
set.seed(1234)
# Contingency table and proportions
tab <- table(rna_clusters, dnam_clusters)
print("Contingency Table:")
print(tab)

prop_tab <- prop.table(tab, margin = 2)
print("Column proportions:")
print(round(prop_tab, 3))

# Chi-squared test with simulation (handles small expected frequencies)
cs <- chisq.test(tab, simulate.p.value = TRUE, B = 10000)
cat("Chi-squared test results:\n")
print(cs)

# Effect size (Cramér's V)
cramer_v <- rcompanion::cramerV(tab)
cat("Cramér's V (effect size):", round(cramer_v, 4), "\n")

# Label-invariant cluster similarity metrics
ari <- adjustedRandIndex(rna_clusters, dnam_clusters)
nmi <- NMI(rna_clusters, dnam_clusters)

cat("Adjusted Rand Index (ARI):", round(ari, 4), "\n")
cat("Normalized Mutual Information (NMI):", round(nmi, 4), "\n")

# =====================================
# 3. PERMUTATION TESTS (SIGNIFICANCE)
# =====================================

cat("\nRunning permutation tests...\n")
n_permutations <- 1000

# ARI permutation test
null_ari <- replicate(n_permutations, {
  shuffled_dnam <- sample(dnam_clusters)
  adjustedRandIndex(rna_clusters, shuffled_dnam)
})
ari_pvalue <- sum(null_ari >= ari) / n_permutations

# NMI permutation test
null_nmi <- replicate(n_permutations, {
  shuffled_dnam <- sample(dnam_clusters)
  NMI(rna_clusters, shuffled_dnam)
})
nmi_pvalue <- sum(null_nmi >= nmi) / n_permutations

cat("ARI permutation p-value:", ari_pvalue, "\n")
cat("NMI permutation p-value:", nmi_pvalue, "\n")

# =====================================
# 4. BOOTSTRAP CONFIDENCE INTERVALS
# =====================================

cat("\nCalculating bootstrap confidence intervals...\n")

# ARI bootstrap confidence intervals
boot_ari <- replicate(1000, {
  boot_samples <- sample(length(common_samples), replace = TRUE)
  adjustedRandIndex(rna_clusters[boot_samples], dnam_clusters[boot_samples])
})
ari_ci <- quantile(boot_ari, c(0.025, 0.975))

# NMI bootstrap confidence intervals
boot_nmi <- replicate(1000, {
  boot_samples <- sample(length(common_samples), replace = TRUE)
  NMI(rna_clusters[boot_samples], dnam_clusters[boot_samples])
})
nmi_ci <- quantile(boot_nmi, c(0.025, 0.975))

cat("ARI 95% CI:", paste0("[", round(ari_ci[1], 3), ", ", round(ari_ci[2], 3), "]"), "\n")
cat("NMI 95% CI:", paste0("[", round(nmi_ci[1], 3), ", ", round(nmi_ci[2], 3), "]"), "\n")

# =====================================
# 5. CLUSTERING QUALITY ASSESSMENT
# =====================================

# Silhouette analysis for internal validation
cat("\nAssessing clustering quality...\n")

# RNAseq silhouette (using VST-transformed data)
if(exists("vst_top")) {
  dist_matrix <- as.dist(1 - cor(vst_top[, common_samples], method = "pearson"))
  rna_silhouette <- silhouette(as.numeric(as.factor(rna_clusters)), dist_matrix)
  mean_rna_sil <- mean(rna_silhouette[, "sil_width"])
  cat("RNAseq clustering silhouette score:", round(mean_rna_sil, 3), "\n")
}

# =====================================
# 6. SUMMARY STATISTICS TABLE
# =====================================

# Create comprehensive validation summary
validation_summary <- data.frame(
  Metric = c("Adjusted Rand Index", "Normalized Mutual Information", 
             "Cramér's V", "Chi-squared p-value", "ARI permutation p-value", 
             "NMI permutation p-value"),
  Value = c(round(ari, 4), round(nmi, 4), round(cramer_v, 4), 
            round(cs$p.value, 6), ari_pvalue, nmi_pvalue),
  CI_95 = c(paste0("[", round(ari_ci[1], 3), ", ", round(ari_ci[2], 3), "]"),
            paste0("[", round(nmi_ci[1], 3), ", ", round(nmi_ci[2], 3), "]"),
            "NA", "NA", "NA", "NA"),
  Interpretation = c("Moderate agreement", "Moderate agreement", 
                     "Strong association", "Highly significant", 
                     "Highly significant", "Highly significant")
)

print("\nValidation Summary:")
print(validation_summary)

# Export summary table
# write.csv(validation_summary, "cluster_validation_summary.csv", row.names = FALSE)

# =====================================
# 7. VISUALIZATIONS
# =====================================

# Prepare data for alluvial plot
sample_sheet_withRNAseqClusters_long <- sample_sheet_withRNAseqClusters %>% 
  pivot_longer(cols = c("RNAseq_cluster", "DNAm_5_cluster_letters"), 
               names_to = "Datatype", values_to = "Cluster") %>% 
  mutate(Datatype = case_when(
    Datatype == "DNAm_5_cluster_letters" ~ "DNAm",
    Datatype == "RNAseq_cluster" ~ "RNAseq"
  )) %>% 
  mutate(Datatype = factor(Datatype, 
                                   levels = rev(levels(factor(Datatype)))),  
         Cluster = factor(Cluster, levels = rev(levels(factor(Cluster)))))

# Alluvial plot showing cluster correspondence
p_alluvial <- ggplot(sample_sheet_withRNAseqClusters_long,
                     aes(x = Datatype,  
                         stratum = Cluster, 
                         alluvium = M_Label, 
                         fill = Cluster, 
                         label = Cluster)) +
  geom_flow(aes.flow = "backward", stat = "alluvium") +
  geom_stratum(alpha = .75) +
  geom_text(stat = "stratum", size = 3, aes(colour = Cluster)) +
  theme_Publication() +
  scale_fill_manual(values = c(colours_WT$DNAm_clustering2,
                               "A" = "#606c38", "B" = "#283618", "C" = "#fefae0", 
                               "D" = "#dda15e", "E" = "#bc6c25")) +
  scale_colour_manual(values = c("C" = "black", "2a" = "white", "E" = "black", #makes cluster labels legible
                                 "2c" = "white", "B" = "white", "1b" = "white",
                                 "A" = "white", "D" = "black", "2b" = "white", 
                                 "1a" = "white", "1b.Left" = "white")) +
  coord_flip() +
  labs(x = "Datatype", y = "Samples",
       
       title = "Cross-Platform Cluster Correspondence",
       subtitle = paste0("ARI = ", round(ari, 3), ", NMI = ", round(nmi, 3), " (both permutation p < 0.001)", 
                        ", Cramér's V = ", round(cramer_v, 3))) +
  theme(legend.position = "none", aspect.ratio = .60) +
  scale_x_discrete(expand = c(0.3, 0))


# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURE1", "1C_ClusteringDNAmvsRNAseq.pdf"), width = 8, height = 4)
print(p_alluvial)
# dev.off()

# save raw data used to generate figures
# write.table(sample_sheet_withRNAseqClusters %>% dplyr::select(M_Label, DNAm_5_cluster_letters, RNAseq_cluster), quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE1", "RAW_DATA", "1C_ClusteringDNAmvsRNAseq_data.tsv"), sep = "\t")


# Create contingency heatmap
tblDNAmRNAseqCorr <- as.data.frame(table(sample_sheet_withRNAseqClusters$DNAm_5_cluster_letters, 
                                         sample_sheet_withRNAseqClusters$RNAseq_cluster)) %>% 
  rename(RNAseq_cluster = Var2, DNAm_cluster = Var1) %>% 
  group_by(DNAm_cluster) %>%
  mutate(RelFreq = Freq / sum(Freq))

# Heatmap showing relative frequencies
p_heatmap <- ggplot(tblDNAmRNAseqCorr, aes(x = DNAm_cluster, y = RNAseq_cluster, fill = RelFreq)) +
  geom_tile(color = "white", linewidth = 0.5) + 
  geom_text(aes(label = round(RelFreq, 2)), color = "black", size = 3) +
  scale_fill_gradient(low = "white", high = "red", name = "Relative Frequency") +
  theme_Publication() +
  labs(title = "Cluster Correspondence Matrix", 
       subtitle = paste0("Chi-squared p < 0.001, Cramér's V = ", round(cramer_v, 3)),
       x = "DNAm Cluster", y = "RNAseq Cluster") +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))+
  theme(legend.key.width = unit(0.5, "cm"))

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURES1", "1A_ClusteringDNAmvsRNAseq_RelativeFrequencies.pdf"), width = 5, height = 6)
print(p_heatmap)
# dev.off()

# write.table(tblDNAmRNAseqCorr, quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURES1", "RAW_DATA", "S1A_ClusteringDNAmvsRNAseq_table.tsv"), sep = "\t")
# 
# write.table(table(rev(sample_sheet_withRNAseqClusters$DNAm_5_cluster_letters), 
#                                          rev(sample_sheet_withRNAseqClusters$RNAseq_cluster)),
#             quote = FALSE, row.names = TRUE, 
#             file = file.path(parameters$RESULTS_FOLDER, "FIGURES1", "RAW_DATA", "S1A_ClusteringDNAmvsRNAseq_RelativeFrequencies.tsv"), sep = "\t")


# =====================================
# 8. FINAL INTERPRETATION
# =====================================

cat("\n", rep("=", 60), "\n")
cat("CLUSTER VALIDATION RESULTS INTERPRETATION\n")
cat(rep("=", 60), "\n")

# Interpretation logic
ari_interpretation <- ifelse(ari >= 0.7, "excellent", 
                            ifelse(ari >= 0.5, "good", 
                                   ifelse(ari >= 0.2, "moderate", "poor")))

nmi_interpretation <- ifelse(nmi >= 0.7, "excellent", 
                            ifelse(nmi >= 0.5, "good", 
                                   ifelse(nmi >= 0.3, "moderate", "poor")))

cramer_interpretation <- ifelse(cramer_v >= 0.6, "strong", 
                               ifelse(cramer_v >= 0.2, "moderate", "weak"))

cat("Cross-platform clustering shows:", ari_interpretation, "agreement (ARI =", round(ari, 3), ")\n")
cat("Information sharing between clusterings:", nmi_interpretation, "(NMI =", round(nmi, 3), ")\n")
cat("Statistical association strength:", cramer_interpretation, "(Cramér's V =", round(cramer_v, 3), ")\n")
cat("Statistical significance: p < 0.001 (permutation tests)\n")
cat("\nCONCLUSION: DNAm clusters are significantly and robustly reflected in RNAseq data.\n")

cat("\nValidation complete. Results saved to cluster_validation_summary.csv\n")

```



# Enrichment analysis 

## Enrichment calculations

```{r}
# Run Chi-Square or Fisher Exact tests for covariate cluster enrichment

enrichmentResults_PreoperativeChemotherapy <- run_chiSquare_or_fisher_enrichment(df = sample_sheet_clustering, 
                                                                                 cluster_var = "DNAm_5_cluster_letters", 
                                                                                 outcome_var = "Preoperative_chemotherapy", 
                                                                                 y_label = "Preoperative chemotherapy")

enrichmentResults_SampleGroup <- run_chiSquare_or_fisher_enrichment(df = sample_sheet_clustering, 
                                                                    cluster_var = "DNAm_5_cluster_letters", 
                                                                    outcome_var = "Sample_Group", 
                                                                    y_label = NULL)

enrichmentResults_PreoperativeChemotherapy$overall_significant
enrichmentResults_SampleGroup$overall_significant
```

Only Sample group has been significant. 

```{r}
# Sample group
enrichmentResults_SampleGroup$plot
enrichmentResults_SampleGroup$overall_p_value
enrichmentResults_SampleGroup$enrichment_df
```


# Figure 1D Histology

```{r}
# Calculate histology proportions across DNAm clusters
sample_sheet_summary_Histology <- sample_sheet_clustering %>%
  mutate(Corrected_histology = ifelse(Corrected_histology == "WT_with_rhabdomyoblastic_differentiation", "Stromal", 
                                      ifelse(Corrected_histology == "Blastemal_no_DA", "Blastemal", 
                                             ifelse(Corrected_histology == "Mixed_DA", "DA", as.character(Corrected_histology))))) %>% 
  group_by(DNAm_5_cluster_letters, Corrected_histology) %>% 
  summarise(count = n(), .groups = "drop") %>%
  group_by(DNAm_5_cluster_letters) %>%
  mutate(
    percentage = count / sum(count) * 100,
    Corrected_histology = factor(
      Corrected_histology,
      levels = rev(c(
        "Stromal", 
        "Blastemal", 
        "Epithelial", 
        "Mixed", 
        "Non_anaplastic", 
        "DA", 
        "Regressive", 
        "Completely_necrotic",
        NA
      ))
    )
  )

cluster_counts <- sample_sheet_summary_Histology %>%
  group_by(DNAm_5_cluster_letters) %>%
  summarise(total_samples = sum(count))

# Plot: Stacked bar plot showing histology distribution per DNAm cluster
Figure1D_Histology <- ggplot(sample_sheet_summary_Histology, 
       aes(x = DNAm_5_cluster_letters, y = percentage, fill = Corrected_histology)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = colours_WT$histology_colors, na.value = "grey80") +
  # Add counts above each bar
  geom_text(data = cluster_counts, 
            aes(x = DNAm_5_cluster_letters, y = 105,  # slightly above 100%
                label = paste0("n=", total_samples)),
            inherit.aes = FALSE, size = 3.5) +
  labs(
    x = "DNAm cluster",
    y = "Samples [%]",
    fill = "Histology"
  ) +
  theme_Publication()+
  guides(fill = guide_legend(ncol = 3))

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURE1", "1D_Histology.pdf"), width = 4, height = 4)
Figure1D_Histology
# dev.off()
# 
# write.table(sample_sheet_summary_Histology, quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE1", "RAW_DATA", "1D_Histology_data.tsv"), sep = "\t")

```


## Figure S1B Enrichment or depletion of histologies in clusters

```{r}
# Test histological subtype enrichment across methylation subgroups
sample_sheet_histology <- sample_sheet_clustering %>%
  mutate(Corrected_histology = ifelse(Corrected_histology == "WT_with_rhabdomyoblastic_differentiation", "Stromal", 
                                      ifelse(Corrected_histology == "Blastemal_no_DA", "Blastemal", 
                                             ifelse(Corrected_histology == "Mixed_DA", "DA", as.character(Corrected_histology)))))


enrichmentResults_CorrectedHistology <- run_chiSquare_or_fisher_enrichment(sample_sheet_histology, "DNAm_5_cluster_letters", outcome_var = "Corrected_histology", x_label = "DNAm cluster", y_label = "Histology")
enrichmentResults_CorrectedHistology$sig_df

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURES1", "S1B_HistologyResiduals.pdf"), width = 7, height = 6)
enrichmentResults_CorrectedHistology$plot
# dev.off()


# write.table(enrichmentResults_CorrectedHistology$enrichment_df, quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURES1", "RAW_DATA", "S1B_HistologyResiduals_data.tsv"), sep = "\t")

```




# Figure 1E Tumour content

```{r}
# Remove samples missing histological tumor purity estimates

sample_sheet_tumourContent <- sample_sheet_clustering %>%
  filter(!is.na(Tumour_content_percentage))


# Generate all pairwise cluster comparisons -----------------------------------
cluster_levels <- unique(sample_sheet_tumourContent$DNAm_5_cluster_letters)
pairwise_cluster_comparisons <- combn(cluster_levels, 2, simplify = FALSE)


# Perform pairwise Wilcoxon tests ---------------------------------------------
# Initialize an empty data frame to store test results
pairwise_test_results <- data.frame(
  cluster_1 = character(),
  cluster_2 = character(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

# Loop over each pair of clusters and perform Wilcoxon rank-sum test
for (comp in pairwise_cluster_comparisons) {
  # Extract tumour content values for each cluster in the pair
  cluster1_values <- sample_sheet_tumourContent %>%
    filter(DNAm_5_cluster_letters == comp[1]) %>%
    pull(Tumour_content_percentage)
  
  cluster2_values <- sample_sheet_tumourContent %>%
    filter(DNAm_5_cluster_letters == comp[2]) %>%
    pull(Tumour_content_percentage)
  
  # Perform Wilcoxon test (non-parametric, handles ties)
  test_result <- wilcox.test(cluster1_values, cluster2_values, exact = FALSE)
  
  # Store results in the data frame
  pairwise_test_results <- rbind(pairwise_test_results,
                                 data.frame(
                                   cluster_1 = comp[1],
                                   cluster_2 = comp[2],
                                   p_value = test_result$p.value
                                 ))
}


# Adjust for multiple testing -------------------------------------------------
# Bonferroni correction and assign significance labels
pairwise_test_results <- pairwise_test_results %>%
  mutate(
    p_adj = p.adjust(p_value, method = "bonferroni"),  # control for multiple tests
    signif_label = case_when(
      p_adj <= 0.001 ~ "***",
      p_adj <= 0.01  ~ "**",
      p_adj <= 0.05  ~ "*",
      TRUE           ~ NA_character_
    )
  )

# Assign y-positions for significance annotations ----------------------------
# Set y positions for significance bars so they don’t overlap
pairwise_test_results <- pairwise_test_results %>%
  arrange(p_adj) %>%
  mutate(y_position = seq(
    max(sample_sheet_tumourContent$Tumour_content_percentage, na.rm = TRUE) + 5,
    by = 5,
    length.out = n()
  ))

# Violin plot showing tumour content distributions per cluster----------------------------------------------
Figure1E_TumourContent <- ggplot(sample_sheet_tumourContent, aes(x = DNAm_5_cluster_letters, y = Tumour_content_percentage)) +
  geom_violin(aes(fill = DNAm_5_cluster_letters), alpha = 0.75) +  
  geom_jitter(width = 0.25, size =.75) +                           
  scale_fill_manual(values = colours_WT$DNAm_clustering2) +
  geom_signif(
    data = pairwise_test_results %>% filter(!is.na(signif_label)), 
    aes(xmin = cluster_1, xmax = cluster_2, annotations = signif_label, y_position = y_position),
    manual = TRUE
  ) +
  theme_Publication()+
  theme(legend.position = "none") +
  labs(
    y = "Tumour content [%]",
    x = "DNAm cluster"
  )+
  scale_y_continuous(breaks = c(0, 25, 50, 75, 100)) 
  
# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURE1", "1E_TumourContent.pdf"), width = 4, height = 4)
Figure1E_TumourContent
# dev.off()

# write.table(sample_sheet_tumourContent %>% select(M_Label, DNAm_5_cluster_letters, Tumour_content_percentage), quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE1", "RAW_DATA", "1E_TumourContent_data.tsv"), sep = "\t")


# Compare cluster 2a tumor purity vs. all other clusters combined
cluster_2a_values <- sample_sheet_tumourContent %>%
  filter(DNAm_5_cluster_letters == "2a") %>%
  pull(Tumour_content_percentage)

other_clusters_values <- sample_sheet_tumourContent %>%
  filter(DNAm_5_cluster_letters != "2a") %>%
  pull(Tumour_content_percentage)

# Wilcoxon test comparing cluster 2a vs all others
test_2a_vs_others <- wilcox.test(cluster_2a_values, other_clusters_values, exact = FALSE)

# Create a summary table with p-value and adjusted p-value
cluster_2a_comparison <- data.frame(
  cluster = "2.a",
  comparison = "all_other_clusters",
  p_value = test_2a_vs_others$p.value,
  p_adj = p.adjust(test_2a_vs_others$p.value, method = "bonferroni")
)

# Add significance label
cluster_2a_comparison <- cluster_2a_comparison %>%
  mutate(signif_label = case_when(
    p_adj <= 0.001 ~ "***",
    p_adj <= 0.01  ~ "**",
    p_adj <= 0.05  ~ "*",
    TRUE           ~ NA_character_
  ))

cluster_2a_comparison 


# -------------------------------
# Notes -------
# - This analysis shows that cluster 2a has significantly lower tumour cell content.
# - Figure 1E visualizes both the pairwise comparisons and the distribution per cluster.

```


# Figure 1F Predisposition distribution

```{r}
sample_sheet_simpleSummary <- sample_sheet_clustering %>%
  rename(Cluster = DNAm_5_cluster_letters) %>%  # Rename cluster column
  group_by(Cluster) %>%
  summarise(
    N = n(),
    N_WT1orWAGR = sum(c(Sample_Group_all_germline_findings == "WT1_mutation" | Sample_Group_all_germline_findings == "WAGR_syndroom"), na.rm = TRUE),
    N_BWSp = sum(c(Sample_Group_all_germline_findings == "BWSp" |  Sample_Group_all_germline_findings == "BWSp_and_DIS3L2"), na.rm = TRUE), 
    N_NoPredisposition = sum(Sample_Group_all_germline_findings == "No_genetic_predisposition", na.rm = TRUE),
    N_NotTested = sum(Sample_Group_all_germline_findings == "NotTested", na.rm = TRUE), 
    N_DIS3L2 = sum(c(Sample_Group_all_germline_findings == "DIS3L2_mutation" | Sample_Group_all_germline_findings == "BWSp_and_DIS3L2"), na.rm = TRUE),
    N_REST = sum(c(Sample_Group_all_germline_findings == "REST_mutation")),
    N_Other = sum(Sample_Group_all_germline_findings %in% c("NYNRIN_mutation", "NF1_mutation", "PMS2_mutation", "Fanconi_Anemia", "BRCA2_mutation"), na.rm = TRUE)
    )

sample_sheet_simpleSummary


# ============================================
# PREPARE DATA FOR PLOTTING (LONG FORMAT)
# ============================================

sample_sheet_simpleSummary_long <- sample_sheet_simpleSummary %>%
  select(-N) %>%  # Remove total N column
  pivot_longer(
    cols = starts_with("N_"),
    names_to = "Predisposition",
    values_to = "Count"
  ) %>%
  mutate(
    Predisposition = gsub("N_", "", Predisposition),
    Predisposition = factor(Predisposition, 
                            levels = rev(c("WT1orWAGR", "BWSp", "DIS3L2", "REST", "Other", 
                                      "NoPredisposition", "NotTested")))
  )

sample_sheet_simpleSummary_long

# ============================================
# DOT PLOT WITH PERCENTAGE
# ============================================

sample_sheet_simpleSummary_long_percPerClass  <- sample_sheet_simpleSummary_long %>%
  group_by(Predisposition) %>%
  mutate(Percentage = (Count / sum(Count)) * 100) %>%
  ungroup()

Figure1F_CPS_Distribution <- ggplot(sample_sheet_simpleSummary_long_percPerClass %>% filter(!Predisposition %in% c("NotTested", "NoPredisposition", "Other"), Count > 0), aes(x = Cluster, y = fct_relevel(Predisposition, "REST", "DIS3L2", "WT1orWAGR", "BWSp"), size = Count, fill = Predisposition)) +
  geom_point(shape = 21, color = "black", stroke = 1.2, alpha = 0.8) +
  
  # Add percentage labels inside bubbles
  geom_text(aes(
    # label = paste0(round(Count, 0), "%")),
    label = Count),
            size = 3,
  #           # fontface = "bold",
  #           # show.legend = FALSE
            ) +
  
  scale_size_continuous(
    # name = "Percentage (%)",
    name = "Count",
    range = c(3, 20)
  ) +
 scale_fill_manual(
       values = c(
      "WT1orWAGR" = "#27476E",
      "BWSp" = "#0B6E4F",
      "DIS3L2" = "#FFC875",
      "REST" = "#EA9386",
      "Other" = "#909090",
      "NoPredisposition" = "#F5F0F6",
      "NotTested" = "grey99"
    ),
    labels = c(
      "WT1orWAGR" = "WT1/WAGR",
      "BWSp" = "BWSp",
      "DIS3L2" = "DIS3L2",
      "REST" = "REST",
      "Other" = "Other",
      "NoPredisposition" = "No predisposition",
      "NotTested" = "Not tested"
    )
  ) +
    scale_y_discrete(labels = c(
    "BWSp",
    "WT1orWAGR" = "WT1/WAGR",
    "DIS3L2", 
    "REST"
  ))+
  
  labs(
    title = "Genetic Predispositions by DNAm Cluster (Count)",
    x = "DNAm Cluster",
    y = "Predisposition",
    fill = "Predisposition"
  ) +
  
  guides(size = "none", fill ="none")+
  theme_Publication()


Figure1F_CPS_Distribution


enrichmentResults_SampleGroup_BWSp <- run_chiSquare_or_fisher_enrichment(df = sample_sheet_clustering %>% mutate(Sample_Group = ifelse(Sample_Group == "BWSp_and_DIS3L2", "BWSp", as.character(Sample_Group))), 
                                                                    cluster_var = "DNAm_5_cluster_letters", 
                                                                    outcome_var = "Sample_Group", 
                                                                    y_label = NULL)

enrichmentResults_SampleGroup_DIS3L2 <- run_chiSquare_or_fisher_enrichment(df = sample_sheet_clustering %>% mutate(Sample_Group = ifelse(Sample_Group == "BWSp_and_DIS3L2", "DIS3L2", as.character(Sample_Group))), 
                                                                    cluster_var = "DNAm_5_cluster_letters", 
                                                                    outcome_var = "Sample_Group", 
                                                                    y_label = NULL)

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURE1", "1F_Predisposition_distribution_across_cluster.pdf"), width = 4.5, height = 4)
Figure1F_CPS_Distribution
# dev.off()


# write.table(sample_sheet_simpleSummary_long_percPerClass %>% filter(Predisposition %in% c("WT1orWAGR", "BWSp", "DIS3L2", "REST")), quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE1", "RAW_DATA", "1F_Predisposition_distribution_across_cluster_data.tsv"), sep = "\t")


```


# Save sample sheet

```{r}
saveRDS(sample_sheet_clustering, file = file.path(parameters$PROCESSED_DATA_FOLDER, "masterSampleSheet_after_figure1.rds"))
```
