---
title: "Manuscript_WT_Figure2"
author: "Franziska Oberhammer"
date: "2026-09-10"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Environment & Setup

```{r libraries}

# Load required packages
suppressPackageStartupMessages({
  library(DESeq2)
  library(cowplot)
  library(ggExtra)
  library(Cairo)
  library(ggpubr)
  library(limma)
  library(ggplot2)
  library(ggnewscale)
  library(dplyr)
  library(ggsignif)  
  library(genefilter) 
  library(corrplot)
  library(dplyr)
  library(viridis)
  library(ggnewscale)
  library(EnhancedVolcano)
  library(plotly)
  library(tidyverse)
  library(circlize)
  library(ggalluvial)
  library(qs)
  library(ConsensusClusterPlus)
  library(gridExtra)
  library(RColorBrewer)
  library(ggsignif)
  library(dplyr)
  library(rstatix)
  library(officer)
  library(qs)
  library(rcompanion)
  library(mclust)
  library(scales)
  library(aricode)
})
```


```{r Configuration}
parameters <- list()

# Configuration ----------------------------------------------------------------
# Root and Working Directories
parameters$WILMS_FOLDER          <- "/hpc/pmc_kuiper/Wilms"
parameters$PROJECT_FOLDER        <- file.path(parameters$WILMS_FOLDER, "METHYLATION")  
parameters$CODE_FOLDER           <- file.path(parameters$PROJECT_FOLDER, "CODE")    
parameters$DATA_FOLDER           <- file.path(parameters$PROJECT_FOLDER, "DATA")       
parameters$ANALYSIS_FOLDER       <- file.path(parameters$PROJECT_FOLDER, "ANALYSIS") 
parameters$PROCESSED_DATA_FOLDER <- file.path(parameters$ANALYSIS_FOLDER, "PROCESSED_DATA") 
parameters$RESULTS_FOLDER        <- file.path(parameters$PROJECT_FOLDER, "RESULTS")

# Primary Input Data Files
parameters$CURRENT_SAMPLESHEET          <- file.path(parameters$PROCESSED_DATA_FOLDER, 
                                                     "masterSampleSheet_after_figure1.rds")
parameters$CURRENT_BETAS                <- file.path(parameters$PROCESSED_DATA_FOLDER,
                                                     "betas_V1V2_combined_probeFiltered_SampleQCFiltered_SNPsQC_postMortemAndVersionFiltered_SexMissmatchFiltered.qs")
parameters$CURRENT_BETAS_batchCorrected <- file.path(parameters$PROCESSED_DATA_FOLDER,
                                                     "betas_V1V2_combined_probeFiltered_SampleQCFiltered_SNPsQC_postMortemAndVersionFiltered_batchCorrected.qs")
parameters$CURRENT_CLUSTERING           <- file.path(parameters$ANALYSIS_FOLDER,
"EXPLORATORY_ANALYSIS/clustering/ConsensusClusterPlus_5000mvCpGs_manuscript/innerLinkageward.D2_finalLinkageward.D2_clusterAlghc_distanceMethpearson", "clustering_tree_5clusters.qs")


# Load Helper Functions, Themes, and Color Palettes
source(file.path(parameters$CODE_FOLDER, "00_theme_Publication.R"))
source(file.path(parameters$CODE_FOLDER, "00_WT_colour_palettes.R"))
source(file.path(parameters$CODE_FOLDER, "00_WT_helper_functions.R"))
```

# Data Loading & Preprocessing
Loading sample sheets, DNA methylation (beta matrices), clustering trees. Sample identifiers are harmonized across modalities.

```{r LoadData}
# Read primary metadata and methylation matrices
sample_sheets <- list()
sample_sheets$sample_sheet_meth   <- readRDS(parameters$CURRENT_SAMPLESHEET) %>% as.data.frame() 
betas                             <- qread(parameters$CURRENT_BETAS)
betas_batchCorrected              <- as.data.frame(qread(parameters$CURRENT_BETAS_batchCorrected))
clustering_tree                   <- qread(parameters$CURRENT_CLUSTERING)
```




```{r PrepData}
# Preprocess methylation beta matrices (keep only train cohort samples)
betas <- betas %>% 
  dplyr::select(-c("CI_V1", "BC", "EB")) 

# Standardize sample column names by removing version suffixes
colnames(betas) <- sub("_.*", "", colnames(betas))

# Change labels to M-labels
colnames(betas) <- sample_sheets$sample_sheet_meth$M_Label[match(colnames(betas), sample_sheets$sample_sheet_meth$Internal_Patient_ID)]

# check whether samples are in same order as in sample sheet
stopifnot(colnames(betas) == sample_sheets$sample_sheet_meth$M_Label)


# repeat for batch corrected ---------------
betas_batchCorrected <- betas_batchCorrected %>% 
  dplyr::select(-c("EB")) #CI_V1 and BC already removed in batch corrected data

# Standardize sample column names by removing version suffixes
colnames(betas_batchCorrected) <- sub("_.*", "", colnames(betas_batchCorrected))

# Change labels to M-labels
colnames(betas_batchCorrected) <- sample_sheets$sample_sheet_meth$M_Label[match(colnames(betas_batchCorrected), sample_sheets$sample_sheet_meth$Internal_Patient_ID)]

# check whether samples are in same order as in sample sheet
stopifnot(colnames(betas_batchCorrected) == sample_sheets$sample_sheet_meth$M_Label)
```




# Figure 2A Overrepresentation of BWSp in 2b
Classifying Beckwith-Wiedemann Spectrum (BWSp) status and assessing enrichment across DNA methylation (DNAm) sub-clusters.

```{r}

# Classify clinical and molecular BWSp status (including mosaicism)
sample_sheets$sample_sheet_withBWSpInfo <- sample_sheets$sample_sheet_meth %>%
  mutate(
    BWSp_status_final = case_when(
      Sample_Group_clinical_BWSp_OthersReduced == "BWSp" &
        !is.na(MethStatus_H19_IGF2_blood) &
        !is.na(MethStatus_H19_IGF2_kidney) &
        MethStatus_H19_IGF2_blood == "Normal" &
        MethStatus_H19_IGF2_kidney == "Hypermethylated" ~ "Mosaic",

      Sample_Group_clinical_BWSp_OthersReduced == "BWSp_and_DIS3L2" &
        !is.na(MethStatus_H19_IGF2_blood) &
        !is.na(MethStatus_H19_IGF2_kidney) &
        MethStatus_H19_IGF2_blood == "Normal" &
        MethStatus_H19_IGF2_kidney == "Hypermethylated" ~ "Mosaic",
      Sample_Group_clinical_BWSp_OthersReduced == "BWSp_and_DIS3L2" ~ "Yes",
      Sample_Group_clinical_BWSp_OthersReduced == "BWSp" ~ "Yes",
      Sample_Group_clinical_BWSp_OthersReduced == "BWSp_clinical_only" ~ "Clinical only",
      TRUE ~ "No"  # all other cases
    ),
    # Set order for stacked bars
    BWSp_status_final = factor(
      BWSp_status_final,
      levels = c("No", "Mosaic", "Clinical only", "Yes")
    ), 
    BWSp_updated  = BWSp_status_final %in% c("Mosaic", "Clinical only", "Yes"),
    BWSp_updatedYesNo = ifelse(BWSp_status_final %in% c("Mosaic", "Clinical only", "Yes"), "Yes", "No"),
    BWSp_updatedYesNoClinicalOnly = ifelse(BWSp_status_final %in% c("Mosaic", "Yes"), "Yes", ifelse(BWSp_status_final == "Clinical only", "Clinical only", "No"))

  )
```


### Statistical Test with for 2A (BWSp)

```{r}

# Statistical test: 
enrichmentResults_BWSp <- run_chiSquare_or_fisher_enrichment(sample_sheets$sample_sheet_withBWSpInfo, cluster_var = "DNAm_5_cluster_letters", outcome_var = "BWSp_updatedYesNo", x_label = "DNAm cluster")


table(sample_sheets$sample_sheet_withBWSpInfo$BWSp_updatedYesNo, sample_sheets$sample_sheet_withBWSpInfo %>% mutate(Cluster2bvsall = ifelse(DNAm_5_cluster_letters == "2b", "2b", "other")) %>% pull(Cluster2bvsall))

enrichmentResults_BWSp

```


```{r}
# Summarize BWSp proportions per cluster
bwsp_summary <- sample_sheets$sample_sheet_withBWSpInfo %>%
  group_by(DNAm_5_cluster_letters, BWSp_updatedYesNo) %>%
  summarise(sample_count = n(), .groups = "drop") %>%
  group_by(DNAm_5_cluster_letters) %>%
  mutate(
    percentage = sample_count / sum(sample_count) * 100
  )

cluster_counts <- bwsp_summary %>%
  group_by(DNAm_5_cluster_letters) %>%
  summarise(total_samples = sum(sample_count))

# Plot: Stacked bar chart of BWSp categories per DNAm cluster ---------------------
Figure2A_BWSp <- ggplot(bwsp_summary, 
       aes(x = DNAm_5_cluster_letters, y = percentage, fill = BWSp_updatedYesNo)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    x = "DNA methylation cluster",
    y = "Samples [%]",
    fill = "BWSp"
  ) +
  scale_fill_manual(values = colours_WT$bwsp_colors, breaks = c("Yes", "Clinical only", "Mosaic", "No")) +
  theme_Publication() +
  # Add a star annotation on top of cluster 2b to highlight significance. (Calculated above)
  geom_text(
    data = bwsp_summary[bwsp_summary$DNAm_5_cluster_letters == "2b", ],
    aes(label = "*", y = 110),
    size = 8, color = "black"
  ) +
  # Add counts above each bar
  geom_text(data = cluster_counts, 
            aes(x = DNAm_5_cluster_letters, y = 105,  # slightly above 100%
                label = paste0("n=", total_samples)),
            inherit.aes = FALSE, size = 3.5) +
  ylim(0, 110)+  # extend y-axis to make room for annotation
  scale_y_continuous(breaks = c(0, 25, 50, 75, 100))   # custom y-axis breaks


# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURE2BWSp", "2A_BWSp_distribution_across_cluster.pdf"), width = 8.5/2.54, height = 7/2.54)
Figure2A_BWSp
# dev.off()

# write.table(bwsp_summary, quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE2BWSp", "RAW_DATA", "2A_BWSp_distribution_across_cluster.tsv"), sep = "\t")


```



# Imprinting IC1/IC2 Analysis

Extracting beta values for genomic loci within Imprinting Control Regions 1 and 2 (ICR1 and ICR2 / IC1 and IC2) and performing consensus clustering.

Rather than including all four CTCF binding sites, clustering was restricted to two specific loci within ICR1 and ICR2. Selecting this targeted region yielded more robust, reproducible clustering across iterations, preventing the artificial merging of hypermethylated and pUPD samples into a single cluster. This genomic targeting aligns with previously established coordinates (Paper Link; ICR1: GRCh37/hg19 chr11:2019379–2020007 / GRCh38 chr11:1998149–1998777; ICR2: GRCh37/hg19 chr11:2720540–2721394 / GRCh38 chr11:2699310–2700164).

```{r}
# Load manifests
annoV1 <- read.delim(file.path(parameters$DATA_FOLDER, "BIN", "EPIC.hg38.manifest.tsv"))
annoV2 <- read.delim(file.path(parameters$DATA_FOLDER, "BIN", "EPICv2.hg38.manifest.tsv"))


# Extract CpG probes covering target IC1 & IC2 regions (GRCh38 / hg38 coordinates)
CpGs_IC1_EPICv1 <- annoV1 %>% 
  filter(CpG_chrm == "chr11", CpG_beg >= 1998149, CpG_end <=1998777) %>% pull(Probe_ID)
CpGs_IC1_EPICv2 <- annoV1 %>% 
  filter(CpG_chrm == "chr11", CpG_beg >= 1998149, CpG_end <=1998777) %>% pull(Probe_ID)
stopifnot(all(CpGs_IC1_EPICv1==CpGs_IC1_EPICv2))

CpGs_IC2_EPICv1 <- annoV1 %>% 
  filter(CpG_chrm == "chr11", CpG_beg >= 2699310, CpG_end <=2700164) %>% pull(Probe_ID)
CpGs_IC2_EPICv2 <- annoV1 %>% 
  filter(CpG_chrm == "chr11", CpG_beg >= 2699310, CpG_end <=2700164) %>% pull(Probe_ID)
stopifnot(all(CpGs_IC2_EPICv2 == CpGs_IC2_EPICv1))


# Subset batch-corrected methylation matrix for IC1/IC2 CpGs
CpGs_IC1 <- CpGs_IC1_EPICv1[CpGs_IC1_EPICv1 %in% rownames(betas_batchCorrected)]
CpGs_IC2 <- CpGs_IC2_EPICv1[CpGs_IC2_EPICv1 %in% rownames(betas_batchCorrected)]

IC1_meth <- betas_batchCorrected[CpGs_IC1,] %>% t()
IC2_meth <- betas_batchCorrected[CpGs_IC2,] %>% t()

colnames(IC1_meth) <- paste0(colnames(IC1_meth), "_IC1")
colnames(IC2_meth) <- paste0(colnames(IC2_meth), "_IC2")


IC1_IC2_combined <- cbind(IC1_meth, IC2_meth)
IC1_IC2_combined <- IC1_IC2_combined[match(sample_sheets$sample_sheet_withBWSpInfo$M_Label, rownames(IC1_IC2_combined)),] 

IC1_IC2_combined_scaled_transformed <- IC1_IC2_combined %>% scale() %>% t()

# Clean temporary manifest objects
rm(annoV1, annoV2, CpGs_IC1_EPICv1, CpGs_IC1_EPICv2, CpGs_IC2_EPICv1, CpGs_IC2_EPICv2, CpGs_IC1, CpGs_IC2, IC1_meth, IC2_meth)
```


## IC Consensus clustering

```{r}
# Perform Consensus Clustering over IC1/IC2 loci
innerLinkage <- "ward.D2"
finalLinkage <- "ward.D2"
clusterAlg <- "hc"
distanceMeth <- "euclidean"

results = ConsensusClusterPlus(
  IC1_IC2_combined_scaled_transformed,
  maxK=16,
  reps=1000,
  pItem=0.8,
  pFeature=1,
  innerLinkage = innerLinkage,
  finalLinkage = finalLinkage,
  clusterAlg = clusterAlg,
  distance = distanceMeth,
  title = file.path(parameters$ANALYSIS_FOLDER,
                    "EXPLORATORY_ANALYSIS/clustering/ConsensusClusterPlus_IC1_IC2_manuscript",
                    paste0("innerLinkage", innerLinkage,
                           "_finalLinkage", finalLinkage,
                           "_clusterAlg", clusterAlg,
                           "_distanceMeth", distanceMeth)),
  seed=1262118388.71279,
  # plot="pdf" # Uncomment to export consensus cluster decision plots
)

# calculate cluster consensus
icl = calcICL(
  results,
  title = file.path(parameters$ANALYSIS_FOLDER,
                    "EXPLORATORY_ANALYSIS/clustering/ConsensusClusterPlus_IC1_IC2_manuscript",
                    paste0("innerLinkage", innerLinkage,
                           "_finalLinkage", finalLinkage,
                           "_clusterAlg", clusterAlg,
                           "_distanceMeth", distanceMeth)),
  # plot = "pdf" # Uncomment to export plots
)



# 4 clusters were found to be most stable 
IC1IC2_consensus_clustering <- results[[4]]
IC1IC2_consensus_clustering_tree <- IC1IC2_consensus_clustering$consensusTree
IC1IC2_consensus_clustering_tree$labels <- rownames(IC1_IC2_combined)




# clean global environment
rm(innerLinkage, finalLinkage, clusterAlg, distanceMeth, icl, IC1_IC2_combined_scaled_transformed)
```

## Figure S2B Heatmap IC1/IC2

```{r}
# Visualize IC1 and IC2 Methylation Profiles

stopifnot(sample_sheets$sample_sheet_withBWSpInfo$M_Label == rownames(IC1_IC2_combined))

# pdf(file = file.path(parameters$RESULTS_FOLDER, "FIGURES2_BWSp", "IC1_IC2_methylation_BWSpInfo.pdf"), width = 8, height = 20)
ComplexHeatmap::Heatmap(IC1_IC2_combined, cluster_columns = FALSE,
                        col = circlize::colorRamp2(seq(0,1,0.01), colours_WT$meth),
                        cluster_rows= as.dendrogram(IC1IC2_consensus_clustering_tree),
                        row_split = 4,
                        name = "beta-value",
                        row_names_side = "left", 
                        row_dend_side = "right",
                        clustering_method_rows = "complete",
                        right_annotation = ComplexHeatmap::rowAnnotation(
                          BWSp             = factor(sample_sheets$sample_sheet_withBWSpInfo$BWSp_updatedYesNoClinicalOnly, levels = c("Yes", "Clinical only", "No")), 
                          col = list(BWSp  = c("Yes" = "#0a6f51", "Clinical only" = "#4cbb83", "No" = "#e0e0df"),
                                     Cluster = colours_WT$DNAm_clustering2
                                     ),
                          na_col = "white")
                      
)
# dev.off()

# write.table(IC1_IC2_combined %>% as.data.frame() %>%  rownames_to_column("M_Label") %>% left_join(sample_sheets$sample_sheet_withBWSpInfo %>% select(M_Label, BWSp_updatedYesNoClinicalOnly, DNAm_5_cluster_letters), by = "M_Label"), quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURES2_BWSp", "RAW_DATA", "IC1_IC2_methylation_data.tsv"), sep = "\t")

```

## Combine IC1/IC2 information with sample sheet

```{r}
# Add information whether sample is in hypermethylation, pUPD, normal or intermediate cluster. --------------------------------------------------------
IC1_IC2_clustering <- IC1IC2_consensus_clustering$consensusClass %>% as.data.frame() %>% rownames_to_column()
colnames(IC1_IC2_clustering) <- c("M_Label", "IC1_IC2_clustering")

# Map reference samples to functional categories
IC1_IC2_clustering <- IC1_IC2_clustering %>%
  mutate(
    imprinting = case_when(
      M_Label == "M904AAB" ~ "pUPD",
      M_Label == "M341AAA" ~ "hyper",
      M_Label == "M926AAD" ~ "normal",
      M_Label == "M070AAH" ~ "intermediate",
      TRUE ~ NA_character_
    )
  )

# Propagate annotations to remaining cluster members
IC1_IC2_clustering_imprintingInfo <- IC1_IC2_clustering %>%
  group_by(IC1_IC2_clustering) %>%
  mutate(
    imprinting = dplyr::first(na.omit(imprinting))
  ) %>%
  ungroup() %>%
  mutate(
    IC1_IC2_clustering = factor(IC1_IC2_clustering),
    imprinting = factor(imprinting)
  )

# write.table(IC1_IC2_clustering_imprintingInfo %>% dplyr::select(M_Label, imprinting), quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURES2_BWSp", "RAW_DATA", "IC1_IC2_clustering_imprintingInfo_data.tsv"), sep = "\t")

# Join with sample sheet metadata
sample_sheets$sample_sheet_withBWSpAndImprintingInfo <- left_join(sample_sheets$sample_sheet_withBWSpInfo,  IC1_IC2_clustering_imprintingInfo %>% select(M_Label, IC1_IC2_clustering, imprinting))

# clean global environment
rm(IC1_IC2_clustering, IC1IC2_consensus_clustering, IC1IC2_consensus_clustering_tree, results, IC1_IC2_combined)
```

```{r}
table(sample_sheets$sample_sheet_withBWSpAndImprintingInfo %>% filter(BWSp_updatedYesNo == "No") %>% pull(imprinting))

table(sample_sheets$sample_sheet_withBWSpAndImprintingInfo %>% pull(imprinting))
```



# Figure 2B Imprinting Distribution Across DNAm Clusters

```{r}
# Summarize imprinting proportions across DNAm clusters
sample_sheet_summary_Imprinting <- sample_sheets$sample_sheet_withBWSpAndImprintingInfo %>%
  group_by(DNAm_5_cluster_letters, imprinting) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(DNAm_5_cluster_letters) %>%
  mutate(percentage = count / sum(count) * 100) %>% 
  mutate(imprinting = factor(imprinting, levels = c( "normal", "intermediate",  "pUPD","hyper")))

# Calculate total samples per cluster
cluster_counts <- sample_sheet_summary_Imprinting %>%
  group_by(DNAm_5_cluster_letters) %>%
  summarise(total_samples = sum(count))

# Plot: Stacked bar plot showing imprinting distribution per DNAm cluster
Figure2B_Imprinting <- ggplot(sample_sheet_summary_Imprinting, 
       aes(x = DNAm_5_cluster_letters, y = percentage, fill = imprinting)) +
  geom_bar(stat = "identity", position = "stack", alpha = 1) +
  scale_fill_manual(values = colours_WT$IC1_IC2, na.value = "grey80") +
  # Add counts above each bar
  geom_text(data = cluster_counts, 
            aes(x = DNAm_5_cluster_letters, y = 105,  # slightly above 100%
                label = paste0("n=", total_samples)),
            inherit.aes = FALSE, size = 3.5) +
  labs(
    x = "DNAm cluster",
    y = "Samples [%]",
    fill = "Imprinting 11p15"
  ) +
  theme_Publication()

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURE2BWSp", "2B_Imprinting_distribution_across_clusters.pdf"), width = 8.5/2.54, height = 6/2.54)
Figure2B_Imprinting
# dev.off()


# write.table(sample_sheet_summary_Imprinting, quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE2BWSp", "RAW_DATA", "2B_Imprinting_distribution_across_clusters.tsv"), sep = "\t")

```

## Figure S2C Imprinting Enrichment test

```{r}
# Run chi-square / Fisher enrichment for 11p15 imprinting across DNAm clusters
enrichmentResults_Imprinting <- run_chiSquare_or_fisher_enrichment(sample_sheets$sample_sheet_withBWSpAndImprintingInfo %>% mutate(imprinting = factor(imprinting, levels = rev(c("normal", "intermediate", "hyper", "pUPD")))), cluster_var = "DNAm_5_cluster_letters", outcome_var = "imprinting", x_label = "DNAm cluster", y_label = "Imprinting")


# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURES2_BWSp", "S2C_ImprintingResiduals.pdf"), width = 8.5/2.54, height = 7/2.54)
enrichmentResults_Imprinting$plot
# dev.off()

# write.table(enrichmentResults_Imprinting$enrichment_df, quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURES2_BWSp", "RAW_DATA", "S2C_ImprintingEnrichmentResults.tsv"), sep = "\t")

table(sample_sheets$sample_sheet_withBWSpAndImprintingInfo$imprinting, sample_sheets$sample_sheet_withBWSpAndImprintingInfo$DNAm_5_cluster_letters)

enrichmentResults_Imprinting

```


## Tumour content across imprinting groups

Testing whether estimated tumor purity varies significantly across 11p15 imprinting groups.

```{r}
# Clean histology labels and filter missing tumor purity values
sample_sheets$sample_sheet_withBWSpAndImprintingInfo_imprinting_plot <- sample_sheets$sample_sheet_withBWSpAndImprintingInfo %>%
  mutate(
    # Combine histology columns
    Hist = case_when(
      !is.na(Corrected_histology) ~ Corrected_histology,
      !is.na(Histology) ~ Histology
    ),
    # Simplify specific histology labels
    Hist = case_when(
      Hist == "WT_with_rhabdomyoblastic_differentiation" ~ "Stromal", 
      Hist == "Blastemal_no_DA" ~ "Blastemal",
      Hist == "Mixed_DA" ~ "DA",
      TRUE ~ Hist
    ),
     # Change imprinting label
    imprinting = case_when(
      imprinting == "hyper" ~ "hypermethylated", # change name
      TRUE ~ imprinting
    ), 
    # Set order for imprinting factor
    imprinting = factor(as.character(imprinting), levels = c("normal", "intermediate", "hypermethylated", "pUPD"))
  ) 

table(sample_sheets$sample_sheet_withBWSpAndImprintingInfo_imprinting_plot$Hist)

# Prepare and clean data ------------------------------------------------------
# Remove rows with missing tumour content
sample_sheets$sample_sheet_withBWSpAndImprintingInfo_imprintingTumourcontent_plot <- sample_sheets$sample_sheet_withBWSpAndImprintingInfo_imprinting_plot %>% 
  filter(!is.na(Tumour_content_percentage))


# Generate all pairwise imprinting comparisons -----------------------------------
# Create a list of all unique impringing labels
imprinting_levels <- levels(sample_sheets$sample_sheet_withBWSpAndImprintingInfo_imprintingTumourcontent_plot$imprinting)

# Generate all combinations of imprinting pairs for comparison
pairwise_imprinting_comparisons <- combn(imprinting_levels, 2, simplify = FALSE)


# Perform pairwise Wilcoxon tests ---------------------------------------------
# Initialize an empty data frame to store test results
pairwise_test_results_imprinting <- data.frame(
  imprinting_1 = character(),
  imprinting_2 = character(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

# Loop over each pair of imprinting and perform Wilcoxon rank-sum test
for (comp in pairwise_imprinting_comparisons) {
  # Extract tumour content values for each imprinting in the pair
  imprinting1_values <- sample_sheets$sample_sheet_withBWSpAndImprintingInfo_imprintingTumourcontent_plot %>%
    filter(imprinting  == comp[1]) %>%
    pull(Tumour_content_percentage)
  
  imprinting2_values <- sample_sheets$sample_sheet_withBWSpAndImprintingInfo_imprintingTumourcontent_plot %>%
    filter(imprinting == comp[2]) %>%
    pull(Tumour_content_percentage)
  
  # Perform Wilcoxon test (non-parametric, handles ties)
  test_result <- wilcox.test(imprinting1_values, imprinting2_values, exact = FALSE)
  
  # Store results in the data frame
  pairwise_test_results_imprinting <- rbind(pairwise_test_results_imprinting,
                                            data.frame(
                                              imprinting_1 = comp[1],
                                              imprinting_2 = comp[2],
                                              p_value = test_result$p.value
                                            ))
}


# Adjust for multiple testing -------------------------------------------------
# Bonferroni correction and assign significance labels
pairwise_test_results_imprinting <- pairwise_test_results_imprinting %>%
  mutate(
    p_adj = p.adjust(p_value, method = "bonferroni"),  # control for multiple tests
    signif_label = case_when(
      p_adj <= 0.001 ~ "***",
      p_adj <= 0.01  ~ "**",
      p_adj <= 0.05  ~ "*",
      TRUE           ~ NA_character_
    )
  )

# Assign y-positions for significance annotations ----------------------------
# Set y positions for significance bars so they don’t overlap
pairwise_test_results_imprinting <- pairwise_test_results_imprinting %>%
  arrange(p_adj) %>%
  mutate(y_position = seq(
    max(sample_sheets$sample_sheet_withBWSpAndImprintingInfo_imprintingTumourcontent_plot$Tumour_content_percentage, na.rm = TRUE) + 5,
    by = 7.5,
    length.out = n()
  ))


Figure_imprintingTumourContent <- sample_sheets$sample_sheet_withBWSpAndImprintingInfo_imprintingTumourcontent_plot %>% 
  ggplot(aes(x = imprinting, y = Tumour_content_percentage)) +

  # ----- Boxplot/violin layer -----
  geom_violin(
  # geom_boxplot(
    aes(fill = imprinting),
    alpha = 0.7,
    show.legend = FALSE   # hide from legend
  ) +
  scale_fill_manual(
    name = "IC1/IC2 imprinting status",
    values = c(colours_WT$IC1_IC2, "hypermethylated" = "#A31621")) +
  guides(fill = guide_legend(ncol = 2)) +

  # ----- Significance annotations -----
  geom_signif(
    data = pairwise_test_results_imprinting %>% filter(!is.na(signif_label)),
    aes(
      xmin = imprinting_2,
      xmax = imprinting_1,
      annotations = signif_label,
      y_position = y_position
    ),
    manual = TRUE
  ) +

  # ----- Points layer with separate fill scale -----
  new_scale_fill() +
  geom_jitter(
    aes(fill = Hist),
    # colour = "black",
    shape = 21,
    width = 0.2,
    height = 0.2,
    size = 1
  ) +
  scale_fill_manual(
    name = "Histology",
    values = colours_WT$histology,
    labels = c(
      "Completely_necrotic" = "Completely necrotic",
      "Non_anaplastic" = "Non anaplastic"
    )
  ) +

  # ----- Labels and axes -----
  labs(
    x = "IC1/IC2 imprinting status",
    y = "Tumour content [%]",
    title = ""
  ) +
  scale_x_discrete() +
  ylim(0, 120) +

  # ----- Theme and legend tweaks -----
  theme_Publication() +
  theme(
    legend.box = "vertical",        # stack legends
    legend.box.just = "left",       # align left
    legend.box.spacing = unit(0.5, "cm") # space between legends
  ) +
  guides(fill = guide_legend(ncol = 3)) 

Figure_imprintingTumourContent

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURES2_BWSp", "S2D_TumourContentAcrossImprinting.pdf"), width = 7/2.54, height = 8/2.54)
Figure_imprintingTumourContent
# dev.off()

# write.table(sample_sheets$sample_sheet_withBWSpAndImprintingInfo_imprintingTumourcontent_plot %>% select(M_Label, Hist, imprinting, Tumour_content_percentage), quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURES2_BWSp", "RAW_DATA", "S2D_TumourContentAcrossImprinting_data.tsv"), sep = "\t")


```




# Stratified Imprinting Analyses (BWSp vs. Non-BWSp)

Evaluating whether cluster preferences differ between BWSp and non-BWSp cases within specific somatic imprinting subgroups (IC1 Hypermethylation vs. pUPD).


```{r}
# Ensure cluster levels
sample_sheets$sample_sheet_withBWSpAndImprintingInfo_Factors <- sample_sheets$sample_sheet_withBWSpAndImprintingInfo %>%
  mutate(
    DNAm_5_cluster_letters = factor(DNAm_5_cluster_letters, levels = c("1a","1b","2a","2b","2c")),
    BWSp_updatedYesNo = factor(BWSp_updatedYesNo, levels = c("Yes","No")))

# Prepare data
sample_sheets$sample_sheet_withBWSpAndImprintingInfo_Factors_plot <- sample_sheets$sample_sheet_withBWSpAndImprintingInfo_Factors %>%
  mutate(DNAm_5_cluster_letters = factor(DNAm_5_cluster_letters,
                                         levels = c("1a","1b","2a","2b","2c"))) %>%
  group_by(BWSp_updatedYesNo, imprinting, DNAm_5_cluster_letters) %>%
  filter(imprinting %in% c("pUPD", "hyper")) %>% 
  summarise(n = n(), .groups = "drop") %>%
  group_by(BWSp_updatedYesNo, imprinting) %>%
  complete(DNAm_5_cluster_letters, fill = list(n = 0)) %>%  # ensure empty clusters appear
  mutate(prop = n / sum(n)) %>%
  ungroup()


# Compute total samples per imprinting × BWSp group
group_countss <- sample_sheets$sample_sheet_withBWSpAndImprintingInfo_Factors %>%
  filter(imprinting %in% c("pUPD", "hyper")) %>% 
  group_by(imprinting, BWSp_updatedYesNo) %>%
  summarise(total_n = n(), .groups = "drop")


# Function that returns a list with plot and contingency table
create_plot_and_table <- function(imp) {
  # Set colors per imprinting
  colscale <- if (imp == "pUPD") c("Yes" = "blue", "No" = "lightblue") else c("Yes" = "red", "No" = "pink")
  
  # Filter data for current imprinting
  df_sub <- sample_sheets$sample_sheet_withBWSpAndImprintingInfo_Factors_plot %>% filter(imprinting == imp)
  
  print(df_sub)
  
  dff_sub <- sample_sheets$sample_sheet_withBWSpAndImprintingInfo_Factors %>% 
    filter(imprinting == imp)
  
  ct <- table(dff_sub$DNAm_5_cluster_letters, dff_sub$BWSp_updatedYesNo)
  
  # Contingency table
  print(ct)
  # Subtitle with sample counts
  subtitle_text <- group_countss %>%
    filter(imprinting == imp) %>%
    mutate(text = paste(BWSp_updatedYesNo, "n =", total_n)) %>%
    pull(text) %>%
    paste(collapse = " | ")
  
  # Create plot
  p <- ggplot(df_sub %>% mutate(BWSp_updatedYesNo = factor(BWSp_updatedYesNo, levels = c( "Yes", "No"))), aes(x = DNAm_5_cluster_letters, y = prop, fill = factor(BWSp_updatedYesNo, levels = c( "Yes", "No")))) +
    geom_bar(stat = "identity", position = position_dodge()) +
    scale_y_continuous(labels = scales::percent_format(), limits = c(0,1)) +
    scale_fill_manual(values = colscale) +
    labs(
      title = paste(imp, "IC1/IC2 status"),
      subtitle = subtitle_text,
      x = "Cluster",
      y = "Proportion within group",
      fill = "BWSp"
    ) +
    theme_Publication() +
    theme(
      plot.title = element_text(face = "bold", hjust = 0),
      plot.subtitle = element_text(hjust = 0)
    ) +
    facet_wrap(~BWSp_updatedYesNo)
  # Return both as a list
  return(list(plot = p, contingencyTable = ct))
}

# Use lapply to iterate over imprinting types
resulstsis <- lapply(c("hyper", "pUPD"), create_plot_and_table)
names(resulstsis) <- c("hyper", "pUPD")  

# Access results
resulstsis$hyper$plot
resulstsis$hyper$contingencyTable
resulstsis$pUPD$plot
resulstsis$pUPD$contingencyTable


fisher.test(resulstsis$hyper$contingencyTable)
fisher.test(resulstsis$pUPD$contingencyTable)



# Combine plots side by side
combined <- resulstsis$hyper$plot | resulstsis$pUPD$plot +
  patchwork::plot_annotation(title = "Cluster distribution: BWSp vs Non-BWSp")


# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURE2BWSp", "2CD_ProportionImprintingAcrossClustersStratifiedPerBWSpStatus_facets.pdf"), width = 15/2.54, height = 8/2.54)
combined
# dev.off()

# write.table(sample_sheets$sample_sheet_withBWSpAndImprintingInfo_Factors_plot %>% filter(imprinting == "hyper"), quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE2BWSp", "RAW_DATA", "2C_ProportionImprintingAcrossClustersStratifiedPerBWSpStatus_hyper_data.tsv"), sep = "\t")
# write.table(sample_sheets$sample_sheet_withBWSpAndImprintingInfo_Factors_plot %>% filter(imprinting == "pUPD"), quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE2BWSp", "RAW_DATA", "2D_ProportionImprintingAcrossClustersStratifiedPerBWSpStatus_pUPD_data.tsv"), sep = "\t")



```

BWSp samples, whether hypermethylated or carrying pUPD, tended to be enriched in cluster 2b, with 73% of hypermethylated BWSp and 50% of pUPD BWSp samples located there. In contrast, non-BWSp samples were more evenly distributed across clusters. Despite this apparent enrichment, Fisher’s exact test showed that the cluster distributions of BWSp samples were not significantly different from non-BWSp samples (hypermethylated: p = 0.25; pUPD: p = 0.25), likely reflecting the limited sample size or the subtle nature of these differences.

Next, we want to show how the proportion of BWSp vs non-BWSp samples varies across the DNAm clusters, separately for the two somatic imprinting patterns (IC1 hypermethylation and pUPD/LOH). In other words, we’re assessing whether BWSp status is enriched in specific DNAm clusters within each imprinting category.

```{r}
# Calculate 100% stacked proportions across clusters
sample_sheets$sample_sheet_withBWSpAndImprintingInfo_Factors_plot_100perc <- sample_sheets$sample_sheet_withBWSpAndImprintingInfo_Factors_plot %>%
  filter(DNAm_5_cluster_letters != "2a") %>% 
  select(BWSp_updatedYesNo, imprinting, DNAm_5_cluster_letters, n) %>%
  mutate(
    BWSpandSomaticImprinting = case_when(
      imprinting == "hyper" & BWSp_updatedYesNo == "No"  ~ "non-BWSp & IC1 hypermethylated",
      imprinting == "hyper" & BWSp_updatedYesNo == "Yes" ~ "BWSp & IC1 hypermethylated",
      imprinting == "pUPD"  & BWSp_updatedYesNo == "No"  ~ "non-BWSp & pUPD/LOH",
      imprinting == "pUPD"  & BWSp_updatedYesNo == "Yes" ~ "BWSp & pUPD/LOH"
    ),
    BWSpandSomaticImprinting = factor(
      BWSpandSomaticImprinting,
      levels = c(
        "non-BWSp & IC1 hypermethylated",
        "BWSp & IC1 hypermethylated",
        "non-BWSp & pUPD/LOH",
        "BWSp & pUPD/LOH"
      )
    )
  )

FigureS2E_BWSpImprinting <- ggplot(
  sample_sheets$sample_sheet_withBWSpAndImprintingInfo_Factors_plot_100perc,
  aes(
    x = DNAm_5_cluster_letters,
    y = n,
    fill = BWSpandSomaticImprinting
  )
) +
  geom_col(position = "fill") +
  facet_wrap(~ imprinting) +
  scale_fill_manual(
    values = c(
      "non-BWSp & IC1 hypermethylated" = "#f8beca",
      "BWSp & IC1 hypermethylated"     = "#ec1e24",
      "non-BWSp & pUPD/LOH"            = "#add7e5",
      "BWSp & pUPD/LOH"                = "#3853a3"
    )
  ) +
  scale_y_continuous(
    labels = percent_format()
  ) +
  labs(
    x = "DNAm cluster",
    y = "Percentage [%]",
    fill = "BWSp & somatic imprinting status"
  ) +
  theme_classic()

FigureS2E_BWSpImprinting


# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURES2_BWSp", "S2E_ProportionBWSpPerClusterStratifiedPerimprinting.pdf"), width = 15/2.54, height = 8/2.54)
FigureS2E_BWSpImprinting
# dev.off()

# write.table(sample_sheets$sample_sheet_withBWSpAndImprintingInfo_Factors_plot_100perc, quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURES2_BWSp", "RAW_DATA", "S2E_ProportionBWSpPerClusterStratifiedPerimprinting.pdf_data.tsv"), sep = "\t")
```


# Save sample sheet
```{r}
saveRDS(sample_sheets$sample_sheet_withBWSpAndImprintingInfo, file = file.path(parameters$PROCESSED_DATA_FOLDER, "masterSampleSheet_after_figure2.rds"))
```
