---
title: "Manuscript_WT_Figure3"
author: "Franziska Oberhammer"
date: "2026-09-10"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
```


# Setup & Environment
Loading required visualization, data manipulation, and genomic utilities alongside global configuration paths.

```{r}
# Load required libraries
suppressPackageStartupMessages({
  library(circlize)          
  library(ComplexHeatmap)    
  library(grid)              
  library(dplyr)            
  library(tidyverse)        
  library(RColorBrewer)
  library(qs)               
  library(ggsignif)
})
```


```{r}

parameters <- list()
parameters$PROJECT_FOLDER         <- "/hpc/pmc_kuiper/Wilms/METHYLATION"  
parameters$ANALYSIS_FOLDER        <- file.path(parameters$PROJECT_FOLDER, "ANALYSIS")
parameters$PROCESSED_DATA_FOLDER  <- file.path(parameters$ANALYSIS_FOLDER, "PROCESSED_DATA")
parameters$CODE_FOLDER            <- file.path(parameters$PROJECT_FOLDER, "CODE")
parameters$RESULTS                <- file.path(parameters$PROJECT_FOLDER, "RESULTS")

# File paths
parameters$CURRENT_CLUSTERING     <- file.path(parameters$ANALYSIS_FOLDER,
                                               "EXPLORATORY_ANALYSIS/clustering/ConsensusClusterPlus_5000mvCpGs_manuscript/innerLinkageward.D2_finalLinkageward.D2_clusterAlghc_distanceMethpearson", "clustering_tree_5clusters.qs")

parameters$CURRENT_SAMPLESHEET    <- file.path(parameters$PROCESSED_DATA_FOLDER, "masterSampleSheet_after_figure2.rds")


# Load helper functions
source(file.path(parameters$CODE_FOLDER, "00_theme_Publication.R"))
source(file.path(parameters$CODE_FOLDER, "00_WT_colour_palettes.R"))
source(file.path(parameters$CODE_FOLDER, "00_WT_helper_functions.R"))

```

# Data Loading & Preprocessing
Loading master sample sheet, consensus clustering tree, and variant metadata for WT1 and CTNNB1. Standardizing internal sample identifiers.

```{r}

# Load sample sheet
sample_sheet <- readRDS(parameters$CURRENT_SAMPLESHEET) %>% 
  mutate(
    Cluster1avsall = ifelse(DNAm_5_cluster_letters == "1a", "Yes", "No"), 
    Cluster1bvsall = ifelse(DNAm_5_cluster_letters == "1b", "Yes", "No"), 
    Cluster2avsall = ifelse(DNAm_5_cluster_letters == "2a", "Yes", "No"), 
    Cluster2bvsall = ifelse(DNAm_5_cluster_letters == "2b", "Yes", "No"), 
    Cluster2cvsall = ifelse(DNAm_5_cluster_letters == "2c", "Yes", "No")
  ) 

dim(sample_sheet)

# Load Consensus Clustering Tree
clustering_tree <- qread(parameters$CURRENT_CLUSTERING)

# Load WT1 and CTNNB1 Variant Annotations
sample_data <- readxl::read_xlsx(
  "/hpc/pmc_kuiper/Wilms/METHYLATION/ANALYSIS/WT1/20260716_wt1overview.xlsx" # this corresponds to supplementary table S2
) %>% 
  dplyr::select(
    Internal_Patient_ID, 
    wt1_germline_testing, wt1_germline_researchAndDiagnosticsCombined, wt1_germline_Diagnostics,  wt1_germline_oncoplot_label,   
    wt1_somatic_1, wt1_somatic_Diagnostics_1, wt1_somatic_1_oncoplot_label,
    wt1_somatic_2, wt1_somatic_Diagnostics_2, wt1_somatic_2_oncoplot_label,
    wt1_loh, 
    ctnnb1_somatic_1, ctnnb1_somatic_2, ctnnb1_somatic_1_oncoplot_label, ctnnb1_somatic_2_oncoplot_label,
    ctnnb1_Diagnostics_somatic_1, ctnnb1_Diagnostics_somatic_2, DataType,
    wt1_somatic_1_Classification, wt1_somatic_2_Classification
  ) %>%
  mutate(sample_id = Internal_Patient_ID) %>% 
  mutate()


# Map internal IDs to M-Labels and filter for train samples
sample_data$M_Label <- sample_sheet$M_Label[match(x = sample_data$Internal_Patient_ID, sample_sheet$Internal_Patient_ID)]

sample_data_filtered <- sample_data %>%   
  filter(M_Label %in% clustering_tree$labels)

# Extract unique variants in the dataset
values_inFile <- unique(c(
  sample_data_filtered$wt1_germline_testing, 
  sample_data_filtered$wt1_germline_oncoplot_label,
  sample_data_filtered$wt1_somatic_1, 
  sample_data_filtered$wt1_somatic_1_oncoplot_label,
  sample_data_filtered$wt1_somatic_2, 
  sample_data_filtered$wt1_somatic_2_oncoplot_label,
  sample_data_filtered$ctnnb1_somatic_1, 
  sample_data_filtered$ctnnb1_somatic_1_oncoplot_label,
  sample_data_filtered$ctnnb1_somatic_2, 
  sample_data_filtered$ctnnb1_somatic_2_oncoplot_label,
  sample_data_filtered$wt1_loh
))

values_inFileOncoplotLabels <- unique(c(
  sample_data_filtered$wt1_germline_oncoplot_label,
  sample_data_filtered$wt1_somatic_1_oncoplot_label,
  sample_data_filtered$wt1_somatic_2_oncoplot_label,
  sample_data_filtered$ctnnb1_somatic_1_oncoplot_label,
  sample_data_filtered$ctnnb1_somatic_2_oncoplot_label,
  sample_data_filtered$wt1_loh
))

```

# Variant Color Schemes & Functional Helpers

```{r}
# Read Variant Class Color Mapping
variant_info <- read.csv2(
  sep = "\t", 
  file = file.path(parameters$CODE_FOLDER, "00_WT_variant_info.tsv")) %>% 
  filter(NewOld == "New" | group %in% c("Copy number/LOH", "Datatype"))


# Function: Fetch Color Code per Variant Type
get_variant_color <- function(variant) {
  if (is.na(variant) || variant == "") {
    return(variant_info$color[variant_info$variant == "None_WT1"])
  }
  
  color_match <- variant_info$color[variant_info$variant == variant]
  if (length(color_match) > 0) return(color_match)
  
  return(variant_info$color[variant_info$variant == "None_WT1"])
}

# Verify that all variants in the data are defined in the color scheme
values_inFile[!values_inFile %in% variant_info$variant]
values_inFileOncoplotLabels[!values_inFileOncoplotLabels %in% variant_info$variant]

# Set color vector for heatmap
variant_colors <- setNames(variant_info$color, variant_info$variant)





# Function: Flag Non-Standard Biomaterial Sources
is_other_biomaterial <- function(variant) {
  if (is.na(variant) || variant == "") return(FALSE)
  
  flag <- variant_info$OtherBiomaterial[variant_info$variant == variant]
  if (length(flag) == 0) return(FALSE)
  
  return(flag %in% c("•", TRUE, "TRUE", 1))
}

# Function: Identify Pathogenic Hits
is_hit <- function(value) {
  if (is.na(value) || value == "") return(FALSE)
  hit_status <- variant_info$is_hit[variant_info$variant == value]
  if (length(hit_status) > 0) return(hit_status)
  return(FALSE)
}

```


# Preprocess variant data

```{r}
# Standardize Missing Data and Low Quality Flags
# Replace placeholders / low-quality flags with "None"

to_replace <- c("NoSNVsOnlyWXSThusNoSVsChecked", "LOH_alreadyInSV", "NoneBadQuality")
to_replace2 <- c("WXSnotAnalysed", "notAnalysed")

ctnnb1_cols <- grepl("CTNNB1", names(sample_data_filtered), ignore.case = TRUE)
wt1_cols    <- grepl("WT1", names(sample_data_filtered), ignore.case = TRUE) & !grepl("loh", names(sample_data_filtered), ignore.case = TRUE)
other_cols  <- !grepl("WT1|CTNNB1", names(sample_data_filtered), ignore.case = TRUE)
wt1loh_cols <- grepl("wt1_loh", names(sample_data_filtered), ignore.case = TRUE)


sample_data_filtered[ctnnb1_cols] <- lapply(sample_data_filtered[ctnnb1_cols], function(col) ifelse(col %in% to_replace, "None_CTNNB1", col))
sample_data_filtered[ctnnb1_cols] <- lapply(sample_data_filtered[ctnnb1_cols], function(col) ifelse(col %in% to_replace2, "noData", col))

sample_data_filtered[wt1loh_cols] <- lapply(sample_data_filtered[wt1loh_cols], function(col) ifelse(col %in% to_replace, "None", col))
sample_data_filtered[wt1loh_cols] <- lapply(sample_data_filtered[wt1loh_cols], function(col) ifelse(col %in% to_replace2, "noData", col))

sample_data_filtered[wt1_cols]    <- lapply(sample_data_filtered[wt1_cols], function(col) ifelse(col %in% to_replace, "None_WT1", col))
sample_data_filtered[wt1_cols]    <- lapply(sample_data_filtered[wt1_cols], function(col) ifelse(col %in% to_replace2, "noData", col))

sample_data_filtered[other_cols]  <- lapply(sample_data_filtered[other_cols], function(col) ifelse(col %in% to_replace, "None", col))

# Check whether all values are in file
values_inFile_afterPrep <- unique(c(sample_data_filtered$wt1_germline_oncoplot_label, 
                                    sample_data_filtered$wt1_somatic_1_oncoplot_label, 
                                    sample_data_filtered$wt1_somatic_2_oncoplot_label, 
                                    sample_data_filtered$ctnnb1_somatic_1_oncoplot_label, 
                                    sample_data_filtered$ctnnb1_somatic_2_oncoplot_label, 
                                    sample_data_filtered$wt1_loh, 
                                    sample_data_filtered$DataType)) 
values_inFile_afterPrep


values_inFile_afterPrep[!values_inFile_afterPrep %in% names(variant_colors)]
```


# Define 'two hits' logic for WT1

```{r}
#--------------------------------------------#
# Define 'two hits' logic for WT1
# WT1 considered two-hit if:
# 1) Germline + somatic
# 2) Germline + LOH
# 3) Somatic + LOH
# 4) Two somatic variants
#--------------------------------------------#
create_wt1_two_hits_row <- function(data1) {
  two_hits <- sapply(1:nrow(data1), function(i) {
    germline_hit <- is_hit(data1$wt1_germline_oncoplot_label[i])
    somatic_hit_1 <- is_hit(data1$wt1_somatic_1_oncoplot_label[i])
    somatic_hit_2 <- is_hit(data1$wt1_somatic_2_oncoplot_label[i])
    loh_hit <- is_hit(data1$wt1_loh[i])
    
    has_two_hits <- (
      (germline_hit && (somatic_hit_1 || somatic_hit_2)) ||
        (germline_hit && loh_hit) ||
        ((somatic_hit_1 || somatic_hit_2) && loh_hit) ||
        (somatic_hit_1 && somatic_hit_2)
    )
    
    return(ifelse(has_two_hits, expression(` [*]`), ""))
  })
  
  mat <- matrix(two_hits, nrow = 1)
  colnames(mat) <- data1$M_Label
  rownames(mat) <- "WT1 Two Hits"
  return(mat)
}
```

# Oncoplot 

```{r}
# -----------------------------
# Oncoplot preparation and drawing
# -----------------------------

# ---- Values to include ----
values_inFile_afterPrep <- unique(c(
  sample_data_filtered$wt1_germline_oncoplot_label, 
  sample_data_filtered$wt1_somatic_1_oncoplot_label, 
  sample_data_filtered$wt1_somatic_2_oncoplot_label, 
  sample_data_filtered$ctnnb1_somatic_1_oncoplot_label, 
  sample_data_filtered$ctnnb1_somatic_2_oncoplot_label, 
  sample_data_filtered$wt1_loh, 
  sample_data_filtered$DataType
))

# ---- Two hits row (WT1 biallelic) ----
wt1_two_hits_mat <- matrix(sapply(1:nrow(sample_data_filtered), function(i) {
  hits <- c(
    is_hit(sample_data_filtered$wt1_germline_oncoplot_label[i]),
    is_hit(sample_data_filtered$wt1_somatic_1_oncoplot_label[i]),
    is_hit(sample_data_filtered$wt1_somatic_2_oncoplot_label[i]),
    is_hit(sample_data_filtered$wt1_loh[i])
  )
  ifelse(sum(hits) >= 2, "TRUE", "FALSE")
}), nrow = 1)
colnames(wt1_two_hits_mat) <- sample_data_filtered$M_Label
rownames(wt1_two_hits_mat) <- "WT1 Two Hits"

# ---- Helper: Make heatmap with dot overlay ----
make_ht <- function(mat, name, show_names = FALSE, cell_fun = NULL) {
  Heatmap(
    mat,
    name = name,
    cluster_columns = as.dendrogram(clustering_tree),
    column_split = 5,
    show_heatmap_legend = FALSE,
    col = variant_colors,
    show_column_names = show_names,
    show_row_names = TRUE,
    row_names_gp = gpar(fontsize = 10),
    column_names_gp = gpar(fontsize = 8),
    height = unit(0.8, "cm"),
    cell_fun = cell_fun
  )
}

# ---- Custom cell function for somatic triangles + dots ----
draw_somatic_cell <- function(j, i, x, y, width, height, fill) {
  somatic1 <- sample_data_filtered$wt1_somatic_1_oncoplot_label[j]
  somatic2 <- sample_data_filtered$wt1_somatic_2_oncoplot_label[j]
  
  # Neutral background
  if(somatic1 == "noData"){
    grid.rect(x, y, width, height, gp = gpar(fill = "#ECECEC", col = NA))
  } else {
    grid.rect(x, y, width, height, gp = gpar(fill = "#D6D6D6", col = NA))
  }
  
  # Bottom-left triangle: somatic1
  if(!is.na(somatic1) && somatic1 != "None_WT1") {
    grid.polygon(
      x = c(x - width/2, x + width/2, x - width/2),
      y = c(y - height/2, y - height/2, y + height/2),
      gp = gpar(fill = get_variant_color(somatic1), col = NA)
    )
  }
  
  # Top-right triangle: somatic2
  if(!is.na(somatic2) && somatic2 != "None_WT1") {
    grid.polygon(
      x = c(x - width/2, x + width/2, x + width/2),
      y = c(y + height/2, y + height/2, y - height/2),
      gp = gpar(fill = get_variant_color(somatic2), col = NA)
    )
  }
  
  # Add white dot if either variant is OtherBiomaterial
  if(is_other_biomaterial(somatic1) || is_other_biomaterial(somatic2)) {
    grid.points(x, y, pch = 16, size = unit(1.8, "mm"), gp = gpar(col = "white"))
  }
}

# ---- Cell function for single variants (Germline, LOH, CTNNB1) ----
draw_variant_cell <- function(j, i, x, y, width, height, fill, variant_column) {
  variant <- sample_data_filtered[[variant_column]][j]
  fill_color <- get_variant_color(variant)
  
  # Draw cell
  grid.rect(x, y, width, height, gp = gpar(fill = fill_color, col = NA))
  
  # Overlay symbol logic
  if(variant == "None_WT1_Diagnostics_butNotCheckedResearch") {
    grid.text("-", x = x, y = y, gp = gpar(col = "grey70", fontsize = 12, fontface = "bold"))
  } else if(is_other_biomaterial(variant)) {
    grid.points(x, y, pch = 16, size = unit(1.8, "mm"), gp = gpar(col = "white"))
  }
}

# ---- Create heatmap matrices ----
create_variant_matrix <- function(data, colname) {
  mat <- matrix(data[[colname]], nrow = 1)
  colnames(mat) <- data$M_Label
  return(mat)
}

wt1_germline_mat   <- create_variant_matrix(sample_data_filtered, "wt1_germline_oncoplot_label")
wt1_somatic_mat    <- matrix(rep("Somatic", nrow(sample_data_filtered)), nrow = 1)
colnames(wt1_somatic_mat) <- sample_data_filtered$M_Label
wt1_loh_mat        <- create_variant_matrix(sample_data_filtered, "wt1_loh")
ctnnb1_somatic_mat <- create_variant_matrix(sample_data_filtered, "ctnnb1_somatic_1_oncoplot_label")
datatype_mat       <- create_variant_matrix(sample_data_filtered, "DataType")

rownames(wt1_germline_mat) <- "WT1 Germline"
rownames(wt1_somatic_mat) <- "WT1 Somatic"
rownames(wt1_loh_mat) <- "LOH/CN-LOH"
rownames(ctnnb1_somatic_mat) <- "CTNNB1 Somatic"
rownames(datatype_mat) <- "Somatic Datatype"



# ---- Create heatmaps ----
ht_wt1_two_hits <- Heatmap(
  wt1_two_hits_mat,
  name = "WT1 Two Hits",
  cluster_columns = as.dendrogram(clustering_tree),
  column_split = 5,
  show_heatmap_legend = FALSE,
  col = c("TRUE" = "white", "FALSE" = "white"), # all background white
  show_row_names = TRUE,
  row_names_gp = gpar(fontsize = 10, fontface = "bold"),
  show_column_names = FALSE,
  height = unit(0.8, "cm"),
  cell_fun = function(j, i, x, y, width, height, fill) {
    # Draw white background rectangle
    grid.rect(x, y, width, height, gp = gpar(fill = "white", col = NA))
    
    # Draw red * in the middle if present
    if(wt1_two_hits_mat[i, j] == "TRUE") {
      grid.text("*", x = x, y = y, gp = gpar(col = "red", fontsize = 14, fontface = "bold"))
    }
  }
)

ht_wt1_germline <- make_ht(
  wt1_germline_mat, "WT1_Germline", show_names = TRUE,
  cell_fun = function(j, i, x, y, width, height, fill) {
    draw_variant_cell(j, i, x, y, width, height, fill, "wt1_germline_oncoplot_label")
  }
)

ht_wt1_somatic <- make_ht(
  wt1_somatic_mat, 
  "WT1_Somatic", 
  cell_fun = draw_somatic_cell
)

ht_wt1_loh <- make_ht(
  wt1_loh_mat, "WT1_LOH", cell_fun = function(j,i,x,y,w,h,f){
    draw_variant_cell(j,i,x,y,w,h,f, "wt1_loh")
  }
)

ht_ctnnb1_somatic <- make_ht(
  ctnnb1_somatic_mat, "CTNNB1_Somatic", cell_fun = function(j,i,x,y,w,h,f){
    draw_variant_cell(j,i,x,y,w,h,f, "ctnnb1_somatic_1_oncoplot_label")
  }
)

ht_datatype <- make_ht(datatype_mat, "Datatype", show_names = TRUE)

# ---- Create legend with "Other biomaterial" dot ----
create_legend <- function(values_inFile) {
  
  legend_order <- c(
    "Truncating_or_Deletion","Truncating/Deletion_VUS","Complex_SV_or_SV","SV_VUS","Missense","Missense_VUS","Splice","None_WT1",
    "Exon3_Missense_Stabilising","Exon3_S45del_Stabilising","PartsExon3Deleted","None_CTNNB1",
    "CN_LOH","LOH","OtherBiomaterialCN_LOH","Gain","None",
    "WGS","WES",
    "NotTested","notAnalysed","WXSnotAnalysed","NoSNVsOnlyWXSThusNoSVsChecked",
    "BadQuality","NoneBadQuality","noData","sampleSwap","LOH_alreadyInSV"
  )
  
  group_order <- c("WT1","CTNNB1","Copy number/LOH","Datatype","Technical / missing / flags")
  
  present_variants <- variant_info %>%
    filter(variant %in% values_inFile) %>% 
    filter(!grepl("OtherBiomaterial", variant))
  
  legend_list <- list()
  
  for(group_name in group_order) {
    if(!group_name %in% present_variants$group) next
    group_data <- present_variants %>%
      filter(group == group_name) %>%
      mutate(variant = factor(variant, levels = legend_order)) %>%
      arrange(variant)
    
    legend_list[[group_name]] <- Legend(
      title = group_name,
      at = group_data$variant,
      labels = group_data$variant_label,
      graphics = lapply(seq_len(nrow(group_data)), function(i){
        fill_col <- group_data$color[i]
        has_dot <- is_other_biomaterial(group_data$variant[i])
        function(x,y,w,h){
          grid.rect(x,y,w,h,gp=gpar(fill=fill_col,col=NA))
          if(has_dot) grid.points(x,y,pch=16,size=unit(1.8,"mm"),gp=gpar(col="white"))
        }
      }),
      title_gp = gpar(fontsize=12,fontface="bold"),
      labels_gp = gpar(fontsize=10)
    )
  }
  
  # Additional information sublegend
  legend_list[["Additional information"]] <- Legend(
    title = "Additional information",
    labels = "Other biomaterial",
    type = "points",
    pch = 16,
    legend_gp = gpar(col = "white"),
    title_gp = gpar(fontsize = 12, fontface = "bold"),
    labels_gp = gpar(fontsize = 10)
  )
  
  packLegend(legends=legend_list, direction="horizontal", column_gap=unit(1.5,"cm"))
}

# ---- Combine & Draw ----
combined_heatmap <- ht_wt1_two_hits %v% ht_wt1_germline %v% ht_wt1_somatic %v% ht_wt1_loh %v% ht_ctnnb1_somatic %v% ht_datatype

draw(
  combined_heatmap,
  heatmap_legend_list = create_legend(values_inFile_afterPrep),
  annotation_legend_side = "bottom",
  column_title = "WT1 and CTNNB1 Variants",
  column_title_gp = gpar(fontsize = 14, fontface = "bold")
)

# Save PDF
# pdf(file = file.path(parameters$RESULTS, "FIGURE3", "WT1_CTNNB1_oncoplot.pdf"), width = 21, height = 6)
draw(
  combined_heatmap,
  heatmap_legend_list = create_legend(values_inFile_afterPrep),
  annotation_legend_side = "bottom",
  column_title = "WT1 and CTNNB1 Variants",
  column_title_gp = gpar(fontsize = 14, fontface = "bold")
)
# dev.off()


# write.table(sample_data_filtered %>% dplyr::select(M_Label, wt1_germline_oncoplot_label, wt1_somatic_1_oncoplot_label, wt1_somatic_2_oncoplot_label, wt1_loh, ctnnb1_somatic_1_oncoplot_label, ctnnb1_somatic_2_oncoplot_label, DataType), quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS, "FIGURE3", "RAW_DATA", "3A_WT1_Oncoplot.tsv"), sep = "\t")

```





# WT1 and CTNNB1 Frequencies Across DNAm Clusters

```{r}
# record unique variants for sanity check
uniqueVariants <- c(sample_data_filtered$wt1_germline_oncoplot_label, sample_data_filtered$wt1_somatic_1_oncoplot_label, sample_data_filtered$wt1_somatic_2_oncoplot_label, sample_data_filtered$ctnnb1_somatic_1_oncoplot_label, sample_data_filtered$ctnnb1_somatic_2_oncoplot_label ) %>% unique()

# get variants of WT1 and CTNNB1
variantsWT1 <- c("CN_LOH", "LOH", "Gain", "Truncating_or_Deletion", "Complex_SV_or_SV", "Splice", "Missense")
variantsCTNNB1 <- c("Exon3_Missense_Stabilising", "Exon3_S45del_Stabilising", "PartsExon3Deleted")
noVariants <- c("None_CTNNB1", "None_WT1", "None_WT1_Diagnostics_butNotCheckedResearch", "Missense_VUS") 
notTestedEtc <- c("NotTested", "noData") 
variantsCTNNB1OtherBiomaterial <- c("Exon3_Missense_Stabilising_OtherBiomaterial", "Exon3_S45del_Stabilising_OtherBiomaterial") # will not be needed here, but add it for sanity check to see whether all possible variants we have in the sheet are in the list!

coveredVariants <- c(variantsWT1, variantsCTNNB1, noVariants, notTestedEtc, variantsCTNNB1OtherBiomaterial)

stopifnot(all(uniqueVariants %in% coveredVariants))


# Classify samples into combined WT1 / CTNNB1 mutation categories
boxplot_data <- sample_data_filtered %>%
  mutate(WT1_CTNNB1 = case_when(
    wt1_germline_oncoplot_label %in% variantsWT1 &  wt1_somatic_1_oncoplot_label %in% variantsWT1 & ctnnb1_somatic_1_oncoplot_label %in% variantsCTNNB1 ~ "WT1biallelicAndCTNNB1",
    wt1_germline_oncoplot_label %in% variantsWT1 &  wt1_somatic_2_oncoplot_label %in% variantsWT1 & ctnnb1_somatic_1_oncoplot_label %in% variantsCTNNB1 ~ "WT1biallelicAndCTNNB1",
    wt1_germline_oncoplot_label %in% variantsWT1 &  wt1_loh %in% variantsWT1 & ctnnb1_somatic_1_oncoplot_label %in% variantsCTNNB1 ~ "WT1biallelicAndCTNNB1",
    wt1_somatic_1_oncoplot_label %in% variantsWT1 &  wt1_somatic_2_oncoplot_label %in% variantsWT1 & ctnnb1_somatic_1_oncoplot_label %in% variantsCTNNB1 ~ "WT1biallelicAndCTNNB1",
    wt1_somatic_1_oncoplot_label %in% variantsWT1 &  wt1_loh %in% variantsWT1 & ctnnb1_somatic_1_oncoplot_label %in% variantsCTNNB1 ~ "WT1biallelicAndCTNNB1",
    wt1_germline_oncoplot_label %in% variantsWT1 &  ctnnb1_somatic_1_oncoplot_label %in% variantsCTNNB1 ~ "WT1monoallelicAndCTNNB1",
    wt1_somatic_1_oncoplot_label %in% variantsWT1 & ctnnb1_somatic_1_oncoplot_label %in% variantsCTNNB1 ~ "WT1monoallelicAndCTNNB1",
    wt1_germline_oncoplot_label %in% variantsWT1 &  wt1_somatic_1_oncoplot_label %in% variantsWT1 ~ "WT1biallelic",
    wt1_germline_oncoplot_label %in% variantsWT1 &  wt1_somatic_2_oncoplot_label %in% variantsWT1 ~ "WT1biallelic",
    wt1_germline_oncoplot_label %in% variantsWT1 &  wt1_loh %in% variantsWT1 ~ "WT1biallelic",
    wt1_somatic_1_oncoplot_label %in% variantsWT1 &  wt1_somatic_2_oncoplot_label %in% variantsWT1  ~ "WT1biallelic",
    wt1_somatic_1_oncoplot_label %in% variantsWT1 &  wt1_loh %in% variantsWT1 ~ "WT1biallelic",
    wt1_germline_oncoplot_label %in% variantsWT1 & wt1_somatic_1 == "noData" ~ "WT1monoallelicGermlineAndNoSomaticData",
    wt1_germline_oncoplot_label %in% variantsWT1 ~ "WT1monoallelic",
    wt1_somatic_1_oncoplot_label %in% variantsWT1  ~ "WT1monoallelic",
    ctnnb1_somatic_1_oncoplot_label %in% variantsCTNNB1 ~ "CTNNB1",
    wt1_somatic_1_oncoplot_label == "noData" ~ "noData",
    TRUE ~ "noVariants"
  )) %>% dplyr::select(Internal_Patient_ID, WT1_CTNNB1, wt1_germline_oncoplot_label, wt1_somatic_1_oncoplot_label, wt1_somatic_2_oncoplot_label, wt1_loh, ctnnb1_somatic_1_oncoplot_label, ctnnb1_somatic_2_oncoplot_label)


sample_sheet_wWTinfo <- left_join( sample_sheet %>% filter(!is.na(DNAm_5_cluster_letters)), boxplot_data) 


sample_sheet_boxplot <- sample_sheet_wWTinfo %>%  
  filter(WT1_CTNNB1!= "noData") %>% 
  mutate(Cluster1bvsOther = factor(case_when(
    DNAm_5_cluster_letters == "1b" ~ "1b",
    TRUE ~ "Other"), levels = c("1b", "Other"))
  )



# Summarize Counts & Relative Percentages
plot_data <- sample_sheet_boxplot %>%
  group_by(Cluster1bvsOther, WT1_CTNNB1) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Cluster1bvsOther) %>%
  mutate(freq = n / sum(n)) %>%   # relative frequency within each group
  mutate(perc = freq*100)

# Control stacking order of WT1_CTNNB1_status
plot_data$WT1_CTNNB1 <- factor(plot_data$WT1_CTNNB1,
                               levels = rev(c(
                                 "WT1biallelicAndCTNNB1",
                                 "WT1biallelic",
                                 "WT1monoallelicAndCTNNB1",
                                 "WT1monoallelicGermlineAndNoSomaticData",
                                 "WT1monoallelic",
                                 "CTNNB1",
                                 "noVariants",
                                 "noData"
                               )))


# Define color mapping
status_colors <- c(
  "WT1biallelicAndCTNNB1"       = "#3d348b",  # dark red
  "WT1biallelic"                = "#7678ed",  # red
  "WT1monoallelicAndCTNNB1"      = "#f35b04",  # dark orange
  # "WT1monoallelicGermlineAndNoSomaticData"= "black",
  "WT1monoallelicGermlineAndNoSomaticData"= "#F27100",
  "WT1monoallelic"               = "#f18701",  # orange
  "CTNNB1"                      = "#f7b801",  # blue
  "noVariants" = "grey90"# grey
)


# number of samples per group
sample_sheet_boxplot %>%
  group_by(Cluster1bvsOther) %>%
  summarise(n = n(), .groups = "drop")

# Plot Relative Frequency Stacked Bar Chart
WT1_CTNNB1_plot <- ggplot(plot_data, aes(x = Cluster1bvsOther, y = freq, fill = WT1_CTNNB1)) +
  geom_bar(stat = "identity", position = "fill") +  # position="fill" makes it 100%
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = status_colors, labels = rev(c("WT1biallelicAndCTNNB1",
                                                           "WT1biallelic",
                                                           "WT1monoallelicAndCTNNB1",
                                                           "WT1monoallelicGermlineAndNoSomaticData",
                                                           "WT1monoallelic",
                                                           "CTNNB1",
                                                           "noVariants"
  ))) +
  scale_fill_manual(values = status_colors) +
  labs(
    x = "Cluster",
    y = "Relative frequency (%)",
    fill = "WT1 / CTNNB1 status",
    title = "Relative frequencies of WT1 / CTNNB1 status per Cluster \n(excluding VUS, considering LOH only for biallelic hit, not including other biomaterials)"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )+
  theme_Publication(base_size = 14)

WT1_CTNNB1_plot


# pdf(file = file.path(parameters$RESULTS, "FIGURE3", "WT1_CTNNB1_relativeFrequencies_1bvsOther.pdf"), width = 8, height = 6)
WT1_CTNNB1_plot
# dev.off()


# write.table(plot_data, quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS, "FIGURE3", "RAW_DATA", "3C_WT1_FrequenciesPerDNAmCluster.tsv"), sep = "\t")
```


```{r}
# Test overall enrichment of WT1 variants in cluster 1b vs others
sample_sheet_boxplot_stats <- sample_sheet_boxplot %>% 
  filter(WT1_CTNNB1 != "noData") %>% 
  mutate(variantsWT1 = ifelse(WT1_CTNNB1 %in% c("WT1biallelicAndCTNNB1", 
                                                "WT1biallelic", 
                                                "WT1monoallelicGermlineAndNoSomaticData",
                                                "WT1monoallelic", 
                                                "WT1monoallelicAndCTNNB1"), "Variant", "NoVariant"))


# proportion of samples with WT1 variant in 1b and in other groups
tab <- table(sample_sheet_boxplot_stats$variantsWT1, sample_sheet_boxplot_stats$Cluster1bvsall)
prop.table(tab, margin =2) # 1b: 75.86% yes, Other cluster 6.17% yes

# test whether difference is significant
chisq.test(table(sample_sheet_boxplot_stats$variantsWT1, sample_sheet_boxplot_stats$Cluster1bvsall))
```




# Statistical Associations for Manuscript Text

Statistical verification of germline/somatic WT1 and CTNNB1 mutation enrichment in cluster 1b.

```{r}
# -------------------- Prepare WT1 two-hit data --------------------
wt1_two_hits_df <- data.frame(
  M_Label = colnames(wt1_two_hits_mat),
  TwoHitsWT1 = as.vector(wt1_two_hits_mat)
)

# -------------------- Simplify mutation annotations --------------------
sample_sheet_simplified <- sample_sheet_wWTinfo %>%
  mutate(
    # WT1 germline status
    WT1_simplified_germline = case_when(
      wt1_germline_oncoplot_label %in% variantsWT1 ~ "Yes",
      wt1_germline_oncoplot_label %in% noVariants ~ "No",
      wt1_germline_oncoplot_label %in% notTestedEtc ~ "NoDataOrNotTested",
      TRUE ~ NA_character_
    ),
    
    # WT1 somatic status
    WT1_simplified_somatic = case_when(
      wt1_somatic_1_oncoplot_label %in% variantsWT1 ~ "Yes",
      wt1_somatic_1_oncoplot_label %in% noVariants ~ "No",
      wt1_somatic_1_oncoplot_label %in% notTestedEtc ~ "NoDataOrNotTested",
      TRUE ~ NA_character_
    ),
    
    # Combined WT1 status
    WT1_simplified_combined = case_when(
      WT1_simplified_germline == "Yes" ~ "Germline",
      WT1_simplified_somatic == "Yes" ~ "Somatic",
      WT1_simplified_somatic == "NoDataOrNotTested" ~ "NoDataOrNotTested",
      WT1_simplified_germline == "No" & WT1_simplified_somatic == "No" ~ "NoVariant",
      TRUE ~ NA_character_
    ),
    
    # CTNNB1 somatic status
    CTNNB1_simplified_somatic = case_when(
      ctnnb1_somatic_1_oncoplot_label %in% variantsCTNNB1 ~ "Yes",
      ctnnb1_somatic_1_oncoplot_label %in% c(noVariants, variantsCTNNB1OtherBiomaterial) ~ "No",
      ctnnb1_somatic_1_oncoplot_label %in% notTestedEtc ~ "NoDataOrNotTested",
      TRUE ~ NA_character_
    ),
    
    # Combined CTNNB1 status
    CTNNB1_simplified_combined = case_when(
      CTNNB1_simplified_somatic == "Yes" ~ "Somatic",
      CTNNB1_simplified_somatic == "NoDataOrNotTested" ~ "NoDataOrNotTested",
      CTNNB1_simplified_somatic == "No" ~ "NoVariant",
      TRUE ~ NA_character_
    )
  ) %>%
  left_join(wt1_two_hits_df, by = "M_Label")

# -------------------- WT1 germline --------------------

# Number of patients with germline WT1 pathogenic variants
table(sample_sheet_simplified$WT1_simplified_germline)

# Distribution across DNAm clusters
table(sample_sheet_simplified$WT1_simplified_germline,
      sample_sheet_simplified$DNAm_5_cluster_letters)

# Perform a chi-square test to assess association between WT1 germline status and DNAm cluster distribution
# Filter to include only samples with available germline WT1 data
# (i.e., exclude samples that were not tested or have missing data)
set.seed(1)
sample_sheet_simplified %>% 
  filter(WT1_simplified_germline != "NoDataOrNotTested") %>% 
  with(table(WT1_simplified_germline, DNAm_5_cluster_letters)) %>% 
  chisq.test()

# Fisher’s exact test was additionally performed:
set.seed(1)
sample_sheet_simplified %>% 
  filter(WT1_simplified_germline != "NoDataOrNotTested") %>% 
  with(table(WT1_simplified_germline, DNAm_5_cluster_letters)) %>% 
  fisher.test()


# -------------------- WT1 somatic --------------------

table(sample_sheet_simplified$WT1_simplified_somatic)

table(sample_sheet_simplified$WT1_simplified_combined)

# Distribution across DNAm clusters
table(sample_sheet_simplified$WT1_simplified_combined,
      sample_sheet_simplified$DNAm_5_cluster_letters)

# -------------------- Biallelic WT1 --------------------
table(sample_sheet_simplified$WT1_simplified_combined,
      sample_sheet_simplified$TwoHitsWT1,
      sample_sheet_simplified$DNAm_5_cluster_letters)

# -------------------- WT1 and CTNNB1 co-occurrence --------------------

# Within cluster 1b
sample_sheet_simplified %>%
  filter(DNAm_5_cluster_letters == "1b") %>%
  with(table(WT1_simplified_combined, CTNNB1_simplified_combined))

# Outside cluster 1b
sample_sheet_simplified %>%
  with(table(WT1_simplified_combined,
             CTNNB1_simplified_combined,
             DNAm_5_cluster_letters))

# -------------------- CTNNB1 --------------------

sample_sheet_simplified %>%
  with(table(CTNNB1_simplified_combined,
             DNAm_5_cluster_letters))

# Perform Fisher’s Exact Test to assess association between CTNNB1 status and cluster assignment
# Filter to include only samples with available CTNNB1 data
# (i.e., exclude samples that were not tested or have missing data)
sample_sheet_simplified %>%
  filter(CTNNB1_simplified_combined != "NoDataOrNotTested") %>% 
  with(table(CTNNB1_simplified_combined,
             Cluster1bvsall)) %>% 
  fisher.test()

# -------------------- Additional summaries --------------------

# Patients not tested for germline WT1
sample_sheet_boxplot %>%
  filter(wt1_germline_oncoplot_label == "NotTested")

```



# Diagnostic vs. Research Variant Calling Comparison

```{r}

sample_sheet_simplified_diagnosticsInfo <- left_join(sample_sheet_simplified, sample_data_filtered %>%
                                                       dplyr::select(Internal_Patient_ID, wt1_germline_Diagnostics,
                                                                     wt1_somatic_Diagnostics_1, wt1_somatic_Diagnostics_2, 
                                                                     ctnnb1_Diagnostics_somatic_1, ctnnb1_Diagnostics_somatic_2) )

# =====================================
# RESEARCH vs DIAGNOSTIC WT1 DETECTION
# =====================================

# Data preparation: filter samples with diagnostic information
samples_with_diagnostics_germline <- sample_sheet_simplified_diagnosticsInfo %>% 
  filter(wt1_germline_Diagnostics != "NA")
samples_with_diagnostics_somatic <- sample_sheet_simplified_diagnosticsInfo %>% 
  filter(wt1_somatic_Diagnostics_1 != "NA")



# =====================================
# GERMLINE WT1: RESEARCH vs DIAGNOSTICS COMPARISON
# =====================================

cat("GERMLINE WT1 ANALYSIS:\n")
cat("======================\n")

cat("Samples with germline diagnostic WT1 information:", nrow(samples_with_diagnostics_germline), "\n\n")

# Comparison table: research classification vs diagnostic detection
germline_comparison <- table(samples_with_diagnostics_germline$WT1_simplified_germline, 
                             samples_with_diagnostics_germline$wt1_germline_Diagnostics)

cat("Research vs Diagnostic detection of germline WT1:\n")
print(germline_comparison)



# =====================================
# SOMATIC WT1: RESEARCH vs DIAGNOSTICS COMPARISON  
# =====================================

cat("SOMATIC WT1 ANALYSIS:\n")
cat("=====================\n")

samples_with_diagnostics_somatic[samples_with_diagnostics_somatic$wt1_somatic_Diagnostics_1 == 'Found' & samples_with_diagnostics_somatic$WT1_simplified_somatic == "No", "wt1_somatic_Diagnostics_1"] <- "NoWT1variant"

cat("Somatic samples checked in diagnostics:", nrow(samples_with_diagnostics_somatic), "\n")

# Comparison table: research classification vs diagnostic detection
somatic_comparison <- table(samples_with_diagnostics_somatic$WT1_simplified_somatic, 
                            samples_with_diagnostics_somatic$wt1_somatic_Diagnostics_1 %>% factor)

cat("\nResearch vs Diagnostic detection of somatic WT1 (first variant):\n")
print(somatic_comparison)

# =====================================
# SOMATIC WT1 VARIANT #2 ANALYSIS
# =====================================

cat("\nSOMATIC WT1 VARIANT #2 ANALYSIS:\n")
cat("=================================\n")

# Filter samples that have somatic WT1 variant #2 information
samples_with_diagnostics_somatic2 <- samples_with_diagnostics_somatic %>% 
  filter(!is.na(wt1_somatic_Diagnostics_2), 
         wt1_somatic_Diagnostics_2 != "NA")

cat("Samples with somatic WT1 variant #2 diagnostic information:", nrow(samples_with_diagnostics_somatic2), "\n")

if(nrow(samples_with_diagnostics_somatic2) > 0) {
  # Create binary classification for somatic variant #2 (research)
  samples_with_diagnostics_somatic2 <- samples_with_diagnostics_somatic2 %>%
    mutate(
      WT1_somatic_2_binary = case_when(
        wt1_somatic_2_oncoplot_label == "None_WT1" ~ "No",
        wt1_somatic_2_oncoplot_label %in% variantsWT1 ~ "Yes",
        TRUE ~ NA
      )
    )
  
  # Comparison table for variant #2
  somatic_comparison_2 <- table(samples_with_diagnostics_somatic2$WT1_somatic_2_binary, 
                                samples_with_diagnostics_somatic2$wt1_somatic_Diagnostics_2 %>% factor)
  
  cat("\nResearch vs Diagnostic detection of somatic WT1 (second variant):\n")
  print(somatic_comparison_2)
  
  # Show detailed breakdown
  cat("\nDetailed breakdown of somatic WT1 variant #2:\n")
  detailed_2 <- samples_with_diagnostics_somatic2 %>% 
    select(Internal_Patient_ID, M_Label, wt1_somatic_2_oncoplot_label, WT1_somatic_2_binary, wt1_somatic_Diagnostics_2)
  print(detailed_2)
} else {
  cat("No samples with somatic WT1 variant #2 diagnostic information found\n")
}


# =====================================
# SUMMARY STATISTICS
# =====================================

cat("\nSUMMARY:\n")
cat("========\n")
{cat("Germline WT1:\n")
  cat("- Total number of samples with germline WT1 variants identified:", sum(samples_with_diagnostics_germline$WT1_simplified_germline == "Yes"), "\n")
  cat("- Number of samples with germline WT1 variants identified by diagnostics:", 
      sum(samples_with_diagnostics_germline$WT1_simplified_germline == "Yes" & 
            samples_with_diagnostics_germline$wt1_germline_Diagnostics == "Found", na.rm = TRUE), "\n")
  cat("- Number of samples with germline WT1 variants missed by diagnostics:", 
      sum(samples_with_diagnostics_germline$WT1_simplified_germline == "Yes" & 
            samples_with_diagnostics_germline$wt1_germline_Diagnostics == "NotFound", na.rm = TRUE), "\n")
  cat("- M-Label of samples with germline WT1 variants missed by diagnostics:", 
      samples_with_diagnostics_germline %>% 
        filter(WT1_simplified_germline == "Yes" & wt1_germline_Diagnostics == "NotFound") %>% 
        pull(M_Label), "\n\n")
}

{cat("Somatic WT1 (variant #1):\n")
  cat("- Samples tested diagnostically:", nrow(samples_with_diagnostics_somatic), "\n")
  cat("- Somatic Variants (1) found by diagnostics:",  sum(samples_with_diagnostics_somatic$WT1_simplified_somatic == "Yes" & 
                                                             samples_with_diagnostics_somatic$wt1_somatic_Diagnostics_1 == "Found", na.rm = TRUE), "\n")
  cat("- Somatic Variants (1) missed by diagnostics:", sum(samples_with_diagnostics_somatic$WT1_simplified_somatic == "Yes" & 
                                                             samples_with_diagnostics_somatic$wt1_somatic_Diagnostics_1 == "NotFound", na.rm = TRUE), "\n")
  cat("- M-Label of samples with somatic WT1 variants (1) missed by diagnostics:", 
      samples_with_diagnostics_somatic %>% 
        filter(WT1_simplified_somatic == "Yes" & wt1_somatic_Diagnostics_1 == "NotFound") %>% 
        pull(M_Label), "\n\n")
}

{cat("Somatic WT1 (variant #2):\n")
  cat("- Samples tested diagnostically:", nrow(samples_with_diagnostics_somatic2), "\n")
  cat("- Somatic Variants (2) found by diagnostics:",  sum(samples_with_diagnostics_somatic2$WT1_somatic_2_binary == "Yes" & 
                                                             samples_with_diagnostics_somatic2$wt1_somatic_Diagnostics_2 == "Found", na.rm = TRUE), "\n")
  cat("- Somatic Variants (2) missed by diagnostics:", sum(samples_with_diagnostics_somatic2$WT1_somatic_2_binary == "Yes" & 
                                                             samples_with_diagnostics_somatic2$wt1_somatic_Diagnostics_2 == "NotFound", na.rm = TRUE), "\n")
  cat("- M-Label of samples with somatic WT1 variants (2) missed by diagnostics:", 
      samples_with_diagnostics_somatic2 %>% 
        filter(WT1_somatic_2_binary == "Yes" & wt1_somatic_Diagnostics_2 == "NotFound") %>% 
        pull(M_Label), "\n\n")
}

```

# Save sample sheet
```{r}
saveRDS(sample_sheet_simplified, file = file.path(parameters$PROCESSED_DATA_FOLDER, "masterSampleSheet_after_figure3.rds"))
```

