---
title: "Classifier"
author: "Franziska Oberhammer"
date: "2026-09-10"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup & Environment
Loading core dependencies, helper utilities, and configuration directories.

```{r load_packages, warning=FALSE, message=FALSE}
suppressPackageStartupMessages({
  library(circlize)
  library(doParallel)
  library(dplyr)
  library(magrittr)
  library(qs)
  library(RColorBrewer)
  library(pROC)
  library(tidyverse)
  library(randomForest)
  library(umap)
  library(data.table)
  library(patchwork)
  library(ggalluvial)
  library()
  library(uwot)
  library(tibble)
})
```



```{r Configuration}
parameters <- list()

# Directory Configuration
parameters$PROJECT_FOLDER        <- "/hpc/pmc_kuiper/Wilms/METHYLATION" 
parameters$CODE_FOLDER           <- file.path(parameters$PROJECT_FOLDER, "CODE") 
parameters$DATA_FOLDER           <- file.path(parameters$PROJECT_FOLDER, "DATA") 
parameters$ANALYSIS_FOLDER       <- file.path(parameters$PROJECT_FOLDER, "ANALYSIS") 
parameters$PROCESSED_DATA_FOLDER <- file.path(parameters$ANALYSIS_FOLDER, "PROCESSED_DATA") 
parameters$RESULTS_FOLDER        <- file.path(parameters$PROJECT_FOLDER, "RESULTS")

# Input Data Paths (Training Cohort)
parameters$sampleSheetTrainCohort                <- file.path(parameters$PROCESSED_DATA_FOLDER, "masterSampleSheet_after_figure3.rds")
parameters$betaValuesTrainCohort_raw             <- file.path(parameters$PROCESSED_DATA_FOLDER, "betas_V1V2_combined_probeFiltered_SampleQCFiltered_SNPsQC_postMortemAndVersionFiltered_SexMissmatchFiltered.qs")
parameters$betaValuesTrainCohort_batchCorrected  <- file.path(parameters$PROCESSED_DATA_FOLDER, "betas_V1V2_combined_probeFiltered_SampleQCFiltered_SNPsQC_postMortemAndVersionFiltered_batchCorrected.qs")
parameters$CLUSTERING_TrainCohort                <- file.path(parameters$ANALYSIS_FOLDER, 
                                                              "EXPLORATORY_ANALYSIS/clustering/ConsensusClusterPlus_5000mvCpGs_manuscript/innerLinkageward.D2_finalLinkageward.D2_clusterAlghc_distanceMethpearson/clustering_tree_5clusters.qs")

# Source Custom Utilities
source(file.path(parameters$CODE_FOLDER, "00_theme_Publication.R"))
source(file.path(parameters$CODE_FOLDER, "00_WT_colour_palettes.R"))
source(file.path(parameters$CODE_FOLDER, "00_WT_helper_functions.R"))
source(file.path(parameters$CODE_FOLDER, "00_WTClassifierMnMHelperFunctions.R"))
```



2. Load & Preprocess Training Cohort

Loading sample sheets and methylation matrices for the discovery training cohort.

```{r LoadData}
sample_sheets <- list()

# Load training metadata
sample_sheets$sample_sheet_train    <- readRDS(parameters$sampleSheetTrainCohort) 

# Load raw and batch-corrected beta value matrices
betas_TrainCohort_raw <- qread(parameters$betaValuesTrainCohort_raw)
betas_TrainCohort_batchCorrected <- as.data.frame(qread(parameters$betaValuesTrainCohort_batchCorrected))
clustering_Train <- qread(parameters$CLUSTERING_TrainCohort)  
```

```{r PrepData}
# Filter CpGs exceeding missingness threshold (> 11% NAs)
betas_TrainCohort_raw <- betas_TrainCohort_raw[rowMeans(is.na(betas_TrainCohort_raw)) < 0.11, ]
betas_TrainCohort_batchCorrected <- betas_TrainCohort_batchCorrected[rowMeans(is.na(betas_TrainCohort_batchCorrected)) < 0.11, ]


# Preprocess methylation beta matrices (keep only train cohort samples)
betas_TrainCohort_raw <- betas_TrainCohort_raw %>% 
  dplyr::select(-c("CI_V1", "BC", "EB"))

betas_TrainCohort_batchCorrected <- betas_TrainCohort_batchCorrected %>% 
  dplyr::select(-c("EB")) #CI_V1 and BC already removed in batch corrected data


# Standardize sample column names by removing version suffixes
colnames(betas_TrainCohort_raw) <- sub("_.*", "", colnames(betas_TrainCohort_raw))

colnames(betas_TrainCohort_batchCorrected) <- sub("_.*", "", colnames(betas_TrainCohort_batchCorrected))


# Change labels to M-labels
colnames(betas_TrainCohort_raw) <- sample_sheets$sample_sheet_train$M_Label[match(colnames(betas_TrainCohort_raw), sample_sheets$sample_sheet_train$Internal_Patient_ID)]
colnames(betas_TrainCohort_batchCorrected) <- sample_sheets$sample_sheet_train$M_Label[match(colnames(betas_TrainCohort_batchCorrected), sample_sheets$sample_sheet_train$Internal_Patient_ID)]


# check whether samples are in same order as in sample sheet
stopifnot(colnames(betas_TrainCohort_raw) == sample_sheets$sample_sheet_train$M_Label)
stopifnot(colnames(betas_TrainCohort_batchCorrected) == sample_sheets$sample_sheet_train$M_Label)
```


# Classifier

## 10-Fold Cross-Validation Framework
Performing stratified 10-fold cross-validation with ANOVA feature selection and Random Forest feature importance reduction.
All additional classifier functions can be found in 00_WTClassifierMnMHelperFunctions.R
Note: The 10-fold cross-validation framework used here was adapted from the MnM RNA-seq classifier (PMC MnM Repositor: https://github.com/princessmaximacenter/MnM/tree/main). While original function architectures and variable names have been retained for workflow alignment, the pipeline has been modified to process DNA methylation M-values and variance-based probe selection instead of gene expression counts. The input should be a beta-matrix, it will be converted to M-values.

```{r}
countDataRef <-  betas_TrainCohort_batchCorrected %>%  as.matrix() 
metaDataRef <-  sample_sheets$sample_sheet_train %>% 
  mutate(Corrected_histology = as.character(Corrected_histology))


classColumn  <- "DNAm_5_cluster_letters"
higherClassColumn <-  "DNAm_5_cluster_letters_splitIn2"
domainColumn <- "Corrected_histology"
sampleColumn = "M_Label"
meanExpression = 0
nANOVAgenes = 10000
nModels = 100 
ntree = 500
nFeatures = 5000
maxSamplesPerType = NA
whichSeed = 1

outputDir <- "/hpc/pmc_kuiper/Wilms/METHYLATION/ANALYSIS/CLASSIFIER/ModelAndCpGs"

`%notin%` <- base::Negate(`%in%`)

# Check input ------------------------------------------
checkFormatInputData(sampleColumn = sampleColumn,
                     classColumn = classColumn,
                     higherClassColumn = higherClassColumn,
                     domainColumn = domainColumn,
                     metaDataRef = metaDataRef,
                     countDataRef = countDataRef,
                     outputDir = outputDir,
                     saveModel = T)

base::rownames(metaDataRef) <- metaDataRef[, sampleColumn]

# Check whether any group has too few samples
tumorEntitiesWithTooFewSamples <- base::table(metaDataRef[,classColumn])[base::table(metaDataRef[,classColumn]) < 3] %>% base::names()
if (base::length(tumorEntitiesWithTooFewSamples) >0) {
  
  metaDataRef %<>% dplyr::filter(!!dplyr::sym(classColumn) %notin% tumorEntitiesWithTooFewSamples)
  base::cat("\nYou have labels within your dataset that have less than 3 available samples. \nPlease note samples with these labels have been removed.\n")
}

# Convert beta matrix to M-value scale for classification modeling------------------------------------
countDataRef <- countDataRef[, base::rownames(metaDataRef)]
countDataRef <- sesame::BetaValueToMValue(countDataRef)

# Set up output directory------------------------------------------------------
directory <- outputDir
modelDirectory <- base::paste0(directory, "/seed", whichSeed)

if (!base::dir.exists(modelDirectory)){
  base::dir.create(modelDirectory)
}


# Specify max samples per tumor type------------------------------------------
if (base::is.na(maxSamplesPerType)) {
  maxSamplesPerType <- base::ceiling(stats::median(base::table(metaDataRef[, classColumn])))
}
countDataRefTransformed <- countDataRef %>% base::t() %>% base::as.data.frame()

# High-variance feature filtering (Top 10% most variable CpGs)
countDataRef <- countDataRef[complete.cases(countDataRef), ]
varVals <- apply(countDataRef, 1, var, na.rm = TRUE)   
threshold <- quantile(varVals, 0.9)             
countDataRef <- countDataRef[varVals >= threshold, ]


cat("\nWe will now start with the cross-validation\n")
# Set seed for reproducibility
base::set.seed(whichSeed)
```


```{r}
# Create splits for cross-validation setup with equal distribution of tumor types------------------------------
folds <- caret::createFolds(metaDataRef[ , classColumn], k = 10, returnTrain = TRUE, list = TRUE) 

featuresAndModels <- foreach::foreach(i=c(1:base::length(folds))) %dopar%{
  
  # Select CpGs and metadata for training samples
  metaDataCV <- metaDataRef[folds[[i]],]
  countDataCV <- countDataRef[, folds[[i]]]
  
  # Run an ANOVA to select the top n CpGs from the training data for use in the further classification process
  base::set.seed(whichSeed)
  interestingANOVAgenes <- selectAnovaGenes(metaDataRef  = metaDataCV,
                                            countDataRef  = countDataCV,
                                            nANOVAgenes = nANOVAgenes,
                                            classColumn = classColumn)
  
  base::cat(base::paste("Found interesting ANOVA genes for Fold ", i, "\n"))
  
  # Select the ANOVA genes within the log-transformed data
  dataLogCV <- countDataRefTransformed[,interestingANOVAgenes]
  
  # Select biomaterial IDs as training data per model
  base::set.seed(whichSeed)
  samplesTrainDefList <- obtainTrainData(metaDataRef = metaDataCV,
                                         classColumn = classColumn,
                                         nModels = nModels,
                                         maxSamplesPerType = maxSamplesPerType)
  
  # Select training data
  dataCV <- dataLogCV[folds[[i]], ]
  
  # Add class label to the dataset
  dataCV$class <- base::as.character(metaDataRef[rownames(dataCV),classColumn])
  
  # Also create test data and specify which biomaterial IDs are in there
  testDataCV <- dataLogCV[ -folds[[i]], ]
  
  # Reduce features using RF feature importance for accuracy
  base::cat(base::paste("Working on reducing Features for Fold", i, "\n"))
  reducedFeatures <- reduceFeatures(dataTrain = dataCV,
                                    samplesTrainDefList = samplesTrainDefList,
                                    ntree = 500,
                                    nModels = nModels,
                                    nFeatures = nFeatures,
                                    nANOVAgenes = nANOVAgenes
  )
  
  dataCV <- dataCV[,c(reducedFeatures, "class")]
  
  # Reduce features of testing data
  testDataCVReduced <- testDataCV[,reducedFeatures]
  base::cat("\nInitiating RF\n")
  
  # Start the modelling of the data within the different compositions of training data
  base::set.seed(whichSeed)
  modelList <- obtainModelsMinorityClassifier(dataTrain = dataCV,
                                              #dataSynth = data
                                              samplesTrainDefList = samplesTrainDefList,
                                              nModels = nModels,
                                              ntree = ntree)
  
  base::cat(base::paste("Generated all models for Fold ", i, "\n"))
  
  # Find the predictions for the test data
  result <- predictTest(modelList, testDataCVReduced)
  
  # Combine results, generated models and selected features into a list
  featuresAndModels <- list(result = result,
                            modelList = modelList,
                            reducedFeatures = reducedFeatures)
  
  # In the end you will end up with the resulting classifications per tumor sample,
  # the generated models and if feature reduction took place the features chosen per fold.
  return(featuresAndModels)
}
base::cat("\nFinished with generating models and results\n")


# Aggregate multi-fold classification predictions
# This is done looping over all folds.
probabilityList     <- list()
modelList           <- list()
reducedFeaturesList <- list()

for (i in seq(1:base::length(featuresAndModels))) {
  result <- featuresAndModels[[i]][["result"]]
  
  # Store (if required) models per fold within a list
  modelList[[i]] <- featuresAndModels[[i]][["modelList"]]
  
  # Store reduced features per fold within a list
  reducedFeaturesList[[i]]  <- featuresAndModels[[i]][["reducedFeatures"]]
  
  classificationResults <- convertResultToClassification(result = result,
                                                         metaDataRef = metaDataRef,
                                                         addOriginalCall = T,
                                                         classColumn = classColumn)
  
  probabilityList[[i]] <- classificationResults$probabilityList
  ultimatePredictions <- classificationResults$classifications
  
  # Add each fold to a dataframe that combines all results
  if (i == 1) {
    wrongClassifications <- ultimatePredictions[ultimatePredictions$predict != ultimatePredictions$originalCall,]
    classifications <- ultimatePredictions
  } else {
    wrongClassifications <- rbind(wrongClassifications, ultimatePredictions[ultimatePredictions$predict != ultimatePredictions$originalCall,])
    classifications <- rbind(classifications, ultimatePredictions)
  }
}

# Check the accuracy of the current run
accuracy <- base::sum(classifications$predict == classifications$originalCall) / base::length(classifications$originalCall)
base::cat(paste("\naccuracy: ", accuracy, "\n"))

# Store the settings of the classifier run within the resulting object
metaDataRun <- data.frame(nModels = nModels,
                          classColumn = classColumn,
                          higherClassColumn = higherClassColumn,
                          domainColumn = domainColumn,
                          maxSamplesPerType = maxSamplesPerType,
                          nANOVAgenes = nANOVAgenes,
                          nFeatures = nFeatures,
                          whichSeed = whichSeed)

crossValidationMinorityResults <- list(classifications = classifications,
                                       wrongClassifications = wrongClassifications,
                                       probabilityList = probabilityList,
                                       reducedFeaturesList = reducedFeaturesList,
                                       metaDataRef = metaDataRef,
                                       metaDataRun = metaDataRun)


base::cat("\nWe have finished the classification process. Please find your results in the generated object.\n")
filename <- base::paste0(modelDirectory, "/crossValidationMinorityResults.rds")

# base::saveRDS(crossValidationMinorityResults, file = filename)
base::cat(base::paste0("Please find the generated R-object with the classification results within ", filename, "\n"))
return(crossValidationMinorityResults)

filenameFeaturesAndModels <- base::paste0(modelDirectory, "/FeaturesAndModels.rds")
# base::saveRDS(featuresAndModels, file = filenameFeaturesAndModels) #uncomment if features&models should be saved
```

```{r}
#Note: Executing the 10-fold cross-validation pipeline is computationally intensive. 
# To save time, you can run the cross-validation once, serialize the output object, 
# and reload it for subsequent downstream analyses
crossValidationMinorityResults <- readRDS(filename)
FeaturesAndModels <- readRDS(base::paste0(filenameFeaturesAndModels))

```


*classifications*: All classifications for the 10x-crossvalidation setup of M&M on the reference cohort
*wrongClassifications*: All classification errors, meaning the predict is not equal to the originalCall. 
*probabilityList*: Contains how many models voted for a specific class label within the Minority classifier
*reducedFeaturesList*: Contains which RNA-transcripts were used as features within the final model generation of the Minority classifier
*metaDataRef*: Reference metadata used to generate class models
*metaDataRun*: Stored the settings that were used to run the Minority classifier with. Used for further downstream analyses.






## Cross-validation Performance & Metrics
Calculating per-class and macro-averaged metrics and constructing cross-validation confusion matrices.

### Overall Performance (Cross validation)

```{r}
# Compute confusion matrix over all cross-validation folds
cm <- caret::confusionMatrix(factor(crossValidationMinorityResults$classifications$predict, levels = c("1b", "1a", "2a", "2b", "2c")), reference = factor(crossValidationMinorityResults$classifications$originalCall, levels = c("1b", "1a", "2a", "2b", "2c")))

# extract F1 score for all classes
crossvalidation_metrics        <- cm[["byClass"]] %>% as.data.frame()
crossvalidation_metrics$Class  <-sub("Class: ", "", rownames(crossvalidation_metrics))
metric_order                   <- rev(c("Specificity", "Sensitivity", "Precision", "F1", "Balanced Accuracy"))# Define metric importance order

# Calculate macro-average performance
overall_crossvalidation_metrics_macro <- crossvalidation_metrics %>%
  select( Sensitivity, Specificity, Precision, F1, `Balanced Accuracy`) %>%
  summarise(across(everything(), mean, na.rm = TRUE)) %>%
  pivot_longer(cols = everything(), names_to = "Metric", values_to = "Value") %>% 
  mutate(Metric = factor(Metric, levels = metric_order))


# Plot Overall Performance
overall_crossvalidation_metrics_macro_plot <- ggplot(overall_crossvalidation_metrics_macro, aes(x = Metric, y = Value)) +
  geom_col(fill = "chocolate1", width = 0.6) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0,1)) +
  labs(
    title = "Overall Performance (Macro-average)",
    x = "",
    y = ""
  ) +
  theme_Publication()

overall_crossvalidation_metrics_macro_plot

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURE4", "4BExtra_CrossValidation_OverallClassifierPerformance.pdf"), width = 8, height = 3.25)
overall_crossvalidation_metrics_macro_plot
# dev.off()

overall_crossvalidation_metrics_macro_forCombinedPlot <- overall_crossvalidation_metrics_macro %>% as.data.frame() %>% 
  mutate(Class = "Overall") %>% 
  dplyr::relocate(Class, .before = Metric)




# Performance per class -----------------------
# Reshape long format per class
crossvalidation_metrics_long <- crossvalidation_metrics %>%
  select(Class, Sensitivity, Specificity, Precision, F1, `Balanced Accuracy`) %>%
  pivot_longer(cols = -Class, names_to = "Metric", values_to = "Value")

crossvalidation_metrics_long %>% 
  filter(Class == "1b")
crossValidationMetric_combined_long <- rbind(overall_crossvalidation_metrics_macro_forCombinedPlot, crossvalidation_metrics_long)

# Prepare long-format data and factor the Metric
crossValidationMetric_combined_long <- crossValidationMetric_combined_long %>%
  filter(Metric %in% metric_order) %>%   # keep only selected metrics
  mutate(
    Metric = factor(Metric, levels = metric_order)
  )

# Prepare plot with overall performance + per class performance
crossvalidation_metrics_plot <- ggplot(crossValidationMetric_combined_long, aes(x = factor(Class, levels = c("Overall", " ", names(colours_WT$DNAm_clustering2))), y = Value, fill = Class)) +
  geom_col(position = "dodge") +
  facet_wrap(~ Metric, scales = "fixed", nrow = 1) +   # same y-axis for all facets
  scale_fill_manual(values = c(colours_WT$DNAm_clustering2, "Overall" = "chocolate1"))+
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Performance Metrics (Overall & per Class)",
    x = "Class",
    y = "",
    fill = "DNAm cluster"
  ) +
  theme_Publication()

crossvalidation_metrics_plot

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURE4", "4B_CrossValidation_ClassifierPerformance_perClassAndOverall.pdf"), width = 12, height = 5)
crossvalidation_metrics_plot
# dev.off()

# write.table(crossValidationMetric_combined_long, quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE4", "RAW_DATA", "4B_CrossValidation_ClassifierPerformance_perClassAndOverall.tsv"), sep = "\t")

# write.table(crossValidationMinorityResults$classifications, quote = FALSE, row.names = TRUE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE4", "RAW_DATA", "4B_CrossValidation_ClassifierResults.tsv"), sep = "\t")
```


### Performance filtering out low-confidence predictions (Cross validation)
```{r}
# Compute confusion matrix over all cross-validation folds without low-confidence predictions (probability <= 72%)
crossValidationMinorityResults$classifications_wo_lowConfidence <- crossValidationMinorityResults$classifications %>% filter(probability > 72)

cm_woLowConfidence <- caret::confusionMatrix(
  factor(crossValidationMinorityResults$classifications_wo_lowConfidence$predict,  levels = c("1b", "1a", "2a", "2b", "2c")), 
  reference = factor(crossValidationMinorityResults$classifications_wo_lowConfidence$originalCall, levels = c("1b", "1a", "2a", "2b", "2c")))

# Calculate macro-average performance
crossvalidation_metrics_woLowConfidence       <- cm_woLowConfidence[["byClass"]] %>% as.data.frame()
crossvalidation_metrics_woLowConfidence$Class <-sub("Class: ", "", rownames(crossvalidation_metrics_woLowConfidence))


overall_crossvalidation_metrics_macro_woLowConfidence <- crossvalidation_metrics_woLowConfidence %>%
  select( Sensitivity, Specificity, Precision, F1, `Balanced Accuracy` ) %>%
  summarise(across(everything(), mean, na.rm = TRUE)) %>%
  pivot_longer(cols = everything(), names_to = "Metric", values_to = "Value") %>% 
  mutate(Metric = factor(Metric, levels = metric_order))

# Prepare overall metrics plot (without low confidence samples)
overall_crossvalidation_metrics_macro_plot_woLowConfidence <- ggplot(overall_crossvalidation_metrics_macro_woLowConfidence, aes(x = Metric, y = Value)) +
  geom_col(fill = "chocolate1", width = 0.6) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0,1)) +
  labs(
    title = "Overall Performance (Macro-average)",
    x = "",
    y = ""
  ) +
  theme_Publication()

overall_crossvalidation_metrics_macro_plot_woLowConfidence

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURES3", "S3Extra1_CrossValidation_OverallClassifierPerformance_woLowConfidence.pdf"), width = 8, height = 3.25)
overall_crossvalidation_metrics_macro_plot_woLowConfidence
# dev.off()



# Performance per class -----
# Reshape long format per class
crossvalidation_metrics_long_woLowConfidence <- crossvalidation_metrics_woLowConfidence %>%
  select( Class, Sensitivity, Specificity, Precision, F1, `Balanced Accuracy` ) %>%
  pivot_longer( cols = -Class, names_to = "Metric", values_to = "Value" )

crossvalidation_metrics_long_woLowConfidence <- crossvalidation_metrics_long_woLowConfidence %>%
  filter(Metric %in% metric_order) %>%   # keep only selected metrics
  mutate(
    Metric = factor(Metric, levels = metric_order)
  )

# Prepare plot with overall performance + per class performance (without low confidence samples)
crossvalidation_metrics_plot_woLowConfidence <- ggplot(crossvalidation_metrics_long_woLowConfidence, aes(x = Class, y = Value, fill = Class)) +
  geom_col(position = "dodge") +
  facet_wrap(~ Metric, scales = "fixed", nrow = 1) +   # same y-axis for all facets
  scale_fill_manual(values = colours_WT$DNAm_clustering2)+
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Performance Metrics per Class",
    x = "Class",
    y = "",
    fill = "DNAm cluster"
  ) +
  theme_Publication()

crossvalidation_metrics_plot_woLowConfidence

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURES3", "S3Extra2_CrossValidation_OverallClassifierPerformance_perClass_woLowConfidence.pdf"), width = 8.5, height = 3.25)
crossvalidation_metrics_plot_woLowConfidence
# dev.off()

# write.table(crossValidationMinorityResults$classifications_wo_lowConfidence, quote = FALSE, row.names = TRUE, file = file.path(parameters$RESULTS_FOLDER, "FIGURES3", "RAW_DATA", "S3B_CrossValidation_ClassifierResults_woLowConfidence.tsv"), sep = "\t")

# write.table(crossvalidation_metrics_long_woLowConfidence, quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURES3", "RAW_DATA", "S3Extra2_CrossValidation_OverallClassifierPerformance_perClass_woLowConfidence.tsv"), sep = "\t")

```


### Confusion matrix (Cross validation)

```{r}
# prepare confusion matrix
conf_mat_crossval <- make_conf_mat_simple(
  crossValidationMinorityResults$classifications,
  original_col = "originalCall",
  predicted_col = "predict"
)

# prepare confusion matrix for cohort without low confidence samples
conf_mat_crossval_wo_lowConfidence <- make_conf_mat_simple(
  crossValidationMinorityResults$classifications_wo_lowConfidence,
  original_col = "originalCall",
  predicted_col = "predict"
)


# Plots of confusion matrix 
plot_conf_mat(conf_mat_crossval, label_type = "count")
plot_conf_mat(conf_mat_crossval, label_type = "count_prob")
plot_conf_mat(conf_mat_crossval, label_type = "count_relfreq")

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURE4", "4CExtra2_CrossValidation_ConfusionMatrix_CountRelativeFreq.pdf"), width = 7, height = 5)
plot_conf_mat(conf_mat_crossval, label_type = "count_relfreq")
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURE4", "4CExtra3_CrossValidation_ConfusionMatrix_CountProability.pdf"), width = 7, height = 5)
plot_conf_mat(conf_mat_crossval, label_type = "count_prob")
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURE4", "4C_CrossValidation_ConfusionMatrix_Count.pdf"), width = 7, height = 5)
plot_conf_mat(conf_mat_crossval, label_type = "count")
# dev.off()

# write.table(conf_mat_crossval %>% dplyr::select(originalCall, predict, n, avg_prob, freq), quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE4", "RAW_DATA", "4C_CrossValidation_ConfusionMatrix_Count.tsv"), sep = "\t")



# Plots of confusion matrix without low confidence samples ------------------------
plot_conf_mat(conf_mat_crossval_wo_lowConfidence, label_type = "count_prob")
plot_conf_mat(conf_mat_crossval_wo_lowConfidence, label_type = "count_relfreq")


# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURES3", "S3Extra3_CrossValidation_ConfusionMatrix_CountRelativeFreq_wo_lowConfidence.pdf"), width = 7, height = 5)
plot_conf_mat(conf_mat_crossval_wo_lowConfidence, label_type = "count_relfreq")
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURES3", "S3Extra3_CrossValidation_ConfusionMatrix_CountProability_wo_lowConfidence.pdf"), width = 7, height = 5)
plot_conf_mat(conf_mat_crossval_wo_lowConfidence, label_type = "count_prob")
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURES3", "S3B_CrossValidation_ConfusionMatrix_CountProability_wo_lowConfidence.pdf"), width = 7, height = 5)
plot_conf_mat(conf_mat_crossval_wo_lowConfidence, label_type = "count")
# dev.off()


# write.table(conf_mat_crossval_wo_lowConfidence %>% dplyr::select(originalCall, predict, n, avg_prob, freq), quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURES3", "RAW_DATA", "S3B_CrossValidation_ConfusionMatrix_CountProability_wo_lowConfidence.tsv"), sep = "\t")

```



```{r}
# Misclassified samples table
misclassified_table <- crossValidationMinorityResults$classifications %>%
  rownames_to_column(var = "sample") %>%
  filter(originalCall != predict) %>%
  select(
    Sample = sample,
    True_Class = originalCall,
    Predicted_Class = predict,
    Probability = probability
  ) %>%
  arrange(True_Class) 

misclassified_table
```


### Snake diagram

```{r}
# Classification flow visualization

# Prepare full classification flow
all_flow <- crossValidationMinorityResults$classifications %>%
  rownames_to_column(var = "sample") %>%
  mutate(classification_status = if_else(originalCall == predict,
                                         "Correct classification",
                                         "Misclassified")) %>%
  group_by(originalCall, predict, classification_status) %>%
  summarise(n = n(), .groups = "drop")

# Define class order for legend
class_order <- c("1a", "1b", "2a", "2b", "2c")  

# Alluvial plot
classification_flow_plot <- ggplot(all_flow,
                                   aes(axis1 = originalCall, axis2 = predict, y = n, fill = originalCall)) +
  geom_alluvium(aes(alpha = classification_status), width = 1/12) +
  geom_stratum(width = 1/12, fill = "grey80", color = "black") +
  geom_text(stat = "stratum", aes(label = paste(after_stat(stratum), "\n", n)), size = 3) +
  scale_x_discrete(limits = c("True class", "Predicted class")) +
  scale_fill_manual(values = colours_WT$DNAm_clustering2,
                    breaks = class_order,
                    labels = class_order) +
  scale_alpha_manual(values = c(`Correct classification` = 0.3,
                                `Misclassified` = 1)) +  # highlight misclassified flows
  theme_minimal() +
  labs(title = "Classification Flow (Cross-Validation)",
       y = "Number of samples", x = "", fill = "True class (DNAm cluster)", alpha = "Classification status")

classification_flow_plot 

# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURES3", "S3Extra6_CrossValidation_ClassificationFlow_misclassified.pdf"), width = 9, height = 8)
classification_flow_plot 
# dev.off()
```



# Create final classifier model
```{r}
saveModel = T
correctRibo = F

countDataOG <- countDataRef
`%notin%` <<- base::Negate(`%in%`)

checkFormatInputData(sampleColumn = sampleColumn,
                     classColumn = classColumn,
                     higherClassColumn = higherClassColumn,
                     domainColumn = domainColumn,
                     metaDataRef = metaDataRef,
                     countDataRef = countDataRef,
                     outputDir = outputDir,
                     saveModel = saveModel)

tumorEntitiesWithTooFewSamples <- base::table(metaDataRef[,classColumn])[base::table(metaDataRef[,classColumn]) < 3] %>% base::names()
base::rownames(metaDataRef) <- metaDataRef[, sampleColumn]

if (base::length(tumorEntitiesWithTooFewSamples) >0) {
  
  metaDataRef %<>% dplyr::filter(!!dplyr::sym(classColumn) %notin% tumorEntitiesWithTooFewSamples)
  base::print("You have labels within your dataset that have less than 3 available samples.  Please note samples with these labels have been removed.")
  #stop("You have tumor subtypes within your dataset that have less than 3 available samples. Please remove all tumor types with too few samples. ")
  
}

# Convert beta matrix to M-value scale for classification modeling
countDataRef <- countDataRef[, base::rownames(metaDataRef)]
countDataRef <- sesame::BetaValueToMValue(countDataRef)

# Prepare Output directory
directory <- outputDir

# transfer data
dataRef <- countDataRef %>% base::t() %>% base::as.data.frame()


# Specify max samples per tumor type
if (base::is.na(maxSamplesPerType)) {
  maxSamplesPerType <- base::ceiling(stats::median(base::table(metaDataRef[, classColumn])))
}

# Filter CpGs
countDataRef <- countDataRef[complete.cases(countDataRef), ] 
varVals <- apply(countDataRef, 1, var, na.rm = TRUE)   
threshold <- quantile(varVals, 0.9)         
countDataRef <- countDataRef[varVals >= threshold, ] 



base::print("We will now start with the selection of ANOVA-genes") # --------------------------------------
# Set seed for reproducibility
base::set.seed(whichSeed)


# Run an ANOVA to select the top n genes from the training data for use in the further classification process
interestingANOVAgenes <- selectAnovaGenes(metaDataRef = metaDataRef,
                                          countDataRef = countDataRef,
                                          nANOVAgenes = nANOVAgenes, # How many ANOVA genes
                                          classColumn = classColumn)

base::print("We have finished with the ANOVA gene selection")
# Select the ANOVA genes within the M-transformed data
dataRef <- dataRef[ ,interestingANOVAgenes]

# Select biomaterial IDs as training data per model
base::set.seed(whichSeed)
samplesTrainDefList <- obtainTrainData(metaDataRef = metaDataRef,
                                       classColumn = classColumn,
                                       nModels = nModels,
                                       maxSamplesPerType = maxSamplesPerType)

# Add class label to the dataset
dataRef$class <- base::as.character(metaDataRef[rownames(dataRef),classColumn])

base::print("Starting to reduce features")
# Reduce features using RF feature importance for accuracy
base::set.seed(whichSeed)
reducedFeatures <- reduceFeatures(dataTrain = dataRef,
                                  samplesTrainDefList = samplesTrainDefList,
                                  ntree = 500,
                                  nModels = nModels,
                                  nFeatures = nFeatures,
                                  nANOVAgenes = nANOVAgenes)


dataRef <- dataRef[,c(reducedFeatures, "class")]

base::print("Initiating RF")

# Start the modelling of the data within the different compositions of training data
base::set.seed(whichSeed)
modelList <- obtainModelsMinorityClassifier(dataTrain = dataRef,
                                            samplesTrainDefList = samplesTrainDefList,
                                            nModels = nModels,
                                            ntree = ntree)

# Store the settings of the classifier run within the resulting object
metaDataRun <- base::data.frame(nModels = nModels,
                                classColumn = classColumn,
                                higherClassColumn = higherClassColumn,
                                domainColumn = domainColumn,
                                maxSamplesPerType = maxSamplesPerType,
                                nANOVAgenes = nANOVAgenes,
                                nFeatures = nFeatures,
                                whichSeed = whichSeed)

createdModelsMinority <- base::list(modelList = modelList,
                                    #                             riboModelList = riboModelList,
                                    reducedFeatures = reducedFeatures,
                                    metaDataRef = metaDataRef,
                                    metaDataRun = metaDataRun,
                                    countDataRef = countDataOG)


filename_finalModel <- base::paste0(directory,  "/seed", whichSeed, "/createdModelsMinority.rds")
# base::saveRDS(createdModelsMinority, file = filename_finalModel)
# base::print(base::paste0("Please find the generated R-object with the created minority classification models within ", filename_finalModel))


createdModelsMinority <- readRDS(filename_finalModel)

```



# Independent Validation Cohort 
Evaluating model performance against internal and external validation datasets.

## Setup Validation

```{r setup Validation}
parameters$sampleSheetValidationCohorts                <- file.path(parameters$PROCESSED_DATA_FOLDER, "mastersamplesheet_afterQC_validationCohort.rds")
parameters$betaValuesValidationCohorts                 <- file.path(parameters$PROCESSED_DATA_FOLDER, "betas_V1V2_combined_probeFiltered_SampleQCFiltered_validationCohort.qs") 
parameters$additionalMetadata_ExternalValidationCohort <- "/hpc/pmc_kuiper/Wilms/EXTERNAL_DATA/2025_Treger_CancerDiscovery_SomaticFootprintPredisposition/DATA/SampleSheetTreger.xlsx"
```



## Load Validation Data

```{r}
# Load beta value matrices
betas_ValidationCohorts <- qread(parameters$betaValuesValidationCohorts) %>% 
  dplyr::rename(DA_FFPE = DA) %>% 
  dplyr::select(-"DA.1")


# Load external metadata mapping
sample_sheets$additionalMetadata_ExternalValidationCohort <- readxl::read_xlsx(parameters$additionalMetadata_ExternalValidationCohort)

# Clean metadata and align patient labels
sample_sheets$sample_sheet_validation <- readRDS(parameters$sampleSheetValidationCohorts) %>% 
  mutate(
    Internal_Patient_ID = ifelse(Internal_Patient_ID == "DA", "DA_FFPE", Internal_Patient_ID),
    BatchDNAm = ifelse(grepl("PD([0-9])\\w+", Internal_Patient_ID), "ExternalValidation", "InternalValidation"),
    bisConversionControl_testResult = as.logical(bisConversionControl_testResult),
    M_Label = ifelse(is.na(M_Label), Internal_Patient_ID, M_Label)
  ) %>% 
  filter(Internal_Patient_ID %in% colnames(betas_ValidationCohorts)) %>% 
  group_by(M_Label) %>%
  mutate(
    M_Label = if (n() == 1) {
      M_Label
    } else {
      paste0(M_Label, "_", row_number())
    }
  ) %>%
  ungroup()

# Check dataset dimensions
dim(betas_ValidationCohorts)
dim(sample_sheets$sample_sheet_validation)

# Synchronize matrix columns with sample sheet IDs
betas_ValidationCohorts <- betas_ValidationCohorts[,sample_sheets$sample_sheet_validation$Internal_Patient_ID]

#Rename betas to use M-value labels instead of Intenral IDs
colnames(betas_ValidationCohorts) <- sample_sheets$sample_sheet_validation$M_Label[match(colnames(betas_ValidationCohorts), sample_sheets$sample_sheet_validation$Internal_Patient_ID)]

stopifnot(colnames(betas_ValidationCohorts) == sample_sheets$sample_sheet_validation$M_Label)



# Separate Internal & External Validation Cohorts ---------------------------
sample_sheets$sample_sheet_internalValidation <- sample_sheets$sample_sheet_validation %>% filter(BatchDNAm == "InternalValidation")%>% 
  select(where(~ !any(is.na(.))))

sample_sheets$sample_sheet_externalValidation <- sample_sheets$sample_sheet_validation %>% filter(BatchDNAm == "ExternalValidation") %>% 
  select(where(~ !any(is.na(.))))

# check number of samples per cohort
dim(sample_sheets$sample_sheet_internalValidation)
dim(sample_sheets$sample_sheet_externalValidation)
```




## IC1/IC2 Methylation Status of Internal & External Validation Cohorts

```{r}
# Load Array Manifests & Extract Imprinting Control Regions (IC1/IC2)
annoV1 <- read.delim(file.path(parameters$DATA_FOLDER, "BIN", "EPIC.hg38.manifest.tsv"))
annoV2 <- read.delim(file.path(parameters$DATA_FOLDER, "BIN", "EPICv2.hg38.manifest.tsv"))


# get CpGs that are in IC1
CpGs_IC1_EPICv1 <- annoV1 %>% 
  filter(CpG_chrm == "chr11", CpG_beg >= 1998149, CpG_end <=1998777) %>% pull(Probe_ID)
CpGs_IC1_EPICv2 <- annoV1 %>% 
  filter(CpG_chrm == "chr11", CpG_beg >= 1998149, CpG_end <=1998777) %>% pull(Probe_ID)
stopifnot(all(CpGs_IC1_EPICv1==CpGs_IC1_EPICv2))

# get CpGs that are in IC2
CpGs_IC2_EPICv1 <- annoV1 %>% 
  filter(CpG_chrm == "chr11", CpG_beg >= 2699310, CpG_end <=2700164) %>% pull(Probe_ID)
CpGs_IC2_EPICv2 <- annoV1 %>% 
  filter(CpG_chrm == "chr11", CpG_beg >= 2699310, CpG_end <=2700164) %>% pull(Probe_ID)
stopifnot(all(CpGs_IC2_EPICv2 == CpGs_IC2_EPICv1))


# Extract validation cohort beta values for IC1 and IC2 probes
CpGs_IC1 <- CpGs_IC1_EPICv1[CpGs_IC1_EPICv1 %in% rownames(betas_ValidationCohorts)]
CpGs_IC2 <- CpGs_IC2_EPICv1[CpGs_IC2_EPICv1 %in% rownames(betas_ValidationCohorts)]

IC1_meth <-betas_ValidationCohorts[CpGs_IC1,] %>% t()
IC2_meth <-betas_ValidationCohorts[CpGs_IC2,] %>% t()

colnames(IC1_meth) <- paste0(colnames(IC1_meth), "_IC1")
colnames(IC2_meth) <- paste0(colnames(IC2_meth), "_IC2")


IC1_IC2_combined <- cbind(IC1_meth, IC2_meth)
IC1_IC2_combined <- IC1_IC2_combined[match(sample_sheets$sample_sheet_validation$M_Label, rownames(IC1_IC2_combined)),] 

IC1_IC2_combined_scaled_transformed <- IC1_IC2_combined %>% scale() %>% t()

# clean global environment
rm(annoV1, annoV2, CpGs_IC1_EPICv1, CpGs_IC1_EPICv2, CpGs_IC2_EPICv1, CpGs_IC2_EPICv2, CpGs_IC1, CpGs_IC2, IC1_meth, IC2_meth)



# Consensus Clustering of CpGs in IC1 IC2 ----------------------------------------
innerLinkage <- "ward.D2"
finalLinkage <- "ward.D2"
clusterAlg <- "hc"
distanceMeth <- "euclidean"

results = ConsensusClusterPlus::ConsensusClusterPlus(
  IC1_IC2_combined_scaled_transformed,
  maxK=16,
  reps=1000,
  pItem=0.8,
  pFeature=1,
  innerLinkage = innerLinkage,
  finalLinkage = finalLinkage,
  clusterAlg = clusterAlg,
  distance = distanceMeth,
  title = file.path(parameters$ANALYSIS_FOLDER,
                    "EXPLORATORY_ANALYSIS/clustering/ConsensusClusterPlus_IC1_IC2_ValidationCohorts",
                    paste0("innerLinkage", innerLinkage,
                           "_finalLinkage", finalLinkage,
                           "_clusterAlg", clusterAlg,
                           "_distanceMeth", distanceMeth)),
  seed=1262118388.71279,
  # plot="pdf" # Uncomment to export consensus cluster decision plots
)

# Calculate Item Consensus
icl = ConsensusClusterPlus::calcICL(
  results,
  title = file.path(parameters$ANALYSIS_FOLDER,
                    "EXPLORATORY_ANALYSIS/clustering/ConsensusClusterPlus_IC1_IC2_ValidationCohorts",
                    paste0("innerLinkage", innerLinkage,
                           "_finalLinkage", finalLinkage,
                           "_clusterAlg", clusterAlg,
                           "_distanceMeth", distanceMeth)),
  # plot = "pdf"  # Uncomment to export plots
)


# 4 clusters were found to be most stable 
IC1IC2_consensus_clustering <- results[[4]]
IC1IC2_consensus_clustering_tree <- IC1IC2_consensus_clustering$consensusTree
IC1IC2_consensus_clustering_tree$labels <- rownames(IC1_IC2_combined)
IC1IC2_consensus_clustering$consensusClass


# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURES3", "S3D_ValidationCohorts_IC1IC2_withoutRownames.pdf"), width = 7, height = 7)
ComplexHeatmap::Heatmap(IC1_IC2_combined, cluster_columns = FALSE,
                        col = circlize::colorRamp2(seq(0,1,0.01), colours_WT$meth),
                        cluster_rows= as.dendrogram(IC1IC2_consensus_clustering_tree),
                        row_split = 4,
                        name = "beta-value",
                        show_row_names = FALSE,
                        row_names_side = "left", 
                        row_dend_side = "right",
                        clustering_method_rows = "complete")
# dev.off()


# write.table(IC1_IC2_combined, quote = FALSE, row.names = TRUE, file = file.path(parameters$RESULTS_FOLDER, "FIGURES3", "RAW_DATA", "S3D_ValidationCohorts_IC1_IC2.tsv"), sep = "\t")


# Annotate Imprinting Status & Merge with Sample Sheets -------------------------------------------
IC1_IC2_clustering <- data.frame("M_Label" = names(IC1IC2_consensus_clustering$consensusClass),
                                 "IC1_IC2_clustering" = (IC1IC2_consensus_clustering$consensusClass))

# Map known representative samples to define cluster-wide imprinting classes
IC1_IC2_clustering <- IC1_IC2_clustering %>%
  dplyr::mutate(
    imprinting = case_when(
      M_Label == "PD49173a" ~ "pUPD",
      M_Label == "PD52241a" ~ "hyper",
      M_Label == "PD52201a" ~ "normal",
      M_Label == "PD50696a" ~ "intermediate",
      TRUE ~ NA_character_
    )
  )

# Propagate mapped imprinting state across all cluster members
IC1_IC2_clustering_imprintingInfo <- IC1_IC2_clustering %>%
  dplyr::group_by(IC1_IC2_clustering) %>%
  dplyr::mutate(
    imprinting = dplyr::first(na.omit(imprinting))
  ) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    IC1_IC2_clustering = factor(IC1_IC2_clustering),
    imprinting = factor(imprinting)
  )


# write.table(left_join(IC1_IC2_combined %>% as.data.frame() %>% rownames_to_column("M_Label"), IC1_IC2_clustering_imprintingInfo %>% dplyr::select(M_Label, imprinting)), quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURES3", "RAW_DATA", "S3D_ValidationCohorts_IC1IC2Methylation_and_imprintingStatus.tsv"), sep = "\t")


# Merge imprinting status back into metadata tables
sample_sheets$sample_sheet_validation <- sample_sheets$sample_sheet_validation %>%
  left_join(
    IC1_IC2_clustering_imprintingInfo %>%
      select(M_Label, imprinting_ref = imprinting),
    by = "M_Label"
  ) %>%
  mutate(imprinting = coalesce(imprinting, imprinting_ref), 
         imprinting_ref = NULL) 

sample_sheets$sample_sheet_externalValidation <- sample_sheets$sample_sheet_externalValidation %>%
  left_join(
    IC1_IC2_clustering_imprintingInfo %>%
      select(M_Label, imprinting),
    by = "M_Label"
  ) 

sample_sheets$sample_sheet_internalValidation <- sample_sheets$sample_sheet_internalValidation %>%
  left_join(
    IC1_IC2_clustering_imprintingInfo %>%
      select(M_Label, imprinting),
    by = "M_Label"
  )
sample_sheets$sample_sheet_internalValidation
```




## Run Classifier on Validation Cohorts

### Internal Validation cohort 

#### Classification

```{r}
validationInternalBetaValues_reducedFeature <- betas_ValidationCohorts[createdModelsMinority$reducedFeatures,sample_sheets$sample_sheet_internalValidation$M_Label]

outputDirValidation <- "/hpc/pmc_kuiper/Wilms/METHYLATION/ANALYSIS/CLASSIFIER/ModelAndCpGs/seed1/InternalAndExternalValidationCohort"

`%notin%` <<- Negate(`%in%`)

# Execute classifier on internal cohort
result_internal <- runMinorityClassifier(
  createdModelsMinorityList = createdModelsMinority,
  MethDataNew = validationInternalBetaValues_reducedFeature,
  outputDir = file.path(outputDirValidation, "Internal"),
  sample_sheet = sample_sheets$sample_sheet_internalValidation,
  whichKimputation = 3,
  saveModel = TRUE
)

# Append binary call (Cluster 1b vs. Other)
result_internal$classifications <- result_internal$classifications %>% 
  mutate(Prediction_binary = ifelse(predict == "1b", "1b", "Other"))

result_internal$classifications
result_internal$metaDataRun

# write.table(result_internal$classifications, quote = FALSE, row.names = TRUE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE4", "RAW_DATA", "4E_ValidationInternal_ClassifierResults.tsv"), sep = "\t")


```

#### Confusion matrix

```{r}
# 1. Map Known WT1 Mutational Status (Somatic, Germline, Combined) ---------------------------------

internalValidationSamples <- list()

# Somatic
internalValidationSamples$internalValidationSamples_WT1_somatic_MLabel <- c("M953AAF") # this is the patient where we could not find a variant, but the expression is so low in the tumour that there MUST be a variant. Additionally the whole chr11 is CN-LOH! so we consider it PV WT1

# germline
internalValidationSamples$internalValidationSamples_WT1_germline_MLabel <- c("M830AAE", "M873AAF", "M065AAG_1", "M065AAG_2")

# combined
internalValidationSamples$internalValidationSamples_WT1_combined <-
  unique(c(internalValidationSamples$internalValidationSamples_WT1_somatic_PatientID,
           internalValidationSamples$internalValidationSamples_WT1_somatic_MLabel,
           internalValidationSamples$internalValidationSamples_WT1_germline_PatientID,
           internalValidationSamples$internalValidationSamples_WT1_germline_MLabel))


# 2. Build Internal Validation Confusion Matrix Dataframes ---------------------------------

result_internal$classifications_inclWT1 <- result_internal$classifications %>% rownames_to_column("M_Label") %>%
  mutate(original_WT1_combined = factor(ifelse(M_Label %in% internalValidationSamples$internalValidationSamples_WT1_combined, "WT1", "No"), levels = c("WT1", "No"))) %>%
  mutate(original_WT1_somatic  = factor(ifelse(M_Label %in% internalValidationSamples$internalValidationSamples_WT1_somatic_MLabel, "WT1_somatic", "No"), levels = c("WT1_somatic", "No"))) %>%
  mutate(original_WT1_germline = factor(ifelse(M_Label %in% internalValidationSamples$internalValidationSamples_WT1_germline_MLabel, "WT1_germline", "No"), levels = c("WT1_germline", "No"))) %>%
  left_join(result_internal$metaDataRun)

result_internal$classifications_inclWT1 %>% 
  select(original_WT1_combined, M_Label, predict)

# Calculate multi-class confusion matrices
conf_mat_int_WT1_combined <- make_conf_mat_simple(result_internal$classifications_inclWT1, original_col  = "original_WT1_combined", predicted_col = "predict")
conf_mat_int_WT1_somatic  <- make_conf_mat_simple(result_internal$classifications_inclWT1, original_col  = "original_WT1_somatic", predicted_col = "predict" )
conf_mat_int_WT1_germline <- make_conf_mat_simple(result_internal$classifications_inclWT1, original_col  = "original_WT1_germline", predicted_col = "predict" )


# Calculate binary confusion matrices (1b vs Other cluster)
conf_mat_int_WT1_combined_binary <- make_conf_mat_simple(result_internal$classifications_inclWT1, original_col  = "original_WT1_combined", predicted_col = "Prediction_binary" )
conf_mat_int_WT1_somatic_binary  <- make_conf_mat_simple(result_internal$classifications_inclWT1, original_col  = "original_WT1_somatic", predicted_col = "Prediction_binary" )
conf_mat_int_WT1_germline_binary <- make_conf_mat_simple(result_internal$classifications_inclWT1, original_col  = "original_WT1_germline", predicted_col = "Prediction_binary" )



# 3. Display & Export Confusion Matrix Plots----------------------
# All clusters
plot_conf_mat(conf_mat_int_WT1_combined, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\ninternal Validation - WT1 (germline & somatic)\nPer sample", high_colour = "orange3", original_col = "original_WT1_combined")
plot_conf_mat(conf_mat_int_WT1_somatic,  label_type = "count_relfreq", title = "Confusion Matrix Heatmap\ninternal Validation - WT1 (somatic only)\nPer sample", high_colour = "orange3", original_col = "original_WT1_somatic")
plot_conf_mat(conf_mat_int_WT1_germline, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\ninternal Validation - WT1 (germline only)\nPer sample", high_colour = "orange3", original_col = "original_WT1_germline")


# Plots binary (1b vs other)
plot_conf_mat(conf_mat_int_WT1_combined_binary, predicted_col = "Prediction_binary", label_type = "count", title = "Confusion Matrix Heatmap\ninternal Validation - WT1 (germline & somatic)\nPer sample\nPredictionBinary", high_colour = "orange3", original_col = "original_WT1_combined")
plot_conf_mat(conf_mat_int_WT1_somatic_binary,  predicted_col = "Prediction_binary", label_type = "count", title = "Confusion Matrix Heatmap\ninternal Validation - WT1 (somatic only)\nPer sample\nPredictionBinary", high_colour = "orange3", original_col = "original_WT1_somatic")
plot_conf_mat(conf_mat_int_WT1_germline_binary, predicted_col = "Prediction_binary", label_type = "count", title = "Confusion Matrix Heatmap\ninternal Validation - WT1 (germline only)\nPer sample\nPredictionBinary", high_colour = "orange3", original_col = "original_WT1_germline")


# Save Plots
# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURE4", "4EExtra1_ValidationInternal_ConfusionMatrix_WT1combined_CountRelativeFreq_perSample.pdf"),
# width = 7, height = 5)
plot_conf_mat(conf_mat_int_WT1_combined, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\ninternal Validation - WT1 (germline & somatic)\nPer sample", high_colour = "orange3", original_col = "original_WT1_combined")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURE4", "4EExtra2_ValidationInternal_ConfusionMatrix_WT1somatic_CountRelativeFreq_perSample.pdf"),
# width = 7, height = 5)
plot_conf_mat(conf_mat_int_WT1_somatic,  label_type = "count_relfreq", title = "Confusion Matrix Heatmap\ninternal Validation - WT1 (somatic only)\nPer sample", high_colour = "orange3", original_col = "original_WT1_somatic")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER,  "FIGURE4", "4EExtra3_ValidationInternal_ConfusionMatrix_WT1germline_CountRelativeFreq_perSample.pdf"),
# width = 7, height = 5)
plot_conf_mat(conf_mat_int_WT1_germline, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\ninternal Validation - WT1 (germline only)\nPer sample", high_colour = "orange3", original_col = "original_WT1_germline")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()


# pdf(file.path(parameters$RESULTS_FOLDER,  "FIGURE4", "4E_ValidationInternal_ConfusionMatrix_WT1combined_Count_perSamplePredictionBinary.pdf"),
# width = 7, height = 5)
plot_conf_mat(conf_mat_int_WT1_combined_binary, predicted_col = "Prediction_binary", label_type = "count", title = "Confusion Matrix Heatmap\ninternal Validation - WT1 (germline & somatic)\nPer sample\nPredictionBinary", high_colour = "orange3", original_col = "original_WT1_combined")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()


# write.table(conf_mat_int_WT1_combined_binary %>% dplyr::select(original_WT1_combined, Prediction_binary, n, avg_prob, freq), quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE4", "RAW_DATA", "4E_ValidationInternal_ConfusionMatrix_WT1combined_Count_perSamplePredictionBinary.tsv"), sep = "\t")

```

#### Performance Metrics

```{r}
# Recode predicted values for caret evaluation
result_internal$classifications_inclWT1_caret <- result_internal$classifications_inclWT1 %>%
  mutate(predict_WT1_combined = ifelse(predict == "1b", "WT1", "No")) 


cm_internal <- caret::confusionMatrix(
  factor(result_internal$classifications_inclWT1_caret$predict_WT1_combined, levels = c("WT1", "No")),
  reference = factor(result_internal$classifications_inclWT1_caret$original_WT1_combined, levels = c("WT1", "No"))
)


# Extract summary metrics internal validation
metrics_internal <- cm_internal[["byClass"]] %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "Metric") %>%
  dplyr::rename(Value = ".") %>%
  dplyr::filter(Metric %in% c("Sensitivity", "Specificity", "Precision", "F1", "Balanced Accuracy"))

metric_order <- rev(c("Specificity", "Sensitivity", "Precision", "F1", "Balanced Accuracy"))
metrics_internal$Metric <- factor(metrics_internal$Metric, levels = metric_order)

# Function to plot performance metrics
plot_metrics <- function(metrics_df, title = "Overall Performance") {
  ggplot(metrics_df, aes(x = Metric, y = Value)) +
    geom_col(fill = "orange3", width = 0.6) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0,1)) +
    labs(title = title, x = "", y = "") +
    theme_Publication()
}

# Plot and save internal cohort performance
overallPerformancePlot_internal <- plot_metrics(metrics_internal, title = "Overall Performance - Internal Validation")
overallPerformancePlot_internal

# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURE4", "4D_ValidationInternal_OverallClassifierPerformance.pdf"), width = 8, height = 3.25)
overallPerformancePlot_internal
# dev.off()

# write.table(metrics_internal, quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE4", "RAW_DATA", "4D_ValidationInternal_OverallClassifierPerformance.tsv"), sep = "\t")

```



### External Validation cohort

#### Classification

```{r}
# Extract classifier probes for external cohort
validationExternalBetaValues_reducedFeature <- betas_ValidationCohorts[createdModelsMinority$reducedFeatures,sample_sheets$sample_sheet_externalValidation$M_Label]

`%notin%` <<- Negate(`%in%`)

# Execute classifier on external cohort
result_external <- runMinorityClassifier(
  createdModelsMinorityList = createdModelsMinority,
  MethDataNew = validationExternalBetaValues_reducedFeature,
  outputDir = file.path(outputDirValidation, "External"),
  sample_sheet = sample_sheets$sample_sheet_externalValidation,
  whichKimputation = 3,
  saveModel = TRUE
)

# Append binary call (Cluster 1b vs Other)
result_external$classifications <- result_external$classifications %>% 
  mutate(Prediction_binary = ifelse(predict == "1b", "1b", "Other"))
result_external$classifications

result_external$metaDataRun

# write.table(result_external$classifications, quote = FALSE, row.names = TRUE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE4", "RAW_DATA", "4H_ValidationExternal_ClassifierResults.tsv"), sep = "\t")

```


#### Confusion Matrix

##### Per sample

```{r}
# 1. Map Known WT1 Cases (External Cohort) --------

externalValidationSamples <- list()

## somatic
externalValidationSamples$externalValidationSamples_WT1_somatic <- sample_sheets$sample_sheet_externalValidation %>% 
  filter(WT1 == "1") %>% 
  pull(M_Label)

## germline
externalValidationSamples$externalValidationSamples_WT1_germline <- sample_sheets$sample_sheet_externalValidation %>% 
  filter(grepl("WT1", `Type of predisposition`)) %>% 
  pull(M_Label)

## combined
externalValidationSamples$externalValidationSamples_WT1_combined <- unique(c(externalValidationSamples$externalValidationSamples_WT1_somatic, externalValidationSamples$externalValidationSamples_WT1_germline))


# 2. Build Per-Sample Dataframe & Calculate Matrices----------------

result_external$classifications_inclWT1 <- result_external$classifications %>% rownames_to_column("M_Label") %>%
  mutate(original_WT1_combined = factor(ifelse(M_Label %in% externalValidationSamples$externalValidationSamples_WT1_combined, "WT1", "No"), levels = c("WT1", "No"))) %>%  
  mutate(original_WT1_somatic  = factor(ifelse(M_Label %in% externalValidationSamples$externalValidationSamples_WT1_somatic, "WT1_somatic", "No"), levels = c("WT1_somatic", "No"))) %>%
  mutate(original_WT1_germline = factor(ifelse(M_Label %in% externalValidationSamples$externalValidationSamples_WT1_germline, "WT1_germline", "No"), levels = c("WT1_germline", "No"))) %>%
  left_join(result_external$metaDataRun)

# What type of samples do we have in the external validation cohort
unique(result_external$classifications_inclWT$`Type of sample`)


# External validation Multi-class matrices -------------
conf_mat_ext_WT1_combined <- make_conf_mat_simple(
  result_external$classifications_inclWT1,
  original_col  = "original_WT1_combined",
  predicted_col = "predict"
)

conf_mat_ext_WT1_somatic <- make_conf_mat_simple(
  result_external$classifications_inclWT1,
  original_col  = "original_WT1_somatic",
  predicted_col = "predict"
)

conf_mat_ext_WT1_germline <- make_conf_mat_simple(
  result_external$classifications_inclWT1,
  original_col  = "original_WT1_germline",
  predicted_col = "predict"
)

# External validation confusion matrices binary-----

conf_mat_ext_WT1_combined_binary <- make_conf_mat_simple(
  result_external$classifications_inclWT1,
  original_col  = "original_WT1_combined",
  predicted_col = "Prediction_binary"
)

conf_mat_ext_WT1_somatic_binary <- make_conf_mat_simple(
  result_external$classifications_inclWT1,
  original_col  = "original_WT1_somatic",
  predicted_col = "Prediction_binary"
)

conf_mat_ext_WT1_germline_binary <- make_conf_mat_simple(
  result_external$classifications_inclWT1,
  original_col  = "original_WT1_germline",
  predicted_col = "Prediction_binary"
)




# 3. Display & Export Plots --------------------

plot_conf_mat(conf_mat_ext_WT1_combined, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline & somatic)\nPer sample", high_colour = "royalblue4", original_col = "original_WT1_combined")
plot_conf_mat(conf_mat_ext_WT1_somatic,  label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (somatic only)\nPer sample", high_colour = "royalblue4", original_col = "original_WT1_somatic")
plot_conf_mat(conf_mat_ext_WT1_germline, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline only)\nPer sample", high_colour = "royalblue4", original_col = "original_WT1_germline")



# Plots binary ----

plot_conf_mat(conf_mat_ext_WT1_combined_binary, label_type = "count", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline & somatic)\nPer sample\nbinary prediction", high_colour = "royalblue4", original_col = "original_WT1_combined", predicted_col = "Prediction_binary")
plot_conf_mat(conf_mat_ext_WT1_somatic_binary,  label_type = "count", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (somatic only)\nPer sample", high_colour = "royalblue4", original_col = "original_WT1_somatic", predicted_col = "Prediction_binary")
plot_conf_mat(conf_mat_ext_WT1_germline_binary, label_type = "count", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline only)\nPer sample", high_colour = "royalblue4", original_col = "original_WT1_germline", predicted_col = "Prediction_binary")



# Save Plots -----
# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURE4", "4HExtra1_ValidationExternal_ConfusionMatrix_WT1combined_CountRelativeFreq_perSample.pdf"),
# width = 7, height = 5)
plot_conf_mat(conf_mat_ext_WT1_combined, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline & somatic)\nPer sample", high_colour = "royalblue4", original_col = "original_WT1_combined")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURE4", "4HExtra2_ValidationExternal_ConfusionMatrix_WT1somatic_CountRelativeFreq_perSample.pdf"),
# width = 7, height = 5)
plot_conf_mat(conf_mat_ext_WT1_somatic,  label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (somatic only)\nPer sample", high_colour = "royalblue4", original_col = "original_WT1_somatic")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURE4", "4HExtra3_ValidationExternal_ConfusionMatrix_WT1germline_CountRelativeFreq_perSample.pdf"),
# width = 7, height = 5)
plot_conf_mat(conf_mat_ext_WT1_germline, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline only)\nPer sample", high_colour = "royalblue4", original_col = "original_WT1_germline")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()






# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURE4", "4HExtra5_ValidationExternal_ConfusionMatrix_WT1combined_Count_perSample_BinaryClassification.pdf"),
# width = 5, height = 5)
plot_conf_mat(conf_mat_ext_WT1_combined_binary, label_type = "count", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline & somatic)\nPer sample\nbinary prediction", high_colour = "royalblue4", original_col = "original_WT1_combined", predicted_col = "Prediction_binary")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURE4", "4HExtra6_ValidationExternal_ConfusionMatrix_WT1somatic_Count_perSample_BinaryClassification.pdf"),
# width = 5, height = 5)
plot_conf_mat(conf_mat_ext_WT1_somatic_binary,  label_type = "count", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (somatic only)\nPer sample", high_colour = "royalblue4", original_col = "original_WT1_somatic", predicted_col = "Prediction_binary")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURE4", "4HExtra7_ValidationExternal_ConfusionMatrix_WT1germline_Count_perSample_BinaryClassification.pdf"),
# width = 5, height = 5)
plot_conf_mat(conf_mat_ext_WT1_germline_binary, label_type = "count", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline only)\nPer sample", high_colour = "royalblue4", original_col = "original_WT1_germline", predicted_col = "Prediction_binary")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()
```



##### Per Patient

```{r}

# 1. Map Known WT1 Cases on a Per-Patient Level

## somatic
externalValidationSamples$externalValidationSamples_WT1_somatic_PerPatient <- sample_sheets$sample_sheet_externalValidation %>% 
  filter(WT1 == "1") %>% 
  pull(Case) %>% 
  unique()

## germline
externalValidationSamples$externalValidationSamples_WT1_germline_PerPatient <- sample_sheets$sample_sheet_externalValidation %>% 
  filter(grepl("WT1", `Type of predisposition`)) %>% 
  pull(Case) %>% 
  unique()

## combined
externalValidationSamples$externalValidationSamples_WT1_combined_PerPatient <- unique(c(externalValidationSamples$externalValidationSamples_WT1_somatic_PerPatient, externalValidationSamples$externalValidationSamples_WT1_germline_PerPatient))


# Aggregate sample-level calls to majority/patient-level calls
patient_predictions <- summarise_patient_predictions(result_external$classifications %>% 
                                                       rownames_to_column("M_Label"))

# Label each patient with the true class and binary predicted class
result_external$classifications_inclWT1_perPatient <- patient_predictions %>%
  mutate(original_WT1_combined = factor(ifelse(patient %in% externalValidationSamples$externalValidationSamples_WT1_combined_PerPatient, "WT1", "No"), levels = c("WT1", "No"))) %>%  
  mutate(original_WT1_somatic  = factor(ifelse(patient %in% externalValidationSamples$externalValidationSamples_WT1_somatic_PerPatient, "WT1_somatic", "No"), levels = c("WT1_somatic", "No"))) %>%
  mutate(original_WT1_germline = factor(ifelse(patient %in% externalValidationSamples$externalValidationSamples_WT1_germline_PerPatient, "WT1_germline", "No"), levels = c("WT1_germline", "No"))) %>%
  mutate(final_prediction_binary = ifelse(final_prediction == "1b", "1b", "Other"))

# Calculate patient-level multi-class confusion matrices
conf_mat_ext_WT1_combined_perPatient <- make_conf_mat_simple(
  result_external$classifications_inclWT1_perPatient,
  original_col  = "original_WT1_combined",
  predicted_col = "final_prediction",
  probability_col = "avg_probability"
)

conf_mat_ext_WT1_somatic_perPatient <- make_conf_mat_simple(
  result_external$classifications_inclWT1_perPatient,
  original_col  = "original_WT1_somatic",
  predicted_col = "final_prediction",
  probability_col = "avg_probability"
)

conf_mat_ext_WT1_germline_perPatient <- make_conf_mat_simple(
  result_external$classifications_inclWT1_perPatient,
  original_col  = "original_WT1_germline",
  predicted_col = "final_prediction",
  probability_col = "avg_probability"
)

# External validation confusion matrices binary-----
conf_mat_ext_WT1_combined_perPatient_binary <- make_conf_mat_simple(
  result_external$classifications_inclWT1_perPatient,
  original_col  = "original_WT1_combined",
  predicted_col = "final_prediction_binary",
  probability_col = "avg_probability"
)

conf_mat_ext_WT1_somatic_perPatient_binary <- make_conf_mat_simple(
  result_external$classifications_inclWT1_perPatient,
  original_col  = "original_WT1_somatic",
  predicted_col = "final_prediction_binary",
  probability_col = "avg_probability"
)

conf_mat_ext_WT1_germline_perPatient_binary <- make_conf_mat_simple(
  result_external$classifications_inclWT1_perPatient,
  original_col  = "original_WT1_germline",
  predicted_col = "final_prediction_binary",
  probability_col = "avg_probability"
)




# 2. Preview & Export Patient-Level Heatmaps ------------------------------------

plot_conf_mat(conf_mat_ext_WT1_combined_perPatient, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline & somatic)\nPer Patient", high_colour = "royalblue4", original_col = "original_WT1_combined", predicted_col = "final_prediction")

plot_conf_mat(conf_mat_ext_WT1_somatic_perPatient,  label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (somatic only)\nPer Patient", high_colour = "royalblue4", original_col = "original_WT1_somatic", predicted_col = "final_prediction")

plot_conf_mat(conf_mat_ext_WT1_germline_perPatient, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline only)\nPer Patient", high_colour = "royalblue4", original_col = "original_WT1_germline", predicted_col = "final_prediction")


# Plots binary ----

plot_conf_mat(conf_mat_ext_WT1_combined_perPatient_binary, label_type = "count", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline & somatic)\nPer Patient\nbinary prediction", high_colour = "royalblue4", original_col = "original_WT1_combined", predicted_col = "final_prediction_binary")

plot_conf_mat(conf_mat_ext_WT1_somatic_perPatient_binary,  label_type = "count", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (somatic only)\nPer Patient\nbinary prediction", high_colour = "royalblue4", original_col = "original_WT1_somatic", predicted_col = "final_prediction_binary")

plot_conf_mat(conf_mat_ext_WT1_germline_perPatient_binary, label_type = "count", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline only)\nPer Patient\nbinary prediction", high_colour = "royalblue4", original_col = "original_WT1_germline", predicted_col = "final_prediction_binary")




# Save Plots -----
# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURE4", "4HExtra8_ValidationExternal_ConfusionMatrix_WT1combined_CountRelativeFreq_perPatient.pdf"),
#     width = 7, height = 5)
plot_conf_mat(conf_mat_ext_WT1_combined_perPatient, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline & somatic)\nPer Patient", high_colour = "royalblue4", original_col = "original_WT1_combined", predicted_col = "final_prediction")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURE4", "4HExtra9_ValidationExternal_ConfusionMatrix_WT1somatic_CountRelativeFreq_perPatient.pdf"),
#     width = 7, height = 5)
plot_conf_mat(conf_mat_ext_WT1_somatic_perPatient,  label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (somatic only)\nPer Patient", high_colour = "royalblue4", original_col = "original_WT1_somatic", predicted_col = "final_prediction")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURE4", "4HExtra10_ValidationExternal_ConfusionMatrix_WT1germline_CountRelativeFreq_perPatient.pdf"),
#     width = 7, height = 5)
plot_conf_mat(conf_mat_ext_WT1_germline_perPatient, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline only)\nPer Patient", high_colour = "royalblue4", original_col = "original_WT1_germline", predicted_col = "final_prediction")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()


# Save binary Plots -----
# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURE4", "4H_ValidationExternal_ConfusionMatrix_WT1combined_Count_perPatient_binary.pdf"),
#     width = 5, height = 5)
plot_conf_mat(conf_mat_ext_WT1_combined_perPatient_binary, label_type = "count", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline & somatic)\nPer Patient\nbinary prediction", high_colour = "royalblue4", original_col = "original_WT1_combined", predicted_col = "final_prediction_binary")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURE4", "4HExtra12_ValidationExternal_ConfusionMatrix_WT1somatic_Count_perPatient_binary.pdf"),
#     width = 5, height = 5)
plot_conf_mat(conf_mat_ext_WT1_somatic_perPatient_binary,  label_type = "count", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (somatic only)\nPer Patient\nbinary prediction", high_colour = "royalblue4", original_col = "original_WT1_somatic", predicted_col = "final_prediction_binary")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURE4", "4HExtra13_ValidationExternal_ConfusionMatrix_WT1germline_Count_perPatient_binary.pdf"),
#     width = 5, height = 5)
plot_conf_mat(conf_mat_ext_WT1_germline_perPatient_binary, label_type = "count", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline only)\nPer Patient\nbinary prediction", high_colour = "royalblue4", original_col = "original_WT1_germline", predicted_col = "final_prediction_binary")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()

# write.table(conf_mat_ext_WT1_combined_perPatient_binary %>% dplyr::select(original_WT1_combined, final_prediction_binary, n, avg_prob, freq), quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE4", "RAW_DATA", "4H_ValidationExternal_ConfusionMatrix_WT1combined_Count_perPatient_binary.tsv"), sep = "\t")


```

#### Performance Metrics

```{r}
# Recode predicted values for caret evaluation
result_external$classifications_inclWT1_caret <- result_external$classifications_inclWT1 
result_external$classifications_inclWT1_perPatient_caret <- result_external$classifications_inclWT1_perPatient 


result_external$classifications_inclWT1_caret <- result_external$classifications_inclWT1_caret %>%
  mutate(predict_WT1_combined = ifelse(predict == "1b", "WT1", "No")) 


result_external$classifications_inclWT1_perPatient_caret <-  result_external$classifications_inclWT1_perPatient_caret %>%
  mutate(predict_WT1_combined = ifelse(final_prediction == "1b", "WT1", "No"))

cm_sample <- caret::confusionMatrix(
  factor(result_external$classifications_inclWT1_caret$predict_WT1_combined,  levels = c("WT1", "No")),
  reference = factor(result_external$classifications_inclWT1_caret$original_WT1_combined,  levels = c("WT1", "No"))
)

cm_patient <- caret::confusionMatrix(
  factor(result_external$classifications_inclWT1_perPatient_caret$predict_WT1_combined,  levels = c("WT1", "No")),
  reference = factor(result_external$classifications_inclWT1_perPatient_caret$original_WT1_combined,  levels = c("WT1", "No"))
)

# Extract sample-level and patient-level metrics
metrics_sample <- cm_sample[["byClass"]] %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "Metric") %>%
  dplyr::rename(Value = ".") %>%
  dplyr::filter(Metric %in% c("Sensitivity", "Specificity", "Precision", "F1", "Balanced Accuracy"))

metrics_patient <- cm_patient[["byClass"]] %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "Metric") %>%
  dplyr::rename(Value = ".") %>%
  dplyr::filter(Metric %in% c("Sensitivity", "Specificity", "Precision", "F1", "Balanced Accuracy"))

metric_order <- rev(c("Specificity", "Sensitivity", "Precision", "F1", "Balanced Accuracy"))
metrics_sample$Metric <- factor(metrics_sample$Metric, levels = metric_order)

# Generate and display plots
plot_metrics <- function(metrics_df, title = "Overall Performance") {
  ggplot(metrics_df, aes(x = Metric, y = Value)) +
    geom_col(fill = "#2E3B87", width = 0.6) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0,1)) +
    labs(title = title, x = "", y = "") +
    theme_Publication()
}

# Plot per sample
overallPerformancePlot_PerSample <- plot_metrics(metrics_sample, title = "Overall Performance - External Validation (Per Sample)")

# Plot per patient
overallPerformancePlot_PerPatient <- plot_metrics(metrics_patient, title = "Overall Performance - External Validation (Per Patient)")


overallPerformancePlot_PerSample
overallPerformancePlot_PerPatient




# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURE4", "4GExtra_ValidationExternal_OverallClassifierPerformance_PerSample.pdf"), width = 8, height = 3.25)
overallPerformancePlot_PerSample
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURE4", "4G_ValidationExternal_OverallClassifierPerformance_PerPatient.pdf"), width = 8, height = 3.25)
overallPerformancePlot_PerPatient
# dev.off()


# write.table(metrics_patient, quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE4", "RAW_DATA", "4G_ValidationExternal_OverallClassifierPerformance_PerPatient.tsv"), sep = "\t")

```

```{r}
# Check misclassifications 
result_external$classifications_inclWT1_perPatient %>% 
  filter(original_WT1_combined == "WT1" & final_prediction != "1b")

result_external$classifications_inclWT1_perPatient %>% 
  filter(original_WT1_combined == "No" & final_prediction == "1b")
```


#### Confusion matrix without low quality samples

##### Per sample

```{r}

# use only samples with high confidence
result_external$classifications_woLowProbability <- result_external$classifications %>% filter(probability > 72)

### per sample
result_external$classifications_inclWT1_woLowProbability <- result_external$classifications_woLowProbability %>% rownames_to_column("M_Label") %>%
  mutate(original_WT1_combined = factor(ifelse(M_Label %in% externalValidationSamples$externalValidationSamples_WT1_combined, "WT1", "No"), levels = c("WT1", "No"))) %>%  
  mutate(original_WT1_somatic  = factor(ifelse(M_Label %in% externalValidationSamples$externalValidationSamples_WT1_somatic, "WT1_somatic", "No"), levels = c("WT1_somatic", "No"))) %>%
  mutate(original_WT1_germline = factor(ifelse(M_Label %in% externalValidationSamples$externalValidationSamples_WT1_germline, "WT1_germline", "No"), levels = c("WT1_germline", "No"))) %>%
  left_join(result_external$metaDataRun)

# what type of samples do we have in the external validation cohort
unique(result_external$classifications_inclWT1_woLowProbability$`Type of sample`)


# External validation confusion matrices -----
conf_mat_ext_WT1_combined_woLowProbability <- make_conf_mat_simple(
  result_external$classifications_inclWT1_woLowProbability,
  original_col  = "original_WT1_combined",
  predicted_col = "predict"
)

conf_mat_ext_WT1_somatic_woLowProbability <- make_conf_mat_simple(
  result_external$classifications_inclWT1_woLowProbability,
  original_col  = "original_WT1_somatic",
  predicted_col = "predict"
)

conf_mat_ext_WT1_germline_woLowProbability <- make_conf_mat_simple(
  result_external$classifications_inclWT1_woLowProbability,
  original_col  = "original_WT1_germline",
  predicted_col = "predict"
)



# Plots ----

plot_conf_mat(conf_mat_ext_WT1_combined_woLowProbability, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline & somatic)\nPer sample (without low probability samples)", high_colour = "royalblue4", original_col = "original_WT1_combined")
plot_conf_mat(conf_mat_ext_WT1_somatic_woLowProbability,  label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (somatic only)\nPer sample (without low probability samples)", high_colour = "royalblue4", original_col = "original_WT1_somatic")
plot_conf_mat(conf_mat_ext_WT1_germline_woLowProbability, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline only)\nPer sample (without low probability samples)", high_colour = "royalblue4", original_col = "original_WT1_germline")



# Save Plots -----
# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURES3",  "3SExtra9_ValidationExternal_ConfusionMatrix_WT1combined_CountRelativeFreq_perSample_woLowProbability.pdf"),
# width = 7, height = 5)
plot_conf_mat(conf_mat_ext_WT1_combined_woLowProbability, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline & somatic)\nPer sample (without low probability samples)", high_colour = "royalblue4", original_col = "original_WT1_combined")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURES3",  "3SExtra10_ValidationExternal_ConfusionMatrix_WT1somatic_CountRelativeFreq_perSample_woLowProbability.pdf"),
# width = 7, height = 5)
plot_conf_mat(conf_mat_ext_WT1_somatic_woLowProbability,  label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (somatic only)\nPer sample (without low probability samples)", high_colour = "royalblue4", original_col = "original_WT1_somatic")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURES3",  "3SExtra11_ValidationExternal_ConfusionMatrix_WT1germline_CountRelativeFreq_perSample_woLowProbability.pdf"),
# width = 7, height = 5)
plot_conf_mat(conf_mat_ext_WT1_germline_woLowProbability, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline only)\nPer sample (without low probability samples)", high_colour = "royalblue4", original_col = "original_WT1_germline")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()

```

##### Per Patient

```{r}

# Aggregate high-confidence predictions per patient
patient_predictions_woLowProbability <- summarise_patient_predictions(result_external$classifications_woLowProbability %>% 
                                                                        rownames_to_column("M_Label"))


# Calculate patient-level matrices
result_external$classifications_inclWT1_perPatient_woLowProbability <- patient_predictions_woLowProbability %>%
  mutate(original_WT1_combined = factor(ifelse(patient %in% externalValidationSamples$externalValidationSamples_WT1_combined_PerPatient, "WT1", "No"), levels = c("WT1", "No"))) %>%   
  mutate(original_WT1_somatic  = factor(ifelse(patient %in% externalValidationSamples$externalValidationSamples_WT1_somatic_PerPatient, "WT1_somatic", "No"), levels = c("WT1_somatic", "No"))) %>%
  mutate(original_WT1_germline = factor(ifelse(patient %in% externalValidationSamples$externalValidationSamples_WT1_germline_PerPatient, "WT1_germline", "No"), levels = c("WT1_germline", "No"))) 

# External validation confusion matrices -----

conf_mat_ext_WT1_combined_perPatient_woLowProbability <- make_conf_mat_simple(
  result_external$classifications_inclWT1_perPatient_woLowProbability,
  original_col  = "original_WT1_combined",
  predicted_col = "final_prediction",
  probability_col = "avg_probability"
)

conf_mat_ext_WT1_somatic_perPatient_woLowProbability <- make_conf_mat_simple(
  result_external$classifications_inclWT1_perPatient_woLowProbability,
  original_col  = "original_WT1_somatic",
  predicted_col = "final_prediction",
  probability_col = "avg_probability"
)

conf_mat_ext_WT1_germline_perPatient_woLowProbability <- make_conf_mat_simple(
  result_external$classifications_inclWT1_perPatient_woLowProbability,
  original_col  = "original_WT1_germline",
  predicted_col = "final_prediction",
  probability_col = "avg_probability"
)



# Plots ----

plot_conf_mat(conf_mat_ext_WT1_combined_perPatient_woLowProbability, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline & somatic)\nPer Patient (without low probability samples)", high_colour = "royalblue4", original_col = "original_WT1_combined", predicted_col = "final_prediction")

plot_conf_mat(conf_mat_ext_WT1_somatic_perPatient_woLowProbability,  label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (somatic only)\nPer Patient (without low probability samples)", high_colour = "royalblue4", original_col = "original_WT1_somatic", predicted_col = "final_prediction")

plot_conf_mat(conf_mat_ext_WT1_germline_perPatient_woLowProbability, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline only)\nPer Patient (without low probability samples)", high_colour = "royalblue4", original_col = "original_WT1_germline", predicted_col = "final_prediction")


# Save Plots -----
# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURES3",  "3SExtra13_ValidationExternal_ConfusionMatrix_WT1combined_CountRelativeFreq_perPatient_woLowProbability.pdf"),
# width = 7, height = 5)
plot_conf_mat(conf_mat_ext_WT1_combined_perPatient_woLowProbability, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline & somatic)\nPer Patient (without low probability samples)", high_colour = "royalblue4", original_col = "original_WT1_combined", predicted_col = "final_prediction")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURES3",  "3SExtra14_ValidationExternal_ConfusionMatrix_WT1somatic_CountRelativeFreq_perPatient_woLowProbability.pdf"),
# width = 7, height = 5)
plot_conf_mat(conf_mat_ext_WT1_somatic_perPatient_woLowProbability,  label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (somatic only)\nPer Patient (without low probability samples)", high_colour = "royalblue4", original_col = "original_WT1_somatic", predicted_col = "final_prediction")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURES3",  "3SExtra15_ValidationExternal_ConfusionMatrix_WT1germline_CountRelativeFreq_perPatient_woLowProbability.pdf"),
# width = 7, height = 5)
plot_conf_mat(conf_mat_ext_WT1_germline_perPatient_woLowProbability, label_type = "count_relfreq", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline only)\nPer Patient (without low probability samples)", high_colour = "royalblue4", original_col = "original_WT1_germline", predicted_col = "final_prediction")+ylab("WT1 PV")+xlab("Prediction")
# dev.off()
```

##### Performance Metrics Without Low-Confidence Samples - External Validation cohort

```{r}

# Recode high-confidence predictions
result_external$classifications_inclWT1_caret_woLowProbability <- result_external$classifications_inclWT1_woLowProbability 
result_external$classifications_inclWT1_perPatient_caret_woLowProbability <- result_external$classifications_inclWT1_perPatient_woLowProbability 


result_external$classifications_inclWT1_caret_woLowProbability <- result_external$classifications_inclWT1_caret_woLowProbability %>%
  mutate(predict_WT1_combined = ifelse(predict == "1b", "WT1", "No")) 


result_external$classifications_inclWT1_perPatient_caret_woLowProbability <-  result_external$classifications_inclWT1_perPatient_caret_woLowProbability %>%
  mutate(predict_WT1_combined = ifelse(final_prediction == "1b", "WT1", "No"))

cm_sample_woLowProbability <- caret::confusionMatrix(
  factor(result_external$classifications_inclWT1_caret_woLowProbability$predict_WT1_combined,  levels = c("WT1", "No")),
  reference = factor(result_external$classifications_inclWT1_caret_woLowProbability$original_WT1_combined,  levels = c("WT1", "No"))
)

cm_patient_woLowProbability <- caret::confusionMatrix(
  factor(result_external$classifications_inclWT1_perPatient_caret_woLowProbability$predict_WT1_combined,  levels = c("WT1", "No")),
  reference = factor(result_external$classifications_inclWT1_perPatient_caret_woLowProbability$original_WT1_combined,  levels = c("WT1", "No"))
)


# Extract performance metrics
metrics_sample_woLowProbability <- cm_sample_woLowProbability[["byClass"]] %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "Metric") %>%
  dplyr::rename(Value = ".") %>%
  dplyr::filter(Metric %in% c("Sensitivity", "Specificity", "Precision", "F1", "Balanced Accuracy"))


metrics_patient_woLowProbability <- cm_patient_woLowProbability[["byClass"]] %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "Metric") %>%
  dplyr::rename(Value = ".") %>%
  dplyr::filter(Metric %in% c("Sensitivity", "Specificity", "Precision", "F1", "Balanced Accuracy"))

metric_order <- c("Specificity", "Sensitivity", "Precision", "F1", "Balanced Accuracy")
metrics_sample$Metric <- factor(metrics_sample$Metric, levels = metric_order)

# Plot function
plot_metrics <- function(metrics_df, title = "Overall Performance") {
  ggplot(metrics_df, aes(x = Metric, y = Value)) +
    geom_col(fill = "#2E3B87", width = 0.6) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0,1)) +
    labs(title = title, x = "", y = "") +
    theme_Publication()
}

# Plot per sample
overallPerformancePlot_PerSample_woLowProbability <- plot_metrics(metrics_sample_woLowProbability, title = "Overall Performance - External Validation \n(Per Sample) (without low probability samples)")

# Plot per patient
overallPerformancePlot_PerPatient_woLowProbability <- plot_metrics(metrics_patient_woLowProbability, title = "Overall Performance - External Validation \n(Per Patient) (without low probability samples)")


overallPerformancePlot_PerSample_woLowProbability
overallPerformancePlot_PerPatient_woLowProbability

# pdf(file.path(parameters$RESULTS_FOLDER,  "FIGURES3",  "3SExtra17_ValidationExternal_OverallClassifierPerformance_PerSample_woLowProbability.pdf"), width = 8, height = 3.25)
overallPerformancePlot_PerSample_woLowProbability
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER,  "FIGURES3",  "3SExtra18_ValidationExternal_OverallClassifierPerformance_PerPatient_woLowProbability.pdf"), width = 8, height = 3.25)
overallPerformancePlot_PerPatient_woLowProbability
# dev.off()

```



## Alluvial Flow Diagrams (External Validation Cohort)

```{r}

# Map true WT1 status to expected methylation clusters-----------------------------------------------------------------
wt1_to_cluster <- list(
  WT1 = c("1b"),
  No = c("1a", "2a", "2b", "2c")
)

# Define class order for plotting
class_order <- c( "WT1", "No")  # original
pred_order <- c("1a", "1b", "2a", "2b", "2c")  # predicted

# 1. Compute Flow Summaries (Per Patient & Per Sample)
validation_flow <- result_external$classifications_inclWT1_perPatient %>%
  mutate(
    originalCall = factor(original_WT1_combined, levels = class_order),
    final_prediction = factor(final_prediction, levels = pred_order),
    # element-wise check
    classification_status = mapply(function(orig, pred) {
      if (pred %in% wt1_to_cluster[[as.character(orig)]]) {
        "Expected cluster"
      } else {
        "Unexpected cluster"
      }
    }, orig = originalCall, pred = final_prediction)
  ) %>%
  group_by(originalCall, final_prediction, classification_status) %>%
  summarise(n = n(), .groups = "drop")

# write.table(result_external$classifications_inclWT1_perPatient %>% dplyr::select(patient, original_WT1_combined, final_prediction_binary), quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURES3", "RAW_DATA", "S3C_ValidationExternal_ClassificationFlow_misclassified_PerPatient.tsv"), sep = "\t")




validation_flow_perSample <- result_external$classifications_inclWT1 %>%
  mutate(
    originalCall = factor(original_WT1_combined, levels = class_order),
    predict = factor(predict, levels = pred_order),
    # element-wise check
    classification_status = mapply(function(orig, pred) {
      if (pred %in% wt1_to_cluster[[as.character(orig)]]) {
        "Expected cluster"
      } else {
        "Unexpected cluster"
      }
    }, orig = originalCall, pred = predict)
  ) %>%
  group_by(originalCall, predict, classification_status) %>%
  summarise(n = n(), .groups = "drop")


# 2. Render & Export Alluvial Plots -----------------------------------------------------------------
# Alluvial plot
validation_flow_plot <- ggplot(validation_flow,
                               aes(axis1 = originalCall, axis2 = final_prediction, y = n, fill = originalCall)) +
  geom_alluvium(aes(alpha = classification_status), width = 1/12) +
  geom_stratum(width = 1/12, fill = "grey80", color = "black") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 3) +
  scale_x_discrete(limits = c("Original WT1 status", "Predicted DNAm cluster")) +
  scale_fill_manual(values = c( "No" = "royalblue4", "WT1" = "red4")) +
  scale_alpha_manual(values = c(`Expected cluster` = 0.3,
                                `Unexpected cluster` = 1)) +  # highlight unexpected flows
  theme_minimal() +
  labs(title = "Validation Cohort: Original WT1 vs Predicted DNAm cluster\nPer patient",
       y = "Number of samples", x = "", fill = "Original WT1", alpha = "Classification")

validation_flow_plot

# Alluvial plot
validation_flow_plot_perSample <- ggplot(validation_flow_perSample,
                                         aes(axis1 = originalCall, axis2 = predict, y = n, fill = originalCall)) +
  geom_alluvium(aes(alpha = classification_status), width = 1/12) +
  geom_stratum(width = 1/12, fill = "grey80", color = "black") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 3) +
  scale_x_discrete(limits = c("Original WT1 status", "Predicted DNAm cluster")) +
  scale_fill_manual(values = c( "No" = "royalblue4", "WT1" = "red4")) +
  scale_alpha_manual(values = c(`Expected cluster` = 0.3,
                                `Unexpected cluster` = 1)) +  # highlight unexpected flows
  theme_minimal() +
  labs(title = "Validation Cohort: Original WT1 vs Predicted DNAm cluster\nPer sample",
       y = "Number of samples", x = "", fill = "Original WT1", alpha = "Classification")

validation_flow_plot_perSample


# pdf(file.path(parameters$RESULTS_FOLDER, "FIGURES3",  "3SC_ValidationExternal_ClassificationFlow_misclassified_PerPatient.pdf"), width = 9, height = 8)
validation_flow_plot
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER,  "FIGURES3",  "3SExtra21_ValidationExternal_ClassificationFlow_misclassified_PerSample.pdf"), width = 9, height = 8)
validation_flow_plot_perSample
# dev.off()

```




### Summary plots: Classification vs WT1 status (Internal and External)

```{r}
# Aggregate prediction counts by true WT1 status
result_internal$classifications_inclWT1_summary <- result_internal$classifications_inclWT1 %>% 
  mutate(Pred = ifelse(Prediction_binary == "1b", "1b", "Other")) %>% 
  mutate(WT1_PV = ifelse(original_WT1_combined == "WT1", "Yes", "No")) %>% 
  group_by(WT1_PV, Pred) %>% 
  summarise(Count = n())


result_external$classifications_inclWT1_perPatient_summary <- result_external$classifications_inclWT1_perPatient %>% 
  mutate(Pred = ifelse(final_prediction_binary == "1b", "1b", "Other")) %>% 
  mutate(WT1_PV = ifelse(original_WT1_combined == "WT1", "Yes", "No")) %>% 
  group_by(WT1_PV, Pred) %>% 
  summarise(Count = n())


# Reorder factor levels
result_internal$classifications_inclWT1_summary$WT1_PV <- factor(result_internal$classifications_inclWT1_summary$WT1_PV, levels = c("Yes", "No"))
result_internal$classifications_inclWT1_summary$Pred   <- factor(result_internal$classifications_inclWT1_summary$Pred, c("Other", "1b"))

result_external$classifications_inclWT1_perPatient_summary$WT1_PV <- factor(result_external$classifications_inclWT1_perPatient_summary$WT1_PV, levels = c("Yes", "No"))
result_external$classifications_inclWT1_perPatient_summary$Pred   <- factor(result_external$classifications_inclWT1_perPatient_summary$Pred, levels = c("Other", "1b"))

# -------------------------
# Absolute count barplots
# -------------------------

plotInternalSummary <- ggplot(result_internal$classifications_inclWT1_summary, aes(x = WT1_PV, y = Count, fill = Pred)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("1b" = "#E69F00", "Other" = "#F5C97A")) +
  labs(title = "Internal", x = "WT1 PV", y = "Count", fill = "Prediction") +
  theme_minimal()

plotExternalSummary <- ggplot(result_external$classifications_inclWT1_perPatient_summary, aes(x = WT1_PV, y = Count, fill = Pred)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("1b" = "#0072B2", "Other" = "#9AD0F3")) +
  labs(title = "External", x = "WT1 PV", y = "Count", fill = "Prediction") +
  theme_minimal()

plotInternalSummary + plotExternalSummary

#--- 
# Calculate relative proportions (%)

result_internal$classifications_inclWT1_summary_perc <- result_internal$classifications_inclWT1_summary %>%
  group_by(WT1_PV) %>%
  mutate(Percent = Count / sum(Count) * 100)

result_external$classifications_inclWT1_perPatient_summary_perc <- result_external$classifications_inclWT1_perPatient_summary %>%
  group_by(WT1_PV) %>%
  mutate(Percent = Count / sum(Count) * 100)

# -------------------------
# Percentage stacked barplots
# -------------------------

plotInternalSummary_Perc <- ggplot(result_internal$classifications_inclWT1_summary_perc, aes(x = WT1_PV, y = Percent, fill = Pred)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("1b" = "orange3", "Other" = "#F5C97A")) +
  labs(title = "Internal", x = "WT1 PV", y = "Percentage (%)", fill = "Prediction") +
  theme(
    panel.grid = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    text = element_text(family = "Helvetica"),
    legend.direction = "vertical") +
  guides(
    colour = guide_legend(ncol = 1),
    fill = guide_legend(ncol = 1, ),
    shape = guide_legend(ncol = 1)
  )

plotExternalSummary_perc <- ggplot(result_external$classifications_inclWT1_perPatient_summary_perc, aes(x = WT1_PV, y = Percent, fill = Pred)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("1b" = "royalblue4", "Other" = "#AFBEE9")) +
  labs(title = "External", x = "WT1 PV", y = "Percentage (%)", fill = "Prediction") +
  # theme_Publication()+ 
  theme(
    panel.grid = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    text = element_text(family = "Helvetica"),
    legend.direction = "vertical") +
  guides(
    colour = guide_legend(ncol = 1),
    fill = guide_legend(ncol = 1, ),
    shape = guide_legend(ncol = 1)
  )

plotInternalSummary_Perc + plotExternalSummary_perc

# --- Save plots and data
# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURE4", "4F_ValidationInternal_SummaryWT1vsClassification.pdf"), width = 8/2.54, height = 8/2.54)
plotInternalSummary_Perc
# dev.off()

# pdf(file.path(parameters$RESULTS_FOLDER,"FIGURE4", "4I_ValidationExternal_SummaryWT1vsClassification.pdf"), width = 8/2.54, height = 8/2.54)
plotExternalSummary_perc
# dev.off()

# write.table(result_internal$classifications_inclWT1_summary_perc, quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE4", "RAW_DATA", "4F_ValidationInternal_SummaryWT1vsClassification.tsv"), sep = "\t")

# write.table(result_external$classifications_inclWT1_perPatient_summary_perc, quote = FALSE, row.names = FALSE, file = file.path(parameters$RESULTS_FOLDER, "FIGURE4", "RAW_DATA", "4I_ValidationExternal_SummaryWT1vsClassification.tsv"), sep = "\t")



```


## Combine all Validation Plots

```{r}
# Apply custom publication styling
theme_pub <- theme_classic(base_size = 6) +
  theme(
    plot.margin = ggplot2::margin(0, 0, 0, 0),
    
    axis.title = element_text( family = "Helvetica", face = "bold", size = 7 ),    # Axis titles
    axis.text = element_text(family = "Helvetica",face = "plain",size = 6),    # Axis tick labels
    legend.title = element_text(family = "Helvetica",face = "bold.italic",size = 6),    # Legend title (header)
    legend.text = element_text(family = "Helvetica",face = "plain",size = 5),    # Legend labels
    
    # Legend placement
    legend.position = "bottom",
    legend.direction = "horizontal",
    
    # Make legend more compact
    legend.key.size = unit(3, "mm"),
    
    legend.spacing.y = unit(0, "mm"),
    
    legend.margin = ggplot2::margin(0, 0, 0, 0),
    legend.box.margin = ggplot2::margin(0, 0, 0, 0),
    
    legend.key.height = unit(1, "mm"),
    legend.key.width  = unit(4, "mm"),
    legend.spacing.x  = unit(1.5, "mm"),
    legend.box.spacing = unit(0.5, "mm"),
    
    # Remove background and grid
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_rect(fill = "white", colour = NA),
    
    # Keep only axes
    axis.line = element_line(colour = "black"),
    
    # Facets (if used)
    strip.background = element_blank(),
    strip.text = element_text(
      family = "Helvetica",
      face = "bold",
      size = 7
    )
  )
systemfonts::match_font("Helvetica")

# Format Internal Panels (4D, 4E, 4F)
plot4D <- overallPerformancePlot_internal+ theme_pub 
plot4E <- plot_conf_mat(conf_mat_int_WT1_combined_binary, predicted_col = "Prediction_binary", label_type = "count", title = "Confusion Matrix Heatmap\ninternal Validation - WT1 (germline & somatic)\nPer sample\nPredictionBinary", high_colour = "orange3", original_col = "original_WT1_combined")+ylab("WT1 PV")+xlab("Prediction") + theme_pub +  ggtitle("") +geom_text(
  aes(label = display_label),
  family = "Helvetica",
  fontface = "plain",
  size = 5/2.845
)
plot4E$layers[[2]] <- NULL
plot4F <- plotInternalSummary_Perc + theme_pub

plot4D/(plot4E + plot4F)

# Format External Panels (4G, 4H, 4I)
plot4G <- overallPerformancePlot_PerPatient + theme_pub
plot4H <- plot_conf_mat(conf_mat_ext_WT1_combined_perPatient_binary, label_type = "count", title = "Confusion Matrix Heatmap\nExternal Validation - WT1 (germline & somatic)\nPer Patient\nbinary prediction", high_colour = "royalblue4", original_col = "original_WT1_combined", predicted_col = "final_prediction_binary")+ylab("WT1 PV")+xlab("Prediction") + theme_pub+  ggtitle("") +geom_text(
  aes(label = display_label),
  family = "Helvetica",
  fontface = "plain",
  size = 5/2.845
)
plot4I <- plotExternalSummary_perc + theme_pub
plot4H$layers[[2]] <- NULL




# Assemble composite layout
left  <- plot4D / (plot4E + plot4F + plot_layout(widths = c(2, 1)))
right <- plot4G / (plot4H + plot4I + plot_layout(widths = c(2, 1)))

left <- left &
  theme(plot.margin = ggplot2::margin(r = 4, unit = "mm"))

right <- right &
  theme(plot.margin = ggplot2::margin(l = 4, unit = "mm"))

combined <- left | right

# Save combined composite PDF figure
# ggsave(
#   filename = file.path(parameters$RESULTS_FOLDER,"FIGURE4", "4D_4Icombined.pdf"),
#   plot = combined,
#   width = 165,
#   height = 82,
#   units = "mm",
#   device = cairo_pdf
# )
```


